## Symbolic manipulations 

Several of the pattern book functions appear so often in *MOSAIC Calculus* that it is worth reviewing how to manipulate them symbolically. As an example, consider the function $$g(x) \equiv e^x e^x\ .$$
This is a perfectly good way of defining $g()$, but it is helpful to be able to recognize that the following definitions are exactly equivalent
$$f(x) \equiv e^{x+x}\\
h(x) \equiv e^{2 x}\ .$$
The ***symbolic manipulation*** we touch on in this chapter involves being able to recall and apply such equivalences. we will need only a small set of them, all or most of which are found in a high-school algebra course.


::: {.callout-note icon=false data-latex=""}
## Why do some functions have parentheses and other don't?

How come some function names, like $\sin()$ are written with other parentheses, while others such as $e^x$ have the input name shown?

The $x$ in the name $e^x$ is  a placeholder. A better, if longer, name would be $\exp()$, which would signal that we mean the abstract concept of the exponential function, and not that function applied to an input named $x$. 

The same is true for functions like $x$ or $1/t$ or $z^2$. If absolute consistency of notation were the prime goal, we could have written this book in a style that gives a name to every pattern-book function in the name/parentheses style. Something like this:

```{r include=FALSE}
library(mosaic)
```

```{r}
reciprocal <- makeFun(1/t ~ t)
one <- makeFun(1 ~ z)
square <- makeFun(x^2 ~ x)
```

These would be used in the ordinary way, for instance:

```{r}
reciprocal(7)
one(123.67)
square(19)
```

Writing reciprocal($x$) instead of $1/x$ is long-winded, which is perhaps why you never see it. But when you see $1/x$ you should think of it as a function being applied to $x$ and not as a bit of arithmetic.

By the way ... I used different names for the inputs in these three functions just to remind the reader that, for functions with one input, the name has no significance. You just have to make sure to use the same name on the left- and right-hand sides of the tilde expression. 
:::

### Exponential and logarithm {#sec-exp-log-symb}

Basic symbolic patterns for exponentials are (i) and (ii)

::: 
$\LARGE\mathbf{(i)}\ \ \ \ \ e^x e^y \leftrightarrow e^{x+y}$ 

.
    
$\LARGE\mathbf{(ii)}\ \ \ \ \ \left(e^x\right)^y \leftrightarrow e^{x y}$
:::


Exponentials with a base other than $e$ can be converted to base $e$.

:::
.
$\LARGE\mathbf{(iii)}\ \ \ \ \ 2^x \leftrightarrow e^{\ln(2) x} = e^{0.69315 x}$

$\LARGE\mathbf{(iv)}\  \ \  \ 10^x \leftrightarrow e^{\ln(10) x} = e^{2.30259 x}$
:::

For mental arithmetic, easier use base 2 or 10. The base $e = 2.718282$ is not conducive to such calculations. In Block 2 we will discuss why it is standard to write an exponential function as $e^x$.

The logarithms are the ***inverse*** function of the exponential: Rule (v).

::: 
$\LARGE\mathbf{(v).a}\ \ \ \ \ \ln(\exp(x)) = x$ 

$\LARGE\mathbf{(v).b}\ \ \ \ \ \exp(\ln(x)) = x$
:::


One place that we will encounter rules (ii) and (v) is in @sec-parameters when we look a "doubling times" and "half lives." There we will deal with expressions such as $2^{x/\tau} = e^{\ln(2) x/\tau}$. 

Important symbolic patterns for logarithms are Rules (vi) through (vii).

::: 
$\LARGE\mathbf{(vi)}\ \ \ln(x\ \ \!y) \leftrightarrow \ln(x) + \ln(y)$

$\LARGE\mathbf{(vii)}\ \ \! \ln(x / y) \leftrightarrow \ln(x) - \ln(y)$

$\LARGE\mathbf{(viii)}\ \ \ \ln(x^p) \leftrightarrow p \ln(x)$$
:::

::: {.callout-tip}
Notice that the symbolic patterns for logarithms involve multiplication, division, and exponentiation, but **not addition**: $\ln(x + y) \neq \ln(x) + \ln(y)$.
:::

### Power-law functions

In power-law functions, the quantity in the exponent is constant: we will call it $m$, $n$, or $p$ in the following examples.

::: 
$\LARGE\mathbf{(ix)}\ \ \ \ x^m x^n \leftrightarrow x^{m+n}$

$\LARGE\mathbf{(x)}\ \ \ \ \ \ \frac{x^m}{x^n} \leftrightarrow x^{m-n}$

$\LARGE\mathbf{(xi)}\ \ \ \ \ \left(x^n\right)^p \leftrightarrow x^{n\,p}$

$\LARGE\mathbf{(xii)}\ \ \ \ \ \ x^{-n} \leftrightarrow \frac{1}{x^n}$

$\LARGE\mathbf{(xiii)}\ \ \ \ \ \ \ x^0 \leftrightarrow 1$
:::

### Sinusoids

$\sin(x)$ is periodic with period $2\pi$. Zero-crossings of $\sin(x)$ are at $x=..., -2\pi, -\pi, 0, \pi, 2\pi, ...$

As we mentioned earlier, we will be calling both the $\sin()$ and $\cos()$ function "sinusoids." They are merely shifted versions of one another:
$$\cos(x) = \sin\left(x + \frac{\pi}{2}\right)\ .$$



### The straight-line function

You are probably used to a function that we call the "straight-line function" $$\line(x) \equiv m x + b\ .$$ The name comes from the shape of a graph of $\line(x)$ versus $x$, which is a straight line. You are likely used to calling the parameter $m$ the "slope" of the line, and the parameter $b$ the "intercept." (In general, by "intercept" we will mean the value of the function output when the input is zero. In high-school, this is often called the "y-intercept.")

There are two simple symbolic manipulations that you will be using often in *MOSAIC Calculus*:

1. Find the input $x=x_0$ for which the output $\line(x=x_0) = 0$. This input has many names in mathematics: the "root," the "$x$-intercept," the "zero crossing," etc. We will call any input value that corresponds to an output of zero to be "a zero of the function."

For the straight-line function, the zero is readily found symbolically: $$x_0 = - b/m\ .$$

2. Re-write the straight-line function in the form $$\line(x) = a \left(x - x_0\right)\ .$$

Here, the slope is designated with $a$. And, of course, $x_0$ is the zero of the function, as you can see by setting $x=x_0$:
$\line(x=x_0) = a (x - x_0){\LARGE\left.\right|}_{x = x_0} = 0\ .$


