To construct an anti-derivative with the Euler Method, you need:

1. Obviously ... The function to be anti-differentiated, which we will call $f(t)$ in the example.
2. Specific numerical values for all parameters of the function, e.g. the $k$ in $e^{kt}$ or the $P$ in $\sin(2\pi t/P)$.
3. The domain over which you want to know the anti-derivative.
4. A numerical value for a "step size" $h$. This should be "small," but "small" depends on the problem being tackled, just as you saw with the graphics in the previous problem.

To see where the Euler Method comes from, recall that anti-differentiation is "undoing" differentiation and remember the discrete time version of differentiation:
$$\partial_x f(x) \approx \frac{f(x+h) - f(x)}{h}$$
To make it clear where the anti-derivative comes in to this, let's rewrite the finite-difference formula using a different notation: $f(x)$ will be the function to be anti-differentiated and $F(x)$ will be the result of the anti-differentiation of $f(x)$. Using this notation:
$$f(x) \approx \frac{F(x+h) - F(x)}{h}$$
You can see that $f(x)$ is the derivative of $F(x)$, but here we know $f(x)$ but do not (yet) know $F(x)$. To find $F(x)$ at a series of discrete values of $x$ spaced by $h$, re-arrange the equation:
$$F(x + h) \approx F(x) + h f(x)\ \ \  \ \ \text{Euler method formula}$$

As an example of using this formula, let's anti-differentiate $\sin(2\pi x/0.3)$ over the domain $1 \leq x \leq 21$ with a step size of $h=0.01$

We start at $x=1$ and

i. Since we don't yet know $F(x)$, let's make our life simple and assume that $F(x=1) \equiv 0$. (You will see why this choice works later on. In fact, you could use any value you like for $F(x=1)$.)
ii. We do know $f(x)$. Calculate $f(x=1)$. This comes out to be `sin(2*pi*1 / 0.3)` or 0.866.
iii. Plug in the value of $f(x=1)$ from step (ii) to the Euler method formula to calculate $F(1 + 0.01)$. The formula, and our assumed value $F(1) = 0$,  gives $$F(1 + 0.01) = 0 + 0.01 \times 0.866 = 0.00866$$
iv. Now move on to the next time step. We know $f(1.01) = 0.743$ and $F(1.01) = 0.00866$. We use the Euler method formula to find
$$F(1.02) = F(1.01) + 0.01\times 0.743 =  0.00866 + 0.00743 = 0.01609$$

```{r eulersandbox, exercise = TRUE, exercise.cap="Sandbox for Euler calculations", exercise.nlines=5}

```

```{r euler1}
askMC(
  "What is the value of $F(1.03)$?",
  "+`0.01609 + 0.01*0.588`+",
  "`0.01609 + 0.01*0.743`" = "You should be using $f(1.02)$, not $f(1.01)$.",
  "`0.00866 + 0.01*0.743`"="You should be using $F(1.02)$, not $F(1.01)$.",
  "`0.00866 + 0.01*0.588`" = "You should be using $F(1.02)$ and  $f(1.02)$, not $F(1.01)$ and $f(1.01)$."
)
```

```{r euler2}
askMC(
  "How many steps will it take from $x=1.03$ to reach $x=1.05$?",
  1, 2, 3, 4, 5, right_one = 2,
  random_answer_order = FALSE
)
```

```{r euler3}
askMC(
  "What is $F(1.05)$?",
  "0.028" = "That's  $F(1.04)$.",
  "+0.030+",
  "0.032",
  "0.033",
  random_answer_order = FALSE
)
```
