---
id: "Basic modeling functions"
author: "Daniel Kaplan"
---

```{r include=FALSE}

knitr::opts_chunk$set(echo = FALSE)
```


Recall that in Math 141Z, we emphasized a small set of functions that are very commonly encountered in modeling the real-world. You learned the shapes of graphs of these functions, situations when each of them is relevant or useful, and the derivatives of these functions. They were:

Family | function | Derivative: $\partial_t$ function
------------------------|----------|:--------------------:
Power_law | $t^p$ | $p t^{p-1}$ (when $p \neq 0$)
Exponential | $e^{kt}$ | $k e^{kt}$
Sinusoid | $\sin(2\pi t/P)$ | $\frac{2 \pi}{P} \cos(2\pi t/P)$
Sinusoid | $\cos(2\pi t/P)$ | $-\frac{2 \pi}{P} \sin(2\pi t/P)$
Logarithm | $\ln(t)$ | $1 / t$



We will continue to focus on these basic modeling functions because 1) they are useful and 2) their anti-derivatives are so simple that you should learn to write them in math notation from memory. (We're leaving out hump & sigmoid functions because their algebra is not so generally useful. But remember that the derivative of a sigmoid is a gaussian function.)

When talking about anti-derivatives, you should think of the above table but with different column names:

Family | Anti-derivative: $\int$function$dt$ | function
------------------------|:----------:|---------------------
Power_law | $t^p$ | $p t^{p-1}$ (when $p \neq 0$)
Exponential | $e^{kt}$ | $k e^{kt}$
Sinusoid | $\sin(2\pi t/P)$ | $\frac{2 \pi}{P} \cos(2\pi t/P)$
Sinusoid | $\cos(2\pi t/P)$ | $-\frac{2 \pi}{P} \sin(2\pi t/P)$
Logarithm | $\ln(t)$ | $1 / t$

And, while we're at it, let's re-arrange the table so that the function itself is presented in a simple form, and we put the scaling etc. in the anti-derivative itself:

Family | Anti-derivative: $\int$function$dx$ | function
------------------------|:----------:|---------------------
Power_law | $\frac{1}{p} t^{p+1}$ | $t^{p}$ (when $p \neq 0$)
Exponential | $\frac{1}{k} e^{kt}$ | $e^{kt}$
Sinusoid | $\frac{P}{2 \pi}\sin(2\pi t/P)$ | $\cos(2\pi t/P)$
Sinusoid | $-\frac{P}{2 \pi}\cos(2\pi t/P)$ | $\sin(2\pi t/P)$
Logarithm | $\ln(t)$ | $1 / t$

Here are some examples of functions that are *not* basic modeling functions:

- $g(x) \equiv x e^{kx}$
- $h(t) \equiv t \sin(t^2)$
- $f(t) \equiv \sin^2(t) \cos(t)$

We'll encounter such functions, and they can be important in specific situations. But the basic modeling functions are so widely encountered that learning their anti-derivatives by heart will pay off throughout your studies in numerous fields.

```{r TF1}
askMC(
  "Is $e^{-0.3 t}$ a basic modeling function with a linear interior function?",
  "+Yes+" = "The linear interior function has general form $ax + b$. Here, $b=0$ and $a= -0.3$.", 
  "No",
  random_answer_order = FALSE
)
```

```{r TF2}
askMC(
  "Is $2 x \\sin(x/4)$ a basic modeling function with a linear interior function?",
  "Yes" = "This is a *product* of two functions: $x$ and $\\sin(x/4)$. Each of those is a basic modeling function with a linear interior function, but the product of the two is an entirely different family of functions.", 
  "+No+" = "The *product* of two basic modeling functions, here $x$ and $\\sin(x/4)$, is generally *not* a basic modeling function. (Exceptions: A power law times a power law is still a power-law function. An exponential times an exponential is still an exponential function.)",
  random_answer_order = FALSE
)
```

```{r TF3}
askMC(
  "Is $\\sqrt{x}$ a basic modeling function with a linear interior function?",
  "+Yes+" = "$\\sqrt{x}$ is a power-law function, which might equally well be written as $(x)^{1/2}$. The interior function here is very simple: $x$. That is indeed of the linear form $ax +b$, with $a=1$ and $b=0$", 
  "No",
  random_answer_order = FALSE
)
```
