---
id: "bee-bid-knife"
created: "Tue Sep  7 14:28:35 2021"
global_id: "wmyERp"
---

Every function has an anti-derivative. But there are many functions, for example $\dnorm()$ where there is no formula for the anti-derivative. Even so, the R/mosaic `antiD()` operator will figure out what is the anti-derivative. 

For each of the following functions, use `makeFun()` to implement the function as `f()`. Then use `antiD()` to construct the anti-derivative of `f()` and give this the name `F()`. Finally, use `D()` to compute the derivative of `F()`. For instance:

```{r warning=FALSE}
f <- makeFun(x*dnorm(x) ~ x)
F <- antiD(f(x) ~ x)
dF <- D(F(x) ~ x)
slice_plot(f(x) ~ x, domain(x=0:10)) %>%
  slice_plot(dF(x) ~ x, color="blue", size=3, alpha = 0.2)
```


```{r bbk3-1, echo=FALSE, results="markup"}
askMC(
  "For the function $f(x) \\equiv \\pnorm(x^2)$, what is $F(5)$? (Pick the closest answer.)",
  4, "+4.5+", 5, 5.5, 6, 6.5, 
  random_answer_order = FALSE
)
```

```{r bbk3-2, echo=FALSE, results="markup"}
askMC(
  "For the function $f(x) \\equiv \\sin(x) \\ln(x^2)$, what is $F(5)$? (Pick the closest answer.)",
  -4, -3.5, -3, "+-2.5+", -2,
  random_answer_order = FALSE
)
```

```{r bbk3-3, echo=FALSE, results="markup"}
askMC(
  "For the function $f(x) \\equiv x^2 \\pnorm(x)$, what is $F(5)$? (Pick the closest answer.)",
  35, 37, 39, "+41+", 43,
  random_answer_order = FALSE
)
```

