---
id: "bird-hold-jacket"
created: "Tue Jul 20 14:31:36 2021"
global_id: "1qDQZz"
---

Go through the steps above to find the anti-derivative of $g(x) \equiv x \cos(x)$. 

Step 1 hint: We know the anti-derivative of $\cos(x)$ is $\sin(x)$, so an appropriate helper function is the function $x\, \sin(x)$. Now do steps (2) and (3): (2) take the derivative of the helper function and then (3) integrate each term in the result. 

```{r parts1, echo=FALSE}
askMC(
  "What is the derivative of the helper function with respect to $x$?",
  "+$\\partial_x \\text{helper}(x) = \\sin(x) + x \\cos(x)$+",
  "$\\partial_x \\text{helper}(x) = \\sin(x) + x \\sin(x)$",
  "$\\partial_x \\text{helper}(x) = \\sin(x) + \\cos(x)$",
  "$\\partial_x \\text{helper}(x) = \\sin(x)\\cos(x)$"
)
```

```{r parts2, echo=FALSE}
askMC(
  "What is $$\\int \\partial_x \\text{helper}(x)\\ ?$$",
  "+$\\text{helper}(x) + C$+" = "We included the constant of integration $C$ just as a reminder.",
  "$\\frac{1}{2} \\text{helper}^2(x)$",
  "$1 / \\text{helper}(x)$",
  "Whatever it is, it's just as complicated as the original integral. No obvious way to do it." = "Actually, the answer is simple, even if it doesn't seem to get us anywhere. Wait for the next problem!")
```

```{r parts3, echo=FALSE}
askMC(
  "What is the integral with respect to $x$ of the first part of the expanded form of the helper function, that is, $\\int \\sin(x) dx$?",
  "+$-\\cos(x)$+" = "This is one of our basic modeling functions.",
  "$\\cos(x)$" = "Close. $\\cos(x) = \\partial_x \\sin(x)$, but we want $\\int \\sin(x) dx.$",
  "$e^x \\sin(x)$", 
  "$e^x \\cos(x)$"
)
```

```{r parts4, echo=FALSE}
askMC(
  "What is the integral with respect to $x$ of the second part of the expanded form of the helper function, that is, $\\int x\\, \\cos(x)$?",
  "It's the same as the original problem! I thought you were showing us how to do the problem. If we didn't know the answer when we started, why should we be able to do it now?" = "An understandable frustration. But think! Now you have the original problem written out as part of an equation with two integrals that you *do* know.",
  "+It's the same as the original problem. I've got an equation involving the original problem and two bits of algebra/calculus that I know how to do. Thanks!+",
  "$\\sin(x)$" = "If this were true, then $\\partial_x \\sin(x)$ would give us $x\\, \\cos(x)$. But, as you know, $\\partial_x \\sin(x) = \\cos(x)$, so this answer must be wrong."
)
```

Solve for the answer to the original function and write the function in R notation here:

```{r parts5, exercise=TRUE, exercise.cap="Answer to the original problem.", exercise.nlines=3, eval=FALSE}
makeFun( ...your stuff here... ~ x)
```

```{r parts5-check, echo=FALSE, eval=FALSE}
grade_result(

  pass_if( ~ sd(.result(1:10) -  ((1:10)*sin(1:10) + cos(1:10))) < 0.001),
  fail_if( ~ TRUE, message = "No!")
)
```


