---
id: "buck-forgive-canoe"
created: "Wed May 25 12:35:29 2022"
global_id: "AMqazT"
---

`r insert_calcZ_exercise("XX.XX", "AMqazT", "/Users/kaplan/MC2/Accumulation/Exercises/buck-forgive-canoe.Rmd")`

The KC-46 is an airborne fuel tanker that can receive or dispense fuel. Suppose
$x(t)$ tracks a KC-46's fuel transfer rate, in pounds per minute, where $t$ is measured in minutes from the start of the mission.  Note that $0 < x(t)$ when the KC-46 is receiving fuel from another tanker and $x(t)<0$ when the KC-46 is offloading fuel to a supported aircraft.

The function $b(t)$ tracks the KC-46's fuel burn rate, also measured in pounds per minute, with the same interpretation of $t$.  Note that $b(t)$ can never be negative (i.e. $0 \leq b(t)$)


```{r bfcu-1, echo=FALSE, results="markup"}
askMC(
  prompt = "Which of the following is the proper representation of the KC-46's fuel state $n$ minutes after mission start?",
  r"($x(t)-b(t)$)",
  r"($x(t)-b(t)+f_0$ where $f_0$ represents the fuel level at the start of the mission.)",
  r"($$\int_0^n x(t)dt-\int_0^n b(t)dt$$)",
  r"(+$$\int_0^n x(t)dt-\int_0^n b(t)dt +f_0$$ where $f_0$ represents the fuel level at the start of the mission+)" =  r"(Remember without an initial condition, we can only talk about net change.)"
)
```

```{r bfcu-2, echo=FALSE, results="markup"}
askMC(
  prompt = "What is the meaning of $$\\int_0^{120} x(t)dt=40,000\\ ?$$ Keep in mind that there are 60 minutes in 1 hour, so 120 minutes is 2 hours.",

  "In the first two hours of the mission, the KC-46 offloaded a total of 40,000 pounds of fuel.",
  "Two hours into the mission, the KC-46's fuel tanks contain 40,000 pounds of fuel.",
  "+In the first two hours of the mission, the KC-46 received 40,000 more pounds of fuel than it offloaded.+" = "The definite integral tells us about the ***net*** change.",
  "In the first two hours of the mission, the KC-46 received a total of 40,000 pounds of fuel.",
  random_answer_order = FALSE)
```
