---
id: "chicken-cost-coat"
created: "Tue Jul 20 14:32:17 2021"
global_id: "YMuG92"
---

Use the same procedure to find the anti-derivative of $x\, \cos(2x)$. Since $$\cos(2x) = \frac{1}{2}\partial_x \sin(2 x)$$ a reasonable guess for a helper function will be $x \sin(2x)$. 


(We have intentionally dropped the $1/2$ to simplify the rest of the procedure. You will see that such multiplicative constants don't matter, since they will be on both sides of the equation showing the derivative of the helper function. You can see this by keeping the $1/2$ in the helper function and watching what happens to it.)

As you work through the steps be very careful about the constants and make sure you check your final answer by differentiating.

```{r parts6, echo=FALSE, results="markup"}
askMC(
  "What is $\\partial_x x\\, \\sin(x)$?",
  "+$\\sin(x) + x\\, \\cos(x)$+",
  "$\\sin(x) + x\\, \\sin(x)$",
  "$\\cos(x) + x\\, \\sin(x)$",
  "$\\cos(x) + x\\, \\cos(x)$"
)
```

```{r parts7, echo=FALSE, results="markup"}
askMC(
  "What is $\\int \\partial_x [ x\\, \\sin(x)] dx$?",
  "+$x\\, \\sin(x)+C$+" = "Integration undoes differentiation!",
  "$\\sin(x)+C$",
  "$\\cos(x)+C$",
  "$x\\, \\cos(x)+C$"
)
```

```{r parts8, echo=FALSE, results="markup"}
askMC(
  "What is $\\int x\\, \\cos(x) dx$?",
  "+$x\\, \\sin(x) + \\cos(x)$+", 
  "$x\\, \\cos(x) + \\cos(x)$",
  "$x\\, \\cos(x) + \\sin(x)$",
  "$x\\, \\sin(x) + \\sin(x)$"
)
```


