---
id: "crow-eat-bottle"
created: "Thu Sep 23 10:49:34 2021"
global_id: "LpYZy4"
---



A giant tortoise (with very good eyesight and standing on an unobstructed plane!) spies a head of lettuce on the ground 65 meters away.  Being hungry (and knowing the shortest path between two points on the plane!), the tortoise takes off in a straight line for the lettuce. She pretty quickly reaches her top speed, but then starts to tire. If her velocity as a function of time (in meters per minute) is modeled by $v(t) = 7 t e^{-0.3t}$, how long does it take the tortoise to reach her lunch? Answer this question by finding an calculus/algebra formula for the tortoise’s displacement and then use it to approximate how long it takes to get to the lettuce.  `r mark(3775)`

We're going to be looking at $\int v(t) dt = 7 \int t e^{-0.3 t} dt$.

We'll call the left side of the equation "displacement(t)". 
Use integration by parts to find displacement(t) as a simple formula in $t$.

The tortoise will reach the cabbage at time $t^\star$ such that $\text{displacement}(t^\star) = 65$ meters. In a `r sandbox_link()`, graph your displacement function to find $t^\star$. Remember that there will be a constant of integration which can be found by noting that $\text{displacement}(0) = 0$. (Note that the graphics domain given in the example code isn't necessarily the best choice for answering the question.) `r mark(3780)`

```{r parts12, eval=FALSE}
displacement <- makeFun(7 * WHAT + C ~ t)
slice_plot(displacement(t) ~ t, domain(t=c(0,5)))
```

```{r parts13, echo=FALSE, results="markup"}
askMC(
  "At what time $t^\\star$ does the tortoise reach the cabbage?",
  "5.95 sec",
  "+10.85 sec+",
  "15.75 sec",
  "Never! (That is, $t^\\star$ is infinite.)",
  random_answer_order = FALSE
)
```


