---
id: "falling-from-pisa"
author: "Daniel Kaplan"
date: "Dec 26, 2020"
---

```{r include=FALSE}

```


Famously, the acceleration due to gravity at the Earth's surface is a constant: roughly -9.8 meters / sec$^2$. But let's write it as a function of time: $$\text{accel}(t) \equiv -9.8$$
Galileo climbs up to the top of the Tower of Pisa, 56.7 meters above the ground. He is holding a small lead ball between thumb and forefinger. He spreads his fingers and the ball starts to fall from a standstill. How fast will the ball be going as it falls.

In this problem, we are given information in the form of acceleration, and want to turn it into a statement about velocity. Remembering that acceleration is the derivative of velocity with respect to time, we see this as an anti-derivative problem: 

$$\text{vel}(t) \equiv \int \text{accel}(t) dt = \int -9.8 dt = -\int 9.8 dt$$
$$\text{vel}(t) = -9.8 t + C\ .$$

```{r galileo-1, echo=FALSE}
askMC(
  "Imagine that the time of release was $t=3$ seconds. Remembering that 9.8 is in meters/sec$^2$, what is the value of $C$ in the velocity function?",
  "29.4 meters per sec$^2$" = "Sorry, but $C$ has dimension of velocity.",
  "+29.4 meters per sec+" = "Right. At time $t=3$ the ball was at a standstill, meaning velocity zero. So $-9.8 \\times 3 + C = 0$ ",
  "29.4 meters",
  "29.4 dimensionless/unitless")
```

Where will the ball be in 2 seconds after release (that is, at 5 seconds)? "Where" is a position question, so we need to convert the original information that we had in terms of acceleration into information in terms of position.

We've already used the relationship between acceleration and velocity (acceleration is $\partial_t$ accel(t)) to compute the velocity vel(t) at any time: vel(t) $= -9.8 t + 29.4$ and we know the units of $t$, $-9.8$ and $29.4$.

Since velocity is the derivative of position with respect to time, we can compute position as the anti-derivative of velocity: $$\text{pos}(t) = \int \text{vel}(t) = \int (29.4-9.8 t) dt$$
$$\text{pos}(t) = 29.4 t - 9.8 \frac{t^2}{2} + C_{pos}$$
Another $C$! we are writing it as $C_{pos}$ to remind us that this $C$ is different than the $C$ that appeared in the vel(t) function. That $C$ was from the accumulation of acceleration to get velocity. This $C_{pos}$, even though it is the same letter, is something different. Remember that $C_{pos}$ is a reminder that pos(t) is the accumulation of velocity up until time $t$, but we haven't (yet) specified when the accumulation started. To calculate the $C_{pos}$ for the pos(t) function, we need to consider the problem situation (Galileo in Pisa) and figure out some $t_0$ for which we know $pos(t_0)$.

What we know here is that at the release time ($t=3$ seconds) the ball was at the top of the Tower of Pisa. So dist(3) = 56.7 meters.

```{r galileo-2, echo=FALSE}
askMC(
  "What's the value of $C_{pos}$ in the distance function given that dist(3) = 56.7 meters?",
  "+$C = 56.7 - 29.4 \\times 3 + 9.8 \\times 3^2 / 2 = 12.6$ meters+",
  "$C = -29.4 \\times 3 + 9.8 \\times 3^2 / 2 = 44.1$ meters" = "to solve for C you always need the equation AND a single point",
  "$C = 56.7$ meters" = "C is not always equal to the intital value, rember the release is at t=3",
  "$C = 0$ meters",
  "A quantity not in meters." = "Since pos(t) is in meters, the $C$ must also be in meters. Look at the formula for pos(t)!"
)
```

```{r galileo-3, echo=FALSE}
askMC(
  "What is the position of the ball two seconds after release? (The units for each of these is \"meters\" above the ground.)",
  "$29.4 \\times 2 - 9.8 \\times 4/2 + 12.6$" = "Two seconds after release (which occurs at $t=3$) is what value of $t$?",
  "+$29.4 \\times 5 - 9.8 \\times 5^2/2 + 12.6$+",
  "$29.4 - 9.8 \\times 5 / 2 + 12.6$" = "Check the formula for pos(t) to make sure where the t appears."
)
```
