---
id: "Find-C"
date: "MMM DD YYYY"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r include=FALSE}
etude2::load_learnr_safely()
library(gradethis)
library(etude2)
library(shiny)
knitr::opts_chunk$set(echo = FALSE)
learnr::tutorial_options(exercise.timelimit = 60,
                 exercise.checker = gradethis::grade_learnr)
```

```{r context="setup", include=FALSE}
library(mosaic)
library(mosaicCalc)
```


The top graph shows a function $f(t)$. 

The bottom graph shows an anti-derivative $$F(t) \equiv \int f(t) dt$$

You can make measurements by *click/dragging* the mouse on the bottom graph.

```{r}
plotOutput("littlef", height="250px")
plotOutput("Fplot", brush=brushOpts("Fbrush", direction="x", fill=NA, stroke="orange", opacity = 1), height="250px")
h5("Stats on F() for orange lines")
htmlOutput("boxstats")
```

```{r context="server"}
f_seed <- reactiveVal(101)
f_function <- reactive({
  mosaic::rfun( ~ t, seed = f_seed())
})

output$boxstats <- renderText({
  if (is.null(input$Fbrush)) return("Drag mouse on graph to measure")
  width  <- input$Fbrush$xmax - input$Fbrush$xmin
  f <- f_function()
  F <- antiD(f(t) ~ t)
  yvals <- F(c(input$Fbrush$xmin, input$Fbrush$xmax))
  deltaF <- diff(yvals)
  glue::glue(
    "<p>Left: {signif(input$Fbrush$xmin,3)}      Right: {signif(input$Fbrush$xmax, 3)}      Width: {signif(width,3)}</p>
    <p>F(left): {signif(yvals[1], 3)}       F(right): {signif(yvals[2], 3)}       ∆F: {signif(deltaF,3)}</p>
    <p>Slope: {signif(deltaF/width, 3)} </p>"
  ) %>% HTML()
})

output[["littlef"]] <- renderPlot({
  f <- f_function()
  
  suppressWarnings({
    P <- slice_plot(f(t) ~ t, domain(t=c(-5, 5))) %>%
      gf_hline(yintercept = 0, color="dodgerblue") %>%
      gf_labs(y = paste("f(t) for seed", input$seed),
              title = "f(t), the derivative of F(t)") %>%
      gf_refine(scale_x_continuous(breaks = (-5):5))
  })
  if (! is.null(input$Fbrush)) {
    Area <- tibble(
      t = seq(input$Fbrush$xmin, input$Fbrush$xmax, length=100),
      y = f(t)
    )
    Neg_area <- Area %>% filter(y < 0)
    P <- P %>% gf_ribbon(0 + y ~ t, data = Area, inherit=FALSE,
                         fill = "dodgerblue", color=NA, alpha = 0.3) 
    if (nrow(Neg_area) > 0)
      P <- P %>%
      gf_ribbon(0 + y ~ t, data=Neg_area %>% filter(y < 0), inherit = FALSE,
                fill = "orange3", color = NA, alpha = 0.3)
  }
  P
})
output$Fplot <- renderPlot({
  f <- f_function()
  F <- antiD(f(t) ~ t)
  suppressWarnings({
    slice_plot(F(t) ~ t, domain(t=c(-5, 5)), npts=200) %>%
      gf_hline(yintercept = 0, color="dodgerblue") %>%
      gf_refine(scale_x_continuous(breaks = (-5):5)) %>%
      gf_labs(y = "F(t)", title="F(t): the anti-derivative of f(t)")
  })
})
```

----

```{r context = "server"}
f_seed <- reactiveVal(864)
```

Sometimes it's easier to think through matters when they are presented in a concrete, everyday context. Imagine that you managing a small-town water tank. Water flows in from the supply system and flows out to consumer demand. At any instant of time, there's a net amount flowing to the tank, which could be measured in cubic-meters-per-second. (If the amount is negative, consumers are momentarily consuming more than is being newly supplied.) 

The amount of water in the tank, measured in cubic meters, also fluctuates over time. We're going to call the water in the tank your "stock" of water. As a water manager, you want to make sure that your stock never goes to zero. (Otherwise, the town's citizens will be without water!)

In terms of functions, the "flow" is $f(t)$. The definite integral of the flow $$\int_a^b f(t) dt$$ gives you some information about the stock: how much the amount of stock has changed from time $a$ to time $b$. This is not, however, the same as the stock. Presumably, there was already some water in the tank at time $a$, so the stock at time $b$ will be that amount plus the amount of the definite integral of $f(t)$.



```{r C1}
askMC(
  "A. From the graph, roughly estimate $$\\int_0^3 f(t)dt$$ Choose the closest numerical value from the following. You could use either graph for this caclulation, but $F(t)$ will be simpler algebra.",
  -46, -26, "0"="F(t) has a net change over the interval", "3"="there is more stock at F(3) than F(0)", "19"="there is more stock at F(3) than F(0)", "26"="F(t) at the right end of the interval is lower than at the left end", right_one = -26,
  random_answer_order = FALSE
)
```

```{r C2}
askMC(
  "B. to construct the anti-derivative whose value at time $t=-3$ will be zero, what constant of integration $C$ should you **add** to the $F(t)$ shown.",
  "-120"= "Would adding this to F(-3) be equal to 0?", "-80"= "Would adding this to F(-3) be equal to 0?", "-50"= "Would adding this to F(-3) be equal to 0?", "0"= "Would adding this to F(-3) be equal to 0?", "50" = "Would adding this to F(-3) be equal to 0?", 
  "80" = "Are you sure you have the sign right?",
  "120" = "Would adding this to F(-3) be equal to 0?",
  right_one = -80,
  random_answer_order = FALSE
)
```

```{r C3}
askMC(
  "C. Examining the stock at time $t=0$, you observe that there are 40 units. Roughly how much stock will there be at $t=5$?",
  "-25" = "Did you shift the curve in the right direction?", "-15" = "That would be the answer if there had been 0 units of stock at time $t=0$.","0"="Imagine F(t) being shifted by the difference of 40 and F(0)", "15"="Imagine F(t) being shifted by the difference of 40 and F(0)",  25, right_one=25,
  random_answer_order = FALSE
)
```

```{r C4}
askMC(
  "D. You start with a stock of 100 units at time $t = -2$. At roughly what time $t$ will the stock be half of this?",
  "-1.2"="$F(t)$-$F(a)$=50, what is the value of a?", -0.3, "0.5"="$F(t)$-$F(a)$=50, what is the value of a?", "1.2"="$F(t)$-$F(a)$=50, what is the value of a?", "1.8"="$F(t)$-$F(a)$=50, what is the value of a?",
  "The stock will never fall so low.",
  right_one = -0.3,
  random_answer_order = FALSE
)
```

```{r C5}
askMC(
  "E. Your stock finally runs out at time $t=2.5$. When did you have 120 units in stock?",
  "+$t=-4$+", "$t=-3$"="This uses a similiar approach to the last question.", "$t=0$"="This uses a similiar approach to the last question.", 
  "There was never such a time."
)

```

```{r C6}
askMC(
  "F. After decreasing for a long time, the stock finally starts to increase from about $t=2.5$ onward. What about $f(t=2.5)$ tells you that $F(t=2.5)$ is increasing?",
  "The derivative is at a minimum."="Does this mean $F(t)$ must be increasing?",
  "The derivative is negative"="Does this mean $F(t)$ must be increasing?",
  "The derivative is near zero." = "Kind of. For the stock $F(t)$ to increase, what has to be true of the derivative at that instant.",
  "+The derivative becomes positive and stays positive.+",
  "The derivative is at a maximum."="Does this mean $F(t)$ must be increasing?",
  random_answer_order = FALSE
)
```

```{r C7}
askMC(
  "G. Find the argmin $t^\\star$ of $f(t)$ and note the sign of $f(t^\\star).$  What does this tell you about $F(t^\\star)$.",
  "$t^\\star$ is also the argmin of $F()$."="look at the corresponding point on $F(t)$.",
  "+$F(t^\\star)$ is decreasing at its steepest rate.+",
  "$F(t^\\star)$ is increasing at it's slowest rate."="look at the corresponding point on $F(t)$.",
  "$F(t^\\star)$ is increasing at its steepest rate."="look at the corresponding point on $F(t)$.",
  "$t^\\star$ is the argmax of $F()$"="look at the corresponding point on $F(t)$."
)
```

```{r C8}
askMC(
  "H. What is the average flow into stock over the period $-5 \\leq t \\leq 1$. (If the flow is *outward from stock*, that's the same as a negative inward flow.)",
  -30, -20, "0" = "But you can see that the stock is diminishing steadily during $-5 \\leq t \\leq 1$, so how could the average flow be zero.", "10", right_one = -20,
  random_answer_order = FALSE
)
```

```{r C9}
askMC(
  "I. Which of the following is an interval when the average flow is approximately zero?",
  "$-1.3 \\leq t \\leq 1.1$"="The area appears to be negative.",
  "+$0.8 \\leq t \\leq 5$+",
  "$-5 \\leq t \\leq 0$"="The area appears to be negative.",
  "None of the above",
  random_answer_order = FALSE
)
```

```{r C10}
askMC(
  "J. From the graph, estimate $$\\int_2^{-4} f(t)dt$$ Choose the closest numerical value from the following.",
  "-120" = "Pay attention to the order of the limits of integration.", -60, 60, 120, 
  "None of these answers are close to being right.", 
  right_one = 120,
  random_answer_order = FALSE
)
```
