---
id: "function-order"
author: "Daniel Kaplan"
---

Although any function has an anti-derivative, that anti-derivative cannot always be presented in algebraic notation. This poses no fundamental problem to the construction of the anti-derivative, particularly when a computer is available to handle the book-keeping of numerical integration.

Still, it is convenient to have an algebraic form *when it can be found.* Many people have devoted considerable effort to constructing extensive collections of functions for which an algebraic form of anti-derivative is known. Think of such collections as a gallery of portraits of people who happen to have red hair. No matter how large the collection, you'll often have to deal who are not redheads. And unlike real redheads, it can be hard to know whether a function has an anti-derivative that can be expressed simply in algebraic form. For instance, $f(x) \equiv \exp(-x^2)$ does not, even though it is ubiquitous in fields such as statistics.

The US National Institute of Standards and Technology (NIST) has been a primary publisher for more than 50 years of information about functions encountered in applied mathematics. The work, [published originally in book form](https://www.youtube.com/watch?v=Exf02R1FnXY), is also available via the internet as the *[NIST Digital Library of Mathematical Functions](https://dlmf.nist.gov/)*.

So, how to organize the gallery of redheads? Let's take a field trip to the NIST DLMF (The US National Institute of Standards and Technology (NIST) has been a primary publisher for more than 50 years of information about functions encountered in applied mathematics. The work, [published originally in book form](https://www.youtube.com/watch?v=Exf02R1FnXY), is also available via the internet as the *[NIST Digital Library of Mathematical Functions](https://dlmf.nist.gov/)*!

Warning! Many visitors to NIST DLMF encounter dizziness, fatigue, and anxiety. Should you experience such symptoms, close your eyes and remember that DLMF is a *reference work* and that you will not be examined on its use. Nonetheless, to help you benefit maximally from the field trip, there are a few questions in this Daily Digital for you to answer from DLMF.

You should also note that the techniques in almost universal use to help you navigate through voluminous collections of data (e.g. Twitter, Facebook, Instagram, YouTube) such as ratings, subscribing, "friending," following, etc. are *entirely absent from DLMF*. There's not even a friendly introduction to each chapter saying who the material might be of interest to.

We're going to focus on Chapter 4, "Elementary Functions," and indeed just a few sections from that chapter. (A better name for the chapter would be "The Functions Most Often Used." They are not "elementary" as in "elementary school" but as in the "periodic table of elements.")

[Section 4.10](https://dlmf.nist.gov/4.10) covers integrals and anti-derivatives of logarithmic, exponential and power-law functions. 

[Section 4.26](https://dlmf.nist.gov/4.26) is similar, but for trigonometric functions.

1. Navigate to [equation 4.10.1](http://dlmf.nist.gov/4.10.i). This is one of the anti-derivatives you are expected to know by heart in CalcZ.
    a. Notice that the variable name $z$ is used. They could have selected any other name; $x$ and $t$ are popular, $y$ less so, and $\xi$ even less so. The use of $z$ is a signal to the <span title="plural noun:
people who are considered to be especially well informed about a particular subject: 'it was hailed by the cognoscenti as one of the best golf courses in Europe.;">*cognoscenti*</span> that the function can be applied to both real and complex numbers.
    b. Look at equation 4.10.8. This is another of the functions whose anti-derivative you should know by heart. 
    c. Perhaps it would have avoided some confusion if 4.10.1 had been written in terms of $$\int \frac{1}{az}dz$$ so that you would know what to do if you had encountered such a function.
    
Some exercises:
    
```{r DLMF1, echo=FALSE}
askMC(
  "Which of these is $$\\int \\frac{1}{az}dz \\ \\text{?}$$",
  "+$$\\frac{1}{a} \\ln(z)$$+" = "The level of the DLMF is such that you would be expected to know on your own that $\\int \\frac{1}{az} dz = \\frac{1}{a}\\int \\frac{1}{z}dz$. In other words, the $\\frac{1}{a}$ here is a scalar multiple of the function $\\frac{1}{z}$ and \"the integral of a scalar multiple of a function is the scalar multiple of the integral of the function.\". Really.",
  "There is no anti-derivative of $1/az$." = "Every function has an anti-derivative.",
  "$$a \\ln(z)$$" = "You lost track of a reciprocal somewhere."
)
```

```{r DLMF2}
askMC(
  "Use Section 4.10 of DLMF to find $$\\int \\frac{1}{3 t + 5}dt$$",
  "+$$\\frac{1}{15}(3t - \\ln(e^{3t} + 5))$$+",
  "There's no such function listed in Section 4.10." = "Although DLFM wrote the function using the variable name $z$, using $t$ instead is perfectly legitimate.",
  "$$\\frac{1}{15}(5 t - \\ln(e^{5t} + 3))$$" = "Looks like you mixed up the $a$ and $b$."
)
```

```{r DLMF3}
askMC(
  "Is $\\frac{1}{e^{az} + b}$ different from $e^{-(az + b)}$?",
  "+Yes+" = "$e^{-(az + b)}$ is a simple exponential with the linear interior function $-(az + b)$.",
  "No" = "They look similar, but $1/(e^{az} + b)$ is not an exponential function and it's not $1/e^{az} + 1/b$ either.",
  "Depends on the value of $b$." = "A technically correct answer but misleading, since it's only for $b=0$ that the two functions are equal."
)
```

```{r DLMF4}
askMC(
  "Using section 4.26, find $\\int \\tan(\\theta) d\\theta$.", 
  "+$-\\ln(\\cos(\\theta)$+",
  "$\\tan(\\theta)$ doesn't have an anti-derivative." = "All functions have an anti-derivative.",
  "There's no $\\theta$ in section 4.26" = "The authors didn't use $\\theta$ as the name of the variable, but $\\theta$ (\"theta\") is every bit as good a name to use as $x$ or $t$."
)
```

