---
id: "geometry"
topic: "Volumes"
author: "Daniel Kaplan"
---

```{r include=FALSE}
library(mosaicCalc)

```

Some of the prettiest examples of the differential modeling strategy are motivated by geometry.

The famous Great Pyramid of Giza originally was 280 Egyptian Royal Cubits tall and had an edge length at the base of 440 Royal Cubits. What's the volume in "cubic-cubits."

**Modeling set up.** Imagine slicing the pyramid horizontally with the slices stacked from base to 280 cubits high. The area of the slice at the very base is 440^2 = 193,600 square-cubits. The volume differential is 193,600 $dh$, where $dh$ is the thickness of the slices. A function for the edge length is $$\text{edge}(h) \equiv 440 - b\times h$$ We need to choose $b$ so that the edge length at $h=280$ gives 0, so $b=440/280$. The differential volume at height $h$ is 
$$\text{edge}^2(h) dh = 440^2 (1 - h/280)^2 dh$$

**Solution**

```{r geom1, echo=FALSE}
askMC(
"Find the anti-derivative of $440^2 (1-h/280)^2$ and evaluate it at $h=0$ and $h=280$ to find the volume. What is that volume? (Just FYI, 1 Royal Cubit is 52.4 cm.)",
  "16,984,415 cubic-cubits",
  "17,542,783 cubic-cubits",
  "+18,069,333 cubic-cubits+",
  "18,663,963 cubic-cubits",
  random_answer_order = FALSE
)
```

The top half of a circle of radius $r$ centered at the origin is given by the function $$y(x) = r \sqrt{1-x^2/r^2}\ \ \ \text{for}\ \ -r < x < r$$

**Find the volume of a spherical ball of radius $r$.**

**Modeling set up** Imagine the ball as a set of thin disks (thickness $dx$) stacked side by side. The disk at position $x$ has radius $y(x)$, area $\pi y^2(x)$ and volume $\pi y^2(x) dx$. 

Solution: Add up the volumes of all the disks. $$V(r) = \int_{-r}^r \pi r^2 (1-x^2/r^2) dr = \pi\left[\int_{-r}^{r}r^2 dx- \int_{-r}^{r} x^2 dx\right]$$ Compute the anti-derivative and evaluate it from $x=-r$ to $r$ to get the volume.
$$\pi r^2 x\left.\right|_{-r}^{r} - \pi x^3/3\left.\right|_{-r}^{r}$$

```{r geom2, echo=FALSE}
askMC(
  "Carry out the evaluation in the line above. What is $$\\pi r^2 x\\left.\\right|_{-r}^r \\ \\text{?}$$ Remember what part of the problem you are doing and the integral involved a dx.",
  "0",
  "$\\pi r^3$",
  "$\\frac{4}{3} \\pi r^3$",
  "+$2 \\pi r^3$+",
  random_answer_order = TRUE
)
```

```{r geom3, echo=FALSE}
askMC(
  "Imagine the following came from a similiar integral, what is $$\\pi r^2 x^2\\left.\\right|_{-r}^r \\ \\text{?}$$",
  "+0+",
  "$\\pi r^3$",
  "$\\frac{2}{3} \\pi r^3$",
  "$\\frac{4}{3} \\pi r^3$",
  random_answer_order = TRUE
)
```

Overall, the volume of a ball of radius $r$ is $$V(r) = \frac{4}{3} \pi r^3$$

Now think of the ball a different way, as a series of concentric spherical shells, each of thickness $dr$. Each shell is a differential. Just as the volume could be found by accumulating the volumes of the shells, so the volume of a spherical shell of radius $r$  and thickness $dr$ can be found from the derivative $\partial_r V(r)$

```{r geom4, echo=FALSE}
askMC(
  "What is volume of the spherical shell of radius $r$ and thickness $dr$?",
  "$\\frac{1}{3} \\pi r^4$" = "That's the integral of V(r), not the derivative.",
  "$\\frac{4}{3} \\pi r^2 dr$",
  "$\\frac{4}{3} \\pi r^2$" = "This leaves out the thickness of the shell.",
  "+$4 \\pi r^2 dr$+"
)
```

Compare your answer to the previous question to the formula for the surface area of a sphere of radius $r$. You can look up the formula easily if you don't remember it.

