---
id: "Glorias-family"
title: "Gloria's family line"
author: "Daniel Kaplan"
---

We have developed a notation that relates a function $f(x)$ to two other functions:

1. The *derivative* of $f(x)$ with respect to $x$, denoted $\partial_x f(x)$.
2. The *anti-derivative of $f(x)$ with respect to $x$, denoted $\int f(x) dx$.

There is a longer chain of derivatives and anti-derivatives of $g(x)$:

$$\ldots,\  \partial_{xxx} g(x), \ \ \partial_{xx} g(x), \ \ \partial_x g(x),\ \ g(x), \ \ \int g(x)dx, \ \ \int\! \int g(x) dxdx, \ \ \ldots$$
Pick one of these functions as a reference point and imagine that it is a person. The operation of differentiation is analogous to giving birth, that is, it produces another person, the child. Giving birth moves you to the left on the chain. Anti-differentiation moves you to the right. If $g(x)$ is Gloria, then $\partial_x g(x)$ is Gloria's child and $\partial_{xx}\! g(x)$ is Gloria's child's child, in other words, Gloria's grandchild. 

To illustrate: suppose "Gloria" is $g(x)\equiv x^2$. Here are the successive generations of Gloria's family:

Function | Using $\int$ and $\partial_x$ | Family connection
---------|-------------------------------|------------------
$x^4 / 12$ | $\int\! \int g(x) dx dx$ | Gloria's grandparent
$x^3 / 3$ | $\int g(x) dx$ | Gloria's parent
$x^2$ | $g(x)$ | Gloria
$2 x$ | $\partial_x g(x)$ | Gloria's child
$2$   | $\partial_{xx} g(x)$ | Gloria's grandchild


$G(x) = \int g(x)dx$ is Gloria's mother. If we apply the give-birth-to operator to Gloria's mother, who is produced? Gloria! In other words, $$\partial_x\! \int g(x) dx = g(x)$$
Gloria's mother giving birth produces Gloria. (Mothers can give birth to multiple children and that breaks the analogy to differentiation. Differentiation is like giving birth if a mother can have only one child.)

```{r gloria-exp, echo=FALSE}
askMC(
  "Suppose that Gloria is $g(x) \\equiv e^{(-2x + 3)}$. Which one of these is Gloria's parent?",
  "$-2 e^{(-2x + 3)}$" = "This is Gloria's child.",
  "$4 e^{(-2x + 3)}$" = "This is Gloria's grandchild.",
  "+$-\\frac{1}{2} e^{(-2x + 3)}$+",
  "$\\frac{1}{4} e^{(-2x + 3)}$" = "This is Gloria's grandmother.",
  "$\\frac{1}{4} e^{(3 x - 2)}$" = "The characteristic trait of Gloria, her ancestors and her descendents is $e^{(-2x +3)}$. We can't place this person anywhere in Gloria's family line.",
  "$4 x^3$" = "This person is from an entirely different tribe than Gloria."
)
```

```{r gloria-sine, echo=FALSE}
askMC(
  "Suppose Susan is $s(t) = 4 \\sin(4 t - 7)$. Which of these is Susan's grandparent?",
  "+$-\\frac{1}{4} \\sin(4 t - 7)$+",
  "$-\\frac{1}{16} \\cos(4 t - 7)$" = "Don't forget that the 4 in $4\\sin()$ is part of who Susan is.",
  "$16 \\cos(4 t + 7)$" = "Susan's child",
  "$-64 \\sin(4 t - 7)$" = "Susan's grandchild.",
  "$\\sin(4 t - 7)$" = "Not in Susan's line, even though a family similarity is evident."
)
```
