---
id: "Graph-anti"
author: "Daniel Kaplan"
---

This exercise uses the "Graph-antiD" web app which enables you to visualize the anti-derivative function as "areas." To use the app, click-drag-and-release to mark part of the domain of the function being displayed. 

::: {.todo}
Add a picture of the app and a link to it.
:::

To answer these questions correctly, **you must set the "Shape of function" box to 864**.

Remember that the graph shows $f(t)$, while these questions are about $F(t) = \int f(t) dt$.

```{r GAD1}
askMC(
  "At which of the following values of t does $F(t)$ have a positive slope?",
  "-3"="The slope of $F(t)$ is what function?",
  "0"="$f(t)$ is negative here, so the accumulated area will be getting more negative.",
  "+3+" = "Right. $f(t)$ is positive here, so the accumulated area $F(t)$ will be growing, which means that $F(t)$ has a positive slope.",
  random_answer_order = FALSE
)
```

```{r GAD2}
askMC(
  "At which of the following values of t is $F(t)$ concave down?",
  "+-3+" = "Good. The concavity of $F(t)$ has the sign of the second derivative $\\partial_{tt} F(t)$. But this second derivative of $F(t)$ is the first derivative of $f(t)$. So $F(t)$ is concave down when the slope of $f(t)$ is negative.",
  "0",
  "3",
  random_answer_order = FALSE
)
```

```{r GAD3}
askMC(
  "Which of the following statements about $F(t)$ is true at the point t=3.5?",
  "There is a maximum value at this point"="This is true about $f(t)$ not $F(t)$",
  "The slope changes from positive to negative"="This is true about $f(t)$ not $F(t)$",
  "+The concavity changes from concave up to concave down+",
  random_answer_order = FALSE
)
```
