---
id: "Units_of_antiD"
title: "Units of Anti-derivatives"
date: "Dec 23, 2020"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r include=FALSE}
library(mosaic)
library(mosaicCalc)
library(etude2)
```

The output of a function, being a quantity, has dimension and units. Suppose the dimension of the output of a function $v(t)$ is $L/T$, for instance, meters-per-second.

The anti-derivative function $V(t) \equiv \int v(t) dt$ will also have dimension and units.

Recall that in constructing the anti-derivative using the Euler method, we multiply the values of $v(t)$ times some small increment in the input, $h$. Therefore the dimension of the output of $V(t)$ will be $[v(t)] [t]$. So if $[v(t)] = L/T$, the dimension $[V(t)] [t] = L T/T = L$. Units for such a dimension would be, for instance, meters. This makes sense, since if you accumulate velocity (meters-per-sec) over an interval of time (sec) you end up with the distance travelled (meters).

Suppose you know the acceleration $a(t)$ of an object as a function of time. The dimension of acceleration is $L/T^2$.

```{r accel}
askMC(
  "What is the dimension of $$\\int a(t) dt\\ \\text{?}$$",
  "$L/T^3$", "$L$", "$LT$", "+$L/T$+", "$LT^2$"
)
```


Suppose you know the power consumed by an appliance $p(t)$ as a function of time. Typically appliances have a cycle and use different amounts of power during different parts of the cycle. (Think of a clothes washer.)

```{r power}
askMC(
  "What will be the dimension of $\\int p(t) dt\\ \\text{?}$",
  "+energy+", "force", "acceleration", "length"
)
```
