---
id: "wind-power"
title: "Accumulating Wind Power"
author: "Daniel Kaplan"
---



```{r include=FALSE}
library("mosaic")
library("mosaicCalc")
tmp <- rfun(~ t, seed=982)
tmp2 <- rfun(~ t, seed = 2932)
tmp3 <- rfun(~ t, seed = 43)
windspeed <- function(t) {
  abs(tmp((t - 5)*3) + tmp2((t - 10)*2) + tmp3((t - 15)*4))
}
speed2power <- function(s) {
  pmin(ifelse(s < 5, 0, (s-2)^3), 5000)
}
slice_plot(speed2power(s) ~ s, domain(s=c(5,5.5)), npts=500)
```

The function `windspeed(t)` records wind speed at the site of a wind-turbine farm over one day, that is, $0 \leq t \leq 24$ hours. The function `speed2power(s)` is the production function for the model of wind turbine used at the farm: the input is speed in miles per hour, the output is in kilowatts. (Both these functions were created for this exercise. They are not about a real turbine at a real wind farm, but are somewhat realistic.) Hint: you can nest a function inside of another function. For instance, if I had a function ('solarpanelpower') that calculates the amount of power a solar panel generates and another function ('sunlight') that tells me the amount of sunlight at time of the day ('TOD'). I could evaluate this in one step like the following: 'solarpanelpower(sunlight(TOD))'. This would give me the amount of power from the solar panel based upon the time of the day.

Your task, find the *total energy* generated over the 24-hour period by the turbine. Reminder: energy $E$ is electric power multiplied by time. Or, more usefully for this problem, the increment energy $dE$ generated at time $t$ is the product of power at time $t$ multiplied by the increment of time $dt$, that is, $dE = p(t) dt$. Consequently, $$E = \int_\text{morning}^\text{night} p(t) dt$$ where "morning" should really be 00:00 h and night 24:00 h on the day in question. 

We don't have an algebraic formula for `windspeed(t)` even though it is a function. You can use `antiD()` to find the anti-derivative of the electric power function. 

The answer you compute should be saved to the name `result`. The units will be in kWh -- kilowatt hours.

```{r wind2, exercise = TRUE, exercise.nlines=6, exercise.cap="Find the energy produced"}
# ignore these definitions. They are setting up the 
# functions for you to use
tmp <- rfun(~ t, seed=982)
tmp2 <- rfun(~ t, seed = 2932)
tmp3 <- rfun(~ t, seed = 43)
windspeed <- function(t) {
  abs(tmp((t - 5)*3) + tmp2((t - 10)*2) + tmp3((t - 15)*4))
}
speed2power <- function(s) {
  pmin(ifelse(s < 5, 0, (s-2)^3), 5000)
}

#Your work starts here

slice_plot(windspeed(t) ~ t, domain(t=c(0, 24)))

# # Uncomment the next lines as you figure out how to fill in the "...blanks..."
# antid_of_power <- antiD( ....power_function_here(t)... ~ t)
# result <- antid_of_power(...night...) - antid_of_power(...morning...) 
# result # this prints out the result
```



```{r wind3, echo=FALSE}
askMC(
  "Wind turbines of this type have a maximum power rating of 5000 kilowatts. Was this rating exceeded at any point during the day?",
  "The maximum instantaneous power was about 3500 kilowatts",
  "The maximum instantaneous power was about 1100 kilowatts",
  "+That threshold was reached about 9 AM+",
  "That threshold was exceeded about 8 AM",
  "The maximum instantaneous power cannot be determined from the information given.",
  random_answer_order = FALSE
)
```

```{r wind4, echo=FALSE}
askMC(
  "At the maximum power rating of 5000 kilowatts, what's the theoretical maximum amount of energy produced by the turbine over a 24-hour day?",
  "+5000 * 24 kilowatt-hours+",
  "5000 / 24 kilowatt-hours" = "The units, kilowatt hours, should remind you that power is multiplied by time.",
  "5000 kilowatts" = "That's  the **power** rating. We want to know the amount of **energy**.",
  "Can't be determined from the information given."
)
```

```{r wind5, echo=FALSE}
askMC(
  "About what fraction of the theoretical maximum energy did the wind turbine generate over the 24-hour period?",
  "About 2.5%",
  "+About 10%+", 
  "About 25%",
  "About 50%",
  "Can't be determined from the information given.",
  random_answer_order = FALSE
)
```

```{r wind6, echo=FALSE}
askMC(
  "A peak time for residential energy consumption is from 7 am to 9 am. The price at which you can sell electrical energy to the grid operator is $0.09 per kilowatt-hour. At that price, how much would the energy produced from 7-9 am be worth?",
  "About 20 cents." = "That would be the answer if the turbine generated only 1 kilowatt of power during that period.",
  "About $150", 
  "+About $350+",
  "About $650",
  random_answer_order = FALSE
)
```

```{r wind7, echo=FALSE}
askMC(
  "What's the average wind speed over the 24-hour period?",
  "+About 5 mph+",
  "About 7 mph",
  "About 9 mph"
)
```

```{r wind8, echo=FALSE}
askMC(
  "Wind speed fluctuates a lot, but imagine that the wind blew steadily at the average wind speed from the previous problem. How much energy would be generated over the 24-hour period?",
  "+0 kilowatt hours+" = "That wind wouldn't be strong enough to turn the turbine.",
  "500 kilowatt hours",
  "1000 kilowatt hours",
  "10,000 kilowatt hours",
  random_answer_order = FALSE
)
```
