# Assembling functions {#sec-assembling}

```{r include=FALSE}
source("../starter.R")
```

When a modeler needs a new function for some purpose, there are three tactics. The best one depends on the circumstances and the availability of data. 

1. Assemble the function from combinations of input-scaled pattern-book functions.

This tactic is well represented in Calculus textbooks. I think this is not so much a statement about the power of the tactic, but the general tendency to avoid working with data. For the applications of applied math, data is a valuable resource. When it's available, as is often the case, there are alternatives to (1) that are available.

2. Use the data to guide the input and output scaling of pattern-book functions and their generalizations such as the power-law functions introduced in @sec-power-law-family. This is often called "**fitting the function**". We will study this tactic and its techniques in @sec-fitting-by-feature.

3. Split the domain of the desired function into small pieces. Then, for each piece, use very simple functions to capture the **local behavior** of a function locally. As the simple functions, the lower-order polynomaials `x ~ a x` or `x ~a x + b` or `x ~ a x^ + b x^2` are widely used. The parameters are set entirely by data. This strategy was originally called "statistical learning," then renamed "machine learning." These days, "artificial intelligence" is used to refer to a particular strategy of *iteration* that slowly and gently incorporates data to tease the model parameters into a function that makes good predictions.

This chapter focusses on the mathematical framework for (1). @sec-fitting-by-feature introduces the incorporation of data used in tactic (2). As for (3), we will need several core concepts of calculus as a basis for understanding, so we mainly defer the tactic to @sec-splines. 

## Linear combination 

@sec-output-scaling introduced the notation of input and output scaling. For instance, @fig-tides-ri2 illustrates how the pattern-book $\sin()$ function can be customized to model a particular situation. Recall that the input scaling involves inserting a straight-line function *inside* the parentheses, as in

$$f_1(t) \equiv \sin\left( a t + b\right)$${#eq-input-scaled-sin}

You may recognize $at + b$ as a straight-line function. Possibly, your recognition would be easier if we wrote $at + b$ using different names for the input and the first parameter: $m x + b$. You have been trained to pronounce $m$ as the "slope" of the line and $b$ as the "y-intercept."

Using both input and output scaling gives a more general kind of function:

$$f_2(t) \equiv A \sin(a t + b) + B$${#eq-input-output-scaled-sin}

@eq-input-output-scaled-sin is an example of a **linear combination** of two functions. The two functions are $g_1(t) \equiv 1$ and $g_2(t) \equiv sin(a t + b)$. 

It might be easier to see @eq-input-output-scaled-sin as a linear combination if the function were written explicitly using the two functions being combined, that is, as

$$f_2 \equiv A g_2(t) + B g_1(t)$$

The combination is made by scaling each of the functions involved then adding the scaled functions together. The two scaling factors, $A$ and $B$, could be called "parameters of $f_2()$," which indeed they are. It would be good to get used to another word that's used specifically for the parameters in a linear combination: "**coefficients**." The advantage of "coefficients" as a name lies in it marking the parameters as those involved in a linear combination, rather than any of the other ways parameters can be used. 

Note that "coefficients" always refers to parameters that are not inside the parentheses of a function. In contrast, often parameters are inside the parentheses as in @eq-input-scaled-sin. The parameters $a$ and $b$ in $\sin(a t + b)$ are *inside* the parentheses. Consequently, $a$ and $b$ shouldn't be called "coefficients." Sometimes, to emphasize this, parameters in parentheses are called "nonlinear parameters" to distinguish them from coefficients like $A$ and $B$. 


To illustrate how linear combination is used to create new functions, consider polynomials, for instance, 
$$f(x) \equiv 3 x^2 + 5 x - 2\ .$${#eq-poly-ex-1}
In high school, polynomials are often presented as puzzles---factor them to find the roots! In calculus, however, polynomials are used as functions for modeling. They are a kind of modeling "clay," which can be shaped as needed.

There are three pattern-book functions in @eq-poly-ex-1. In  polynomials the functions being combined are all power-law functions: $g_0(x) \equiv 1$, $g_1(x) \equiv x$, and $g_2(x) \equiv x^2$. With these functions defined, we can write the polynomial $f(x)$ as $$f(x) \equiv 3 g_2(x) + 5 g_1(x) - 2 g_0(x)$$ Each of the functions is being scaled by a quantity: 3, 5, and -2 respectively. Then the scaled functions are added up. That is a linear combination; scale and add. 



Linear combination is an extremely important tactic that quantitative workers use throughout their careers. For instance, many physical systems are described by linear combinations. For instance, the motion of a vibrating molecule, a helicopter in flight, or a building shaken by an earthquake are described as simple "modes" which are linearly combined to make up the entire motion. More down to Earth, the timbre of a musical instrument is set by the scalars in a linear combination of pure tones. And throughout work with data in science, commerce, government and other fields a primary data analysis method---called "regression"---is about finding the best linear combination of a set of explanatory variables to create a model function of the response variable.

::: {.callout-note}
## Identical vs proportional vs straight-line

One of the pattern book functions is very simple; the output is identical to the input:

$$\text{identity}(x) \equiv x$$
Multiplying the identity() function by a parameter gives a function that can well be called "proportional()."

$$\text{proportional}(x) \equiv a\ \text{identity}(x) = a\ x$$
The parameter $a$ is often called the "**constant of proportionality**." 

It's common to call this closely related function the "linear" function, but a better name is the "straight-line" function. "Straight-line" is the name we shall use in this book.

$$\text{straight\_line}(x) \equiv a\ x + b$$
:::

## Function composition {#sec-function-composition}

**Function composition** refers to combining functions by taking the input of one function and feeding it as input to another.  "$g()$ composed with $h()$" means $g(h(x))$. 

To illustrate, consider again the function defined in @eq-input-output-scaled-sin:
$$f(t) \equiv A \sin\left( a t + b\right) + B$$

You've already seen how @eq-input-output-scaled-sin is a linear combination of two functions $f_1(t) \equiv 1$ and $f_2(t) \equiv sin(a t + b)$. But $f_2()$ is not a linear combination. Instead, it is a function composition. The two functions being composed are $sin(x)$ and $a t + b$, producing $sin(a t + b)$. Here, $\sin()$ is the **outer** function in the composition and $at + b$ is the **inner** function.

In function *composition*, the order of the functions matters: $f(g(x))$ and $g(f(x))$ are in general completely different functions.

## Function multiplication {#sec-function-multiplication}

Multiplication is the third in our repertoire of methods for making new functions. To review, here are the first two methods involving combining a function $f()$ with a function $g():

1. Linear combinations. Example: $n_1(t) \equiv 5 f(t) + 1.2 g(t)$.
2. Function composition. There are two possibilities which produce two distinct functions:
    i. $c_1(t) \equiv f(g(t))$, that is, $g()$ is the *inner function*.
    ii. $c_2(t) \equiv g(f(t))$, that is, $g()$ is the *outer function*.

Now a new method:

3. Multiplication of the output of two functions. Example: $m(t)\equiv f(t) g(t)$. This is perfectly ordinary multiplication. Multiplication is *commutative*, as illustrated by the equality of $5 \times 2$ and $2 \times 5$. Owing to the commutativity of multiplication, it doesn't matter whether $f()$ is first or $g()$ is first.


$$\underbrace{f(x) \times g(x)}_\text{multiplication}\ \ \ \ \underbrace{{\Large f(}g(x){\Large)} \ \ \text{or}\ \ \ {\Large g(}f(x){\Large)}}_\text{composition}$$

In function composition, only one of the functions---the ***interior function*** is applied to the overall input, $x$ in the above example. The ***exterior function*** is fed its input from the output of the interior function. 

In multiplication, each of the functions is applied to the input individually. Then their outputs are multiplied to produce the overall output.

::: {#thm-rise-and-fall style="display: none;"}
---What goes up may come down.
:::

::: {.callout-note icon=false data-latex=""} 
## @thm-rise-and-fall Modeling rise and fall

The initial rise in popularity of the social media platform [Yik Yak](https://en.wikipedia.org/wiki/Yik_Yak) was exponential. Then popularity leveled off, promising a steady, if static, business into the future. But, the internet being what it is, popularity collapsed to near zero and the company closed.

One way to model this pattern is by multiplying a sigmoid by an exponential. (See @fig-yikyak.)

::: {#fig-yikyak layout-ncol=2}
```{r echo=FALSE}
#| fig-subcap: 
#| - A sigmoid (orange) and an exponential function, shifted to be centered in mid-2014
slice_plot(pnorm(year, mean=2014.5, sd=0.7) ~ year, domain(year = 2012:2018), color="orange") |>
  slice_plot(exp(-(year-2014.5))/10 ~ year, color="blue")
```

```{r echo=FALSE}
#| fig-subcap:
#| - Multiplying the sigmoid and exponential produces a hump.
yikyak <- makeFun(
  pnorm(year, mean=2014.5, sd=0.7) * exp(-1.6*(year-2014.5)) ~ year)
slice_plot(yikyak(year) ~ year, domain(year=2012:2018))
```

Building a model of a steep rise and gentler fall by multiplying a sigmoid by an exponential.  Subscriptions to the web messaging service Yik Yak grew exponentially in 2013 and 2014, then collapsed. The company closed in 2017.
:::
:::

::: {#thm-transient-vibration style="display: none;"}
---Sounds of short duration can be modeled by multiplying sines by a local function such as the gaussian.
:::
::: {.callout-note}
## @thm-transient-vibration Transient vibration

A guitar string is plucked to produce a note. The sound is, of course, vibrations of the air created by vibrations of the string. 

After plucking, the note fades away. An simple model of this is a sinusoid (of the correct period to correspond to the frequency of the note) times a gaussian.



::: {#fig-wave-packet layout-ncol=2}
```{r echo=FALSE}
#| fig-subcap: 
#| - "The two components of the wave packet: an envelope and an oscillation."
#| out-width: "140%"
sinusoid <- makeFun(sin(2*pi*x/P) ~ x, P=10/5)
envelope <- makeFun(2*dnorm((x - center)/width) ~ x, center=3, width=4)
Ptop <- slice_plot(envelope(x) ~ x, bounds(x = c(-15, 15)), color="dodgerblue") %>%
  gf_labs(title="envelope")
Pbottom <- slice_plot(sinusoid(x) ~ x, bounds(x = c(-15, 15)),
           color="orange3", npts=500) %>%
  gf_labs(title="oscillation")
gridExtra::grid.arrange(Ptop, Pbottom, ncol=1)
```

```{r echo=FALSE}
#| fig-subcap: 
#| - "The wave packet constructed by multiplication"

wave_packet <- function(x, A=1, center = 0, width = 5, P=1, phase = 0) {
  A*sin(2*pi*x/P)*dnorm((x - center)/width)
}
slice_plot(wave_packet(x, A=2, center=3, width  = 4, P  = 10/5 ) ~ x, 
           bounds(x = c(-15, 15)), npts = 500)
```

A *wave packet* constructed by multiplying a sinusoid and a gaussian function.
:::

:::

Function multiplication is used so often in modeling that you will see it in many modeling situations. Here's one example that is important in physics and communication: the ***wave packet***. Overall, the wave packet is a localized oscillation as in @fig-wave-packet. The packet can be modeled with the product of two pattern-book functions: a gaussian times a sinusoid.

## Splitting the domain {#sec-piecewise-intro}

Consider the familiar absolute-value function, written
$$\text{abs}(x) \equiv \left|x\right|$$ in traditional notation. Written this way, the definition of abs() is a tautology: unless you already know what $\left|x\right|$ means, you will have no clue what's going on.

Can we assemble abs() out of pattern-book functions? What's distinctive about $abs(x)$ is the break at $x=0$. There is no similarly sharp transition in any of the pattern-book functions.

Defining a function piecewise, that is, constructing it from two different functions on two non-overlapping domains, provides the means to create a sharp transition. For the absolute value function, one domain is the negative inputs: $x < 0$. The other domain is the positive inputs: $0 \leq x$. Note that the domains do not overlap.

{{< include absolute-value.qmd >}}

A function defined separately on different pieces of its domain is called a ***piecewise function***. In the conventional mathematical notation, there is a large $\LARGE\left\{\right.$ followed by two or more lines. Each line gives a formula for that part of the function and indicates to which interval the formula applies.



{{< include heaviside.qmd >}}


The Heaviside function is defined on the same two pieces of the number line as $abs()$. To the right of zero, Heaviside is identical to constant(). To the left, it is identical to 0 times constant$()$. 

The vertical gap between the two pieces of the Heaviside function is called a ***discontinuity***. Intuitively, you cannot draw a discontinuous function ***without lifting the pencil from the paper***. The Heaviside's discontinuity occurs at input $x=0$.

{{< include ramp.qmd >}}

### Computing notation

The usual mathematical notation for piecewise functions, spread out over multiple lines that are connected with a tall brace, is an obvious non-candidate for computer notation. In R, the stitching together of the two pieces can be done with the function `ifelse()`. The name is remarkably descriptive. The `ifelse()` function takes three arguments. The first is a question to be asked, the second is the value to return if the answer is "yes," and the third is the value to return for a "no" answer. 

To define abs() or Heaviside() the relevant question is, "Is the input on the right or left side of zero on the number line?" In widely-used computing languages such as R, the format for asking a question does not involve a question mark. For example, to ask the question, "Is 3 less than 2?" use the expression: 

```{webr-r}
3 < 2
```

In mathematics notation, $3 < 2$ is a *declarative* statement and is an impossibility. More familiar would be $x < 2$, which is again a declarative statement putting a restriction on the possible values of the quantity $x$.

In computing notation, `3 < 2` or `x < 2` is not a declaration, it is an *imperative* statement that directs the computer to do the calculation to find out if the statement is true or false for the particular values given.


Here's a definition of `Heaviside()` written with `ifelse()`.

```{webr-r}
Heaviside <- makeFun(ifelse(0 <= x, 1, 0) ~ x)
```

Since Heaviside() is a function, the question $0 \leq x$ does not get answered at the time the function is defined. Instead, when the function is evaluated we will know the value of input $x$, can answer the $0 \leq x$ question, and use the answer to select either 1 or 0 as the output.

@tbl-R-questions shows computer notation for some common sorts of questions.  

::: {#tbl-R-questions}
R notation              | English
------|:--------------
`x > 2`      | "Is $x$ greater than 2?"
`y >= 3`     | "Is $y$ greater than or equal to 3?"
`x == 4`     | "Is $x$ exactly 4?"
`2 < x & x < 5`| "Is $x$ between 2 and 5?" Literally, "Is $x$ both greater than 2 and less than 5?"
`x < 2 | x > 6` | "Is $x$ either less than 2 or greater than 6?"
`abs(x-5) < 2` | "Is $x$ within two units of 5?"

: Each of these imperative statements in R asks a question about numbers.
:::


::: {#thm-heating-with-gas style="display: none;"}
---Two uses of gas---heating and cooking---call for a piecewise function.
:::
::: {.callout-note icon=false data-latex=""} 
## @thm-heating-with-gas Heating with gas

@fig-gas-use-two is a graph of monthly natural gas use in the author's household versus average temperature during the month. (Natural gas is measured in cubic feet, abbreviated *ccf*.) 

::: {#fig-gas-use-two}
```{r echo=FALSE}
ramp <- makeFun(ifelse(0 < x, x, 0) ~ x)
gf_point(ccf ~ temp, data = Home_utilities, alpha=0.5) %>%
  gf_labs(title="Household natural gas use", x = "Average temperature for the month (deg. F)", y = "Volume of gas used (cubic feet)") %>%
  gf_lims(x = c(0, 85)) %>%
  slice_plot(4.3*ramp(62 - temp) + 15 ~ temp, color="dodgerblue", size=2, alpha=0.5)
```

The amount of natural gas used for heating the author's home varies with the outside temperature.
:::

The graph looks somewhat like a hockey stick. A sloping straight-line dependence of ccf on temperature for temperatures below $60^\circ$F and constant for higher temperatures.  The shape originates from the dual uses of natural gas. Gas is used for cooking and domestic hot water, the demand for which is more or less independent of outdoor temperature at about 15 ccf per month. Gas is also used for heating the house, but that is  needed only when the temperature is less than about $60^\circ$F.  

We can accomplish the hockey-stick shape with a linear combination of the ramp() function and a constant. The ramp function represents gas used for heating, the constant is the other uses of gas (which are modeled as not depending on temperature. Overall, the model is  $$\text{gas}(x) \equiv 4.3\,  \text{ramp}(62-x)  + 15\ .$$

Note that the input to ramp() is 62 - $x$. This *input scaling* (@sec-input-scaling) turns the ramp around so that it rises to the left. The transition between the flat and ramp sections occurs at $x=62$.

:::



## Computing outside the domain

Each of our pattern-book functions, with two exceptions, has a domain that is the entire number line $-\infty < x < \infty$. No matter how big or small is the value of the input, the function has an output. Such functions are particularly nice to work with since we never have to worry about the input going out of bounds. 

The two exceptions are:

1. the logarithm function, which is defined only for $0 < x$.
2. some of the power-law functions: $x^p$. 
    - When $p$ is negative, the output of the function is undefined when $x=0$. You can see why with a simple example: $g(x) \equiv x^{-2}$. Most students had it drilled into them that "division by zero is illegal," and $g(0) = \frac{1}{0} \frac{1}{0}$, a double law breaker. 
    - When $p$ is not an integer, that is $p \neq 1, 2, 3, \cdots$ the domain of the power-law function does not include negative inputs. To see why, consider the function $h(x) \equiv x^{1/3}$. 
    
::: {.callout-tip}
## Division by zero on the computer
It can be tedious to make sure that you are on the right side of the law when dealing with functions whose domain is not the whole number line. The designers of the hardware that does computer arithmetic, after several decades of work, found a clever system to make it easier. It is a standard part of such hardware that whenever a function is handed an input that is not part of that function's domain, one of two special "numbers" is returned. To illustrate:

```{r warning=FALSE}
sqrt(-3)
(-2)^0.9999
1/0
```

`NaN` stands for "not a number." Just about any calculation involving `NaN` will generate `NaN` as a result, even those involving multiplication by zero or cancellation by subtraction or division.^[One that does produce a number is `NaN^0`.] For instance:

```{r warning=FALSE}
0 * NaN
NaN - NaN
NaN / NaN
```

Division by zero produces `Inf`, whose name is reminiscent of "infinity." `Inf` infiltrates any calculation in which it takes part: 

```{r warning=FALSE}
3 * Inf
sqrt(Inf)
0 * Inf
Inf + Inf
Inf - Inf
1/Inf
```

To see the benefits of the `NaN` / `Inf` system let's plot out the logarithm function over the graphics domain $-5 \leq x \leq 5$. Of course, part of that graphics domain, $-5 \leq x \leq 0$ is not in the domain of the logarithm function and the computer is entitled to give us a slap on the wrists. The `NaN` provides some room for politeness. 
:::

