---
id: "bear-lay-plant"
created: "Sun Jul 25 12:57:19 2021"
global_id: "Qj0JAr"
topic: "Scale invariance"
---

Some of our pattern-book functions have a distinctive property called **scale invariance**. This means the graph of the function looks the same even when plotted on very different horizontal and vertical axes. The function $\ln(x)$ plotted on two different scales in @fig-log-scale-invariance shows that the graph of the function has practically the same shape on either scale. 

```{r fig.show="hold", echo=FALSE}
#| label: fig-log-scale-invariance
#| out-width: "50%"
#| fig-cap: "The logarithm function has the same overall shape even when plotted on domains of very different scales."
slice_plot(log(x) ~ x, domain(x=c(.1,10))) %>%
  gf_labs(title="Logarithm on domain [0 < x < 10]")
slice_plot(log(x) ~ x, domain(x=c(exp(8),1e6))) %>%
  gf_labs(title="Logarithm on domain 0 < x < 1000000")
```

@fig-square-invariance shows a power-law function, $g(x) \equiv x^2$, which is also scale invariant.

```{r fig.show="hold", echo=FALSE}
#| label: fig-square-invariance
#| out-width: "50%"
#| fig-cap: "The function $x^2$ shown on two very different domain scales has the same overall shape."
slice_plot(x^2 ~ x, domain(x=c(-10,10))) %>%
  gf_labs(title="Square(x) on domain -10 < x < 10")
slice_plot(x^2 ~ x, domain(x=c(-1000000,1000000))) %>%
  gf_labs(title="Square(x) on domain -1000000 < x < 1000000")
```

Other pattern-book functions are not scale invariant, for example $\sin(x)$.

```{r fig.show="hold", echo=FALSE}
#| label: fig-sin-invariance
#| out-width: "50%"
#| fig-cap: "The sin( ) function is not scale invariant."
slice_plot(sin(x) ~ x, domain(x=c(-10,10))) %>%
  gf_labs(title="sin(x) on domain -10 < x < 10")
slice_plot(sin(x) ~ x, domain(x=c(-100,100)), npts=1000) %>%
  gf_labs(title="sin(x) on domain -100 < x < 100")
```

In contrast to scale-invariant functions, some of our pattern-book functions have a ***characteristic scale***. This is a domain length over which the whole of a characteristic feature of the function is evident. Graphing on larger domains simply squashes down the characteristic feature to a small part of the graphic domain. For instance, in the $\sin()$ function the cycle is a characteristic feature. The cycle in the pattern-book sinusoid has a characteristic length of $2 \pi$, the length of the cycle. Consequently, the graph looks different depending on the length of the graphics domain in multiples of the characteristic length. You can see from @fig-sin-invariance that the graph on the domain $-10 < x < 10$, that is, about 3 times the characteristic scale, looks different from the graph on the larger domain that has a length 30 times the characteristic scale.

The output of the sigmoid function runs from 0 to 1 but reaches these values only asymptotically, as $x \rightarrow \pm \infty$. In defining a characteristic scale, it would be reasonable to look at the length of the domain that takes the output from, say, 0.01 to 0.99. In other words, we want the characteristic scale to be defined in a way that captures **almost** all the action in the output of the function. For a gaussian, a reasonable definition of a characteristic scale would be the length of domain where the output falls to about, say, 1% of its peak output.
 
```{r blp1-1, echo=FALSE, results="markup"}
askMC(
  "The gaussian (hump) function `dnorm()` has a characteristic scale. Which of these is a domain length that can encompass the characteristic shape of the gaussian?",
  0.1, 1, "+6+" = "The domain $-3 < x < 3$ supports practically everything.", 16, 256,
  random_answer_order = FALSE
)
```

```{r blp1-2, echo=FALSE, results="markup"}
askMC(
  "The sigmoid function `pnorm()` also has a characteristic scale. Which of these is a domain length that can encompass the characteristic shape of the sigmoid?",
  0.1, 1, "+6+", 16, 256,
  random_answer_order = FALSE
)
```

Throughout science, it is common to set a standard approach to defining a characteristic scale. For instance, the characteristic scale of an aircraft could be taken as the length of body. Gaussian and sigmoids are so common throughout science that there is a convention for defining the characteristic scale called the ***standard deviation***. For the pattern book gaussian and sigmoid, the standard deviation is 1. That is much shorter than the domain that captures the bulk of action of the gaussian or sigmoid. For this reason, statisticians in practice use a characteristic scale of $\pm 2$ or $\pm 3$ standard deviations. 
