---
id: "beech-tug-lamp"
created: "Sun Apr  3 17:42:10 2022"
global_id: "evn4OP"
skill: "Handling data, plotting data"
---


TURN THIS INTO AN EXERCISE FOR WORKING WITH DATA



Here is a small part of a data frame about the size of penguins in the Palmer Archipelago in Antarctica. `r mark(710)`

```{r echo=FALSE}
library(palmerpenguins)
set.seed(101)
knitr::kable(
  penguins %>% sample_n(10) %>%
    select(species, body_mass_g, flipper_length_mm, 
           bill_length_mm) %>%
    rename(body_mass = "body_mass_g", 
           flipper_length = "flipper_length_mm",
           bill_length = "bill_length_mm") %>%
    mutate(body_mass = body_mass/1000,
           flipper_length=flipper_length/10,
           bill_length = bill_length/10)
) %>% 
    kableExtra::kable_styling()
```

::: {.scaffolding  data-latex=""}
The output above shows what a data frame looks like. Now we need to look at how to use a data frame in an R command. The particular data frame shown just above has the name `penguins` in R. To refer to the data frame, just use the name.

Open a `r sandbox_link()` and give the name itself as a command. This simple command causes the data frame to be displayed on the screen.

```{r eval=FALSE}
penguins
```

You should see as a result a table that can be scrolled through interactively with all 344 rows of the `penguins` data frame.

*Looking* at a data frame, or part of one, is sometimes what's required. But in general you will use the name of the data frame of interest as an ***argument*** to an R function. Typically the argument will be named `data = ` as you'll see below.

The name `penguins` is very general and it wouldn't be at all surprising if some other penguin fan or researcher created a different set of data with the same name. R, like many other modern languages, makes room for such multiple uses of a name by a facility called a ***namespace***. R programmers can create their own namespace. This is not something we will do in this course. But it's helpful to know when the namespace mechanism is at work. For instance, the particular `penguins` we've provided with this course comes from a namespace called `palmerpenguins`. Thus, the "full" name of the data fr is `palmerpenguins::penguins`. You may see such name constructions occasionally in exercises in this course. You don't have to understand how namespaces work other than to know that when you are given the full name, you should use it, double colons and all.
:::

Before you can use `gf_point()` to plot a pair of columns from a data frame, you have to know the names of the columns. A useful function for this is `names()`. Similarly, to look at the first few rows, use the `head()` function. Try these commands in a sandbox:

```{r results="hide"}
names(penguins)
head(penguins)
```


A ***point plot*** displays two columns from a data frame. Each row in the data frame is represented by one point in the graphic. For instance:

```{r warning=FALSE}
#| label: fig-penguin-flipper
#| fig-cap: "Data from 344 penguins on flipper length (mm) and body mass (g)."
gf_point(flipper_length_mm ~ body_mass_g, 
         data=penguins) 
```

Each penguin has a specific mass and flipper length. If you focus your attention on the penguins with a mass of around 4 kg (that is, 4000 gm), you can see that the different penguins have a variety of flipper lengths. That's typical natural variability.


