---
id: "cow-type-kayak-new"
perm: "asevss"
title: "Cooking with alcohol"
author: "Daniel Kaplan"
---

```{r include=FALSE, context="setup"}
library(etude2)
library(mosaic)
library(mosaicCalc)
```



The Wikipedia entry on "Common Misconceptions" used to contain this item:

> *Some cooks believe that food items cooked with wine or liquor will be non-alcoholic, because alcohol's low boiling point causes it to evaporate quickly when heated. However, a study found that some of the alcohol remains: 25% after 1 hour of baking or simmering, and 10% after 2 hours.*

The modeler's go-to function type for events like the evaporation of alcohol is exponential: The amount of alcohol that evaporates would, under constant conditions (e.g. an oven's heat), be proportional to the amount of alcohol that hasn't yet evaporated. 

```{r CTK1-1, echo=FALSE}
askMC(
  prompt = "A) Assume that 25% of the alcohol remains after 1 hour? If the process were exponential, how much would remain after 2 hours?",
    "10%" = "That's what the study is reported to have found, but that's not consistent with an exponential process that decays to 25% after one hour",
    "25%" = "Exponentials decay to zero eventually, so don't expect things to stay still after one hour.",
    "+25% of 25%+" = "We know that 75% is eliminated over 1 hour, leaving 25%. The continuing exponential process will, over the next hour eliminate 75% of the amount at the start of that hour. So after hour 2 we'll have 25% of the amount we had at hour 1, which was 25% of the original amount.",
    "75%" = "That's how much was eliminated in the first hour, not how much remains after 2 hours.",
    "75% of 75%" = "In an exponential process, at any moment the rate of decay (e.g. 75% per hour) is a constant proportion of the amount that is still there. After one hour, there is 25% of the alcohol remaining. That will decay at a rate of 75% per hour. Over the next hour, we'll lose 75% of the original 25%, giving us 25% of the original amount.",
  random_answer_order = FALSE
)
```

```{r CTK1-2, echo=FALSE}
askMC(
  prompt = "B) What is the half-life of an exponential process that decays to 25% after one hour?",
    "15 minutes" = "This provides time for four halvings in one hour, which would leave $\\frac{1}{2} \\times \\frac{1}{2} \\times \\frac{1}{2} \\times \\frac{1}{2} = 1/16$ of the original not 1/4.",
    "+30 minutes+" = "This gives time for two halvings in one hour, bringing us to 25% as observed.",
    "45 minutes" = "Two halvings bring us down to 25%. At this rate, it would take 90 minutes to get down to 25%, not 60 minutes as observed.",
    "none of the above" = "",
  random_answer_order = FALSE
)
```

Let's change pace and think about the "10% after 2 hours" observation. First, recall that the amount left after $n$ halvings is 
$\text{amount.left}(n) \equiv \frac{1}{2}^n$ 
This is an exponential function with base 1/2.

You're going to carry out a ***guess-and-check*** procedure to find $n$ that gives $\text{amount.left}(n) = 0.10$.  

Open a sandbox and copy over the scaffolding, which includes the definition of the `amount.left()` function. We started with a guess of $n=0$, which is wrong.  Change the guess until you get the output 10%.

::: {.scaffolding  data-latex=""}
```{r results="hide"}
amount.left <- makeFun((1/2)^n ~ n)
amount.left(0)
```
:::

```{r CTK1-3, echo=FALSE}
askMC(
  prompt = "C) Use `amount.left()` as defined in the scaffolding to guess-and-check how many halvings it takes to bring something down to 10% of the original amount.",
  "2.58", "+3.32+" , "3.62", "3.94", "4.12",
  random_answer_order = FALSE
)
```

Another way to find the input $n$ that generates an output of 10% is to construct the ***inverse function*** to $\text{amount.left}()$. 

The computer already provides you with inverse functions for $2^n$ and $e^n$ and $10^n$. Their names are `log2()`, `log()`, and `log10()` repectively.  Using `log2()`, write a function named `log_half()` that gives the inverse function to $(1/2)^n$. Remember, the input to the inverse function corresponds to 10%; the output to the $n$.

::: {.scaffolding  data-latex=""}
```{r eval=FALSE}
log_half <- makeFun( - log2(...your stuff here ...) ~ x)
```
:::

```{r CTK1-4, echo=FALSE}
askMC(
  prompt = "D) The answer you got in part C) is the **number** of halvings needed to reach 10%. If this number occurs in 2 hours (that is, 120 minutes),  what is the half life stated in minutes.",
  30, 35, "+36+", 38, 42, 47,
  random_answer_order = FALSE
)
```

Suppose you compromise between the half-life needed to reach 25% after one hour and the half-life needed to reach 10% after two hours. Use, say, 33 minutes as the compromise half life. Using the sandbox, calculate how much would be left after 1 hour for this compromise half life, and how much left after 2 hours.

```{r CTK1-5, echo=FALSE}
askMC(
  prompt = "E) How much is left after 1 hour and after 2 hours when the half life is 33 minutes?",
    "+28% and  8%+",
    "31% and 4%",
    "30% and 9%",
    "27% and 9%"
)
```

