---
id: "horse-lend-bottle"
created: "Wed May 18 15:25:35 2022"
global_id: "jywRQQ"
---

For each of the following, plot out the function over an appropriate graphics domain to determine whether the statement is true or false. 

```{r drill-hlbv-1,  echo=FALSE, results='markup'}
askMC(
  random_answer_order = FALSE,
  prompt = 
r"(The function $$f(t)\equiv \sin(t)$$ has a **local max** of $$1$$ somewhere on the domain $$-\infty < t < \infty$$.)",
"+True+", "False")
```


```{r drill-hlbv-2,  echo=FALSE, results='markup'}
askMC(
  random_answer_order = FALSE,
  prompt = r"(The function $$g(x)\equiv e^x$$ has an **inflection point** on the domain $$-\infty < x < \infty$$.)",
  "True", "+False+")
```


```{r drill-hlbv-3,  echo=FALSE, results='markup'}
askMC(
  random_answer_order = FALSE,
  prompt = r"(The function $$g(x)\equiv e^x$$ is **monotonic** on the domain $$-\infty < x < \infty$$.)",

(X) True ( ) False

----

```{r drill-hlbv-4,  echo=FALSE, results='markup'}
askMC(
  random_answer_order = FALSE,
  prompt = r"(The function $$g(x)\equiv e^x$$ has a **vertical asymptote**.)",
"True", "+False+")
```


```{r drill-hlbv-5,  echo=FALSE, results='markup'}
askMC(
  random_answer_order = FALSE,
  prompt = r"(The function $$h(x)\equiv \ln(x)$$ has a **vertical asymptote**.)",
  "+True+", "False")
```

```{r drill-hlbv-6,  echo=FALSE, results='markup'}
askMC(
  random_answer_order = FALSE,
  prompt = r"(The function $$h(x)\equiv \ln(x)$$ has a **horizontal asymptote**.)",
 "True", "+False+")
```


```{r drill-hlbv-7,  echo=FALSE, results='markup'}
askMC(
  random_answer_order = FALSE,
  prompt = r"(The function $$\text{inv}(val)\equiv 1/val$$ has a **vertical asymptote**.)",
  "+True+", "False")
```

```{r drill-hlbv-8,  echo=FALSE, results='markup'}
askMC(
  random_answer_order = FALSE,
  prompt = r"(The function $$\text{inv}(val)\equiv 1/val$$ has a **horizontal asymptote**.)",
  "+True+", "False")
```

```{r drill-hlbv-9,  echo=FALSE, results='markup'}
askMC(
  random_answer_order = FALSE,
  prompt = r"(The function $$f(x)\equiv 1/x$$ is **monotonically decreasing** (i.e., monotonic AND decreasing) on the domain $$0<x<\infty$$. (You need only check in the interval $0<x\leq 100.$))",
  "+True+", "False")
```

```{r drill-hlbv-10,  echo=FALSE, results='markup'}
askMC(
  random_answer_order = FALSE,
  prompt = r"(The function $$\text{inv}(val)\equiv 1/val$$ is **continuous**.)",
  "True", "+False+")
```
