---
id: "hump-intro"
global-id: "IELWV"
title: "Hump function"
author: "Daniel Kaplan"
---

```{r include=FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
```

The gaussian function is implemented in R with `dnorm(x, mean, sd)`.  The input called `mean` corresponds to the *center* of the hump. The input called `sd` gives the *width* of the hump.

::: {.scaffolding  data-latex=""}
In a sandbox, make a slice plot of `dnorm(x, mean=0, sd=1)`. By varying the value of `width`, figure out how you could read that value directly from the graph. 
```{r eval=FALSE}
slice_plot(dnorm(x, mean=0, sd=1) ~ x,
           domain(x = c(-4, 4)))
```
:::

In the plot below, one of the double-headed arrows represents the `width` parameter. The others are misleading. 

```{r echo=FALSE}
Arrows <- data.frame(xleft = 0.5 + rep(0, 3), xright = 0.5 + c(1.5, 2.2, 2.8),
                     yleft = c(0.16, 0.09, 0.045))
slice_plot(dnorm(x, 0.5, 1.5) ~ x, domain(x = c(-4, 4))) %>%
  gf_segment(yleft + yleft ~ xleft + xright, data = Arrows,
             arrow = arrow(ends = "both",
                           length = unit(3, "mm")))
```

```{r daily-digital-08-QA5, echo=FALSE}
askMC(
  prompt = "Which arrow shows correctly the `width` parameter of the gaussian function in the graph with arrows?",
    "+top+" = "",
    "middle" = "",
    "bottom" = "",
    "none of them" = "",
  random_answer_order = FALSE
)
```


```{r daily-digital-08-QA6, echo=FALSE}
message <- "The `center` parameter is the argmax of the function."
askMC(
  prompt = "What is the value of `center` in the graph with arrows?",
  "-2" = message, "-1" = message, 
  "-0.5"=message, "0"=message, "+0.5+", "1"=message, "2"=message,
  random_answer_order = FALSE
)
```
