---
title: "Invertible functions"
author: "Daniel Kaplan"
---

```{r include=FALSE, context="setup"}
library(etude2)
```

In this class  we will almost exclusively be dealing with functions that take *quantities* as inputs and produce a single *quantity* as an output. (In Block 2, we'll start to  look at inputs and outputs that are *vectors*, that is, collections of quantities.)

But functions can take as inputs and produce as outputs things that are not quantities. For instance, words or labels (which, in statistics  are called *levels* of a  categorical variable).

Open up a new tab in your web browser and navigate to this web site:  `https://what3words.com/butterfly.fortune.clings`
(We ask you to  open a new tab so that you  don't lose your place in  the Daily Digital document.)

You reasonably think of this  as a URL, an address on  the web. But  actually, the web server is set  up  differently. The function name  is `https://what3words.com`. When you navigate there,  the function is being evaluated  for three inputs,  each of which is a word. In  our notation, we would write:

$$\text{what3words.com}(\text{"butterfly"}, \text{"fortune"}, \text{"clings"}) $$

The output of the function is a location, displayed on a map because that's convenient for human users who  are trying to find a location  specified  in the three-word system. (You might have to zoom out to  make sense of the location.  It happens to  be  the location  of Dr. Brilleslyper's  and Dr. Kaplan's offices,  which are adjacent. The three-word system doesn't have a mechanism for specifying  altitude.  Those offices are on  the 6th floor of the building.)

For using the system, you need to be able to calculate the *inverse function*, that is, turning location into the corresponding three input words.

```{r daily-digital-02-Essay1, echo=FALSE}
etude2::etudeEssay(
  prompt = "Is there an inverse function? Explain in everyday terms whether the web site provides a way to calculate an inverse."
)
```

```{r daily-digital-02-TF1, echo=FALSE}
askMC(
  prompt = "Does the order of the inputs make a difference? That is, is the output of the function different when the arguments are changed in order.",
  "+TRUE+", 
  "FALSE",
  random_answer_order = FALSE
  )
```

```{r daily-digital-02-Essay2, echo=FALSE}
etude2::etudeEssay(
  prompt = "Try various orders for the three inputs \"butterfly\", \"fortune\", \"clings\". Give a description of the output for those different combinations."
)
```

An important concept (that we have not yet encountered in this course) is *continuity*. Continuity means that a small change in the input produces a small change in the output.


```{r daily-digital-02-TF2, echo=FALSE}
askMC(
  prompt = "True or false: The forward function is continuous.  That is, a small change in the inputs (that is, the words) produces a small change in the location output",
  "TRUE",
  "+FALSE+",
  random_answer_order = FALSE
  )
```

```{r daily-digital-02-TF3, echo=FALSE}
askMC(
  prompt = "True or false: The inverse function is continuous.  That is, a small change in location leads to a small change in the three words.",
  "TRUE",
  "+FALSE+",
  random_answer_order = FALSE)
```



In the class lesson, we point out that a function with two or more quantities as inputs and a quantity as an output does not have an inverse. That is, each output cannot be associated with a *unique* input.

Yet the `what3words.com` system, to be useful in communicating location over a noisy channel such as a telephone, needs to have an  inverse. The next question is hard, and involves some  math  you  have not  studied, but give it your  best try:

```{r daily-digital-02-Essay3, echo=FALSE}
etude2::etudeEssay(
  prompt = "What is it about  the  `what3words.com` function that enables it to have an inverse even though it  is  a  function of three inputs?"
)
```

Finally, a note about discontinuity. The location `year.nibbles.glasses` is in USAFA. The location `year.nibble.glasses` is in Baja California. So even if the words are easier to transmit than the many digits of latitude and longitude, the system still leaves plenty of space for mischief and error. `what3words.com` would have been well advised to add a fourth word to their system, which would enable them to implement error correction.
