---
author: "Danny Kaplan"
date: 2020-06-03 
tags: [respiration, sinusoid, third]
id: lamb-talk-gloves
global-id: YELXG
---

```{r lamb-talk-gloves-setup, include = FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
library(math141Z)
```

A person breathes in and out roughly every three seconds. The volume $V$ of air in the person's lungs varies between a minimum of $2$ liters and a maximum of $4$ liters. Assume time $t$ is measured in seconds.

Remember that a full cycle of the sine wave $\sin(x)$ involves $x$ going  from its starting value to that value **plus** $2 \pi$.  

```{r lamb-talk-gloves-QA1, echo=FALSE}
wrong_amp <- "This varies between a minimum of 0 and a maximum of 2."
askMC(
  prompt = "Which of the following is the most appropriate of these models for $V(t)$?",
    "$V(t) \\equiv 2 \\sin \\left( \\frac{\\pi}{3} t \\right) + 2$" = wrong_amp,
"+$V(t) \\equiv \\sin \\left( \\frac{2\\pi}{3} t \\right) + 3$+" = "Good. In this class, we generally write the sine function like  $\\sin(2 \\pi t/P)$ which means that the overall argument to  the sine function will go from 0 to $2 \\pi$ when $t$ goes from 0 to $P$.",
"$V(t) \\equiv 2 \\sin \\left( \\frac{2\\pi}{3} t \\right)+ 2$" = wrong_amp,
"$V(t) \\equiv  \\sin \\left( \\frac{\\pi}{3} t \\right) + 3$" = "Right amplitude and baseline: the minimum will be 2 liters and the maximum 4 liters. But the period is wrong. Going from $t=0$ to  $t=3$ should produce a full cycle of the sine function. But here  the argument would go only  from 0 to $\\frac{\\pi}{3} 3 = $\\pi$. After 3 seconds, only half  a cycle has been  completed.",
  inline = FALSE,
  random_answer_order = FALSE
)
```




A respiratory cycle can be divided into two parts: inspiration and expiration. Please do an experiment. Using a clock or watch, breath with a total period of 3 seconds/breath, that is, complete one breath every three seconds. Once you  have practiced this and can do it without forcing either phase of breathing,  make a rough estimate of what fraction of the cycle is inspiration and what fraction is expiration. (The "correct/incorrect" answers here are right for most people. Your natural respiration might be different.)

```{r lamb-talk-gloves-QA3, echo=FALSE}
askMC(
  prompt = "Which is true?",
    "+Inspiration lasts longer than expiration+" = "",
    "Expiration lasts longer than inspiration" = "Maybe it is for you, but not for most people. Try breathing in while counting 1-2-3 then exhaling while counting 1-2-3-4-5-6. Likely,  that's not a very  natural  pattern for you.",
    "Inspiration and expiration each consume about  the same fraction of the complete  cycle." = "People can do this consciously by counting 1-2-3  for inspiration  and another 1-2-3 for expiration. This usually feels forced and unnatural.",
  inline = FALSE
)
```


