---
title: "MMAC ยง1.1 Exercise 16"
author: "YY"
runtime: shiny_prerendered
---

```{r include=FALSE, context="setup"}

etude2::load_learnr_safely()
# Calculus/graphics/data packages
library(mosaic)
library(mosaicCalc)
```

In Exercises 16 and 17, explain why the table of data does or does not define a function.  Also, if it is a function, state the domain and range.

**Exercise 16**


Average debt (D) in thousands of dollars at the end of the spring term in each year (Y) for bachelor's degree recipients attending public four-year colleges and universities who borrowed money to finance their education.

Y    | D
:----|:------
2001 |  20.4
2003 |  20.9
2005 |  21.5
2006 |  21.8

```{r daily-digital-09-TF1, echo=FALSE}
askMC(
  prompt = "TRUE or FALSE: The table defines a function D(Y).",
  "+TRUE+",
  "FALSE" = "Is there a unique output for each input?",
  random_answer_order = FALSE
  )
```


```{r daily-digital-09-Essay1, echo=FALSE}
etude2::etudeEssay(
  prompt = "What is the domain and range? (If the table does define a function.)"
)
```

**Exercise 17**

Price of gas (P) at different gas stations in Los Angeles, California on June 11-12, 2015 (D)


D    |   P
:----|:-----
6/11  | 3.79
6/11  | 3.50
6/12  | 3.79
6/12  | 3.49

```{r mmac-1_1-TF2, echo=FALSE, results="markup"}
askMC(
  prompt = "TRUE or FALSE: The table defines a function P(D).",
  "TRUE" = "Is there a unique output for each input?",
  "+FALSE+",
  random_answer_order = FALSE
  )
```


```{r mmac-1_1-Essay2, echo=FALSE, results="markup"}
etude2::etudeEssay(
  prompt = "What is the domain and range? (If the table does define a function.)"
)
```

