---
id: "pattern-book-descriptions"
author: "Daniel Kaplan"
global-id: "H2eu2"
skill: "Bk-Fun-103b"
---

```{r include=FALSE}

```

Answer these questions. You may want to refer to the graphs in  @fig-monomial-graphs through [-@fig-non-integer-graphs].

Choose from among these possibilities:

- **Flat**: not-concave
- **Down**: concave down on the whole domain
- **Up**: concave up on the whole domain
- **DU**: concave down on the left and up on the right
- **UD**: concave up on the left and down on the right
- **UDU**: concave down in the center and concave up on both flanks

```{r ndK-1, echo=FALSE, results="markup"}
askMC(
  "Which of these best describes the concavity of the gaussian function?",
  "Flat"= "If it curves, it is either concave up or down.",
  "Down"="In some places, but not in others.",
  "Up",
  "+UDU+",
  "DU" = "Look again",
  "UD",
  random_answer_order=FALSE
)
```

```{r ndK-2, echo=FALSE, results="markup"}
askMC(
  "Which of these best describes the concavity of the sigmoid function?",
  "Flat"= "If it curves, it is either concave up or down.",
  "Down"="In some places, but not in others.",
  "Up",
  "UDU",
  "DU" = "Look again",
  "+UD+"
)
```

```{r ndK-3, echo=FALSE, results="markup"}
askMC(
  "Which of these best describes the concavity of the second-order monomial $m_2(x) \\equiv x^2$?",
  "Flat"= "If it curves, it is either concave up or down.",
  "Down"="Is it a smile or a frown?",
  "+Up+",
  "UDU",
  "DU" = "Look again",
  "UD",
  random_answer_order=FALSE
)
