---
id: "pattern-book-descriptions"
author: "Daniel Kaplan"
permid: "H2eu2"
---

```{r include=FALSE}
library(etude2)
```

Answer these questions about the pattern-book functions. You can refer to the graphs in Figures \@ref(fig:monomial-graphs) through \@ref(fig:non-integer-graphs).

```{r ndK-1, echo=FALSE, results="markup"}
askMC(
  "Which of these best describes the concavity of the gaussian function?",
  "It's not concave."= "If it curves, it's either concave up or down.",
  "It's concave down."="In some places, but not in others.",
  "+It's concave down in the center and concave up on both flanks.+",
  "It's concave down on the left and concave up on the right" = "Look again"
)
```

```{r ndK-2, echo=FALSE, results="markup"}
askMC(
  "Which of these best describes the concavity of the sigmoid function?",
  "It's not concave."= "If it curves, it's either concave up or down.",
  "It's concave down."="In some places, but not in others.",
  "It's concave down on the left and concave up on the right." = "Look again",
  "+It's concave up on the left and concave down on the right.+"
)
```

```{r ndK-3, echo=FALSE, results="markup"}
askMC(
  "Which of these best describes the concavity of the second-order monomial $m_2(x) \\equiv x^2$?",
  "It's not concave."= "If it curves, it's either concave up or down.",
  "It's concave down."="Is it a smile or a frown?",
  "It's concave down on the left and concave up on the right." = "Look again",
  "+It's concave up everywhere.+"
)
