---
title: "Introduction to sandboxes"
author: "Daniel Kaplan"
---

```{r include=FALSE, context="setup"}

```

Computing is an essential part of the uses of calculus in today's world. This was not always the case. After all, the first calculus textbook was published in 1697, but modern computers only became widely available in the 1980s. 

You will notice that calculus concepts are presented both in traditional math notation and as workable computer operations presented in a mainstream computer language: R. An important technique for teaching and learning computing is to present ***scaffolding*** for computer commands. At first, the scaffolding may be complete, correct commands that can be cut-and-paste into a ***sandbox*** where the calculation will be carried out. Other times it will be left to the student to fill in some part that is  been left out of the scaffolding. For example, when we introduce drawing graphs of functions and the choice of a domain, you might see a scaffold that has blanks to be filled in:

::: {.scaffolding}
```{r eval=FALSE}
slice_plot( exp(-3*t) ~ t, bounds( --fill in domain-- ))
```
:::

You can hardly be expected at this point to make sense of any part of the above command, but soon you will. 
