---
title: "Arithmetic with functions"
author: "Daniel Kaplan"
global-id: "DLLVA"
---

```{r include=FALSE}

library(mosaic)
library(mosaicCalc)
```


Epidemics are often modeled as a series of "waves." Each wave involves different  people. Let's imagine two epidemic waves, one passing through the "blue" population  and a smaller one through  the "orange" population, like this:

```{r echo=FALSE}
f1 <- makeFun(1000 * pnorm(t, 50, 20) ~ t)
f2 <- makeFun(500 * pnorm(t, 125, 10) ~ t)
slice_plot(f1(t) ~ t, domain(t=c(0,200)), 
           color = "dodgerblue",  
           label_text = "wave1(t)") %>%
  gf_labs(y = "Cumulative cases",  x = "Day of year") %>%
  slice_plot(f2(t) ~ t, color="orange3", label_text  = "wave2(t)")
```



```{r daily-digital-09-QA15, echo=FALSE}
askMC(
  prompt = "Assuming that there were no way to distinguish between red and blue cases, the epidemic would look like just one curve. Which of the following curves would the overall epidemic look like? (The colors are just  to  help you  distinguish the different functions)",
    "A" = "The cumulative number of cases cannot possibly go down!",
    "B" = "The cumulative number of cases cannot possibly go down!",
    "C" = "You cannot see the separate waves in (C) even though the rise in the blue wave is distinct in  time from the rise in  the red wave. ",
    "+D+" = "Yes. Each wave adds another layer to the cumulative number of cases"
)
```


```{r echo=FALSE}
slice_plot(f1(t) + f2(t) ~ t, domain(t=c(0,200)), label_text="(D)") %>%
  slice_plot(2*f1(t) - 2*f2(t) ~ t, label_text="(B)",  color  = "gray") %>%
  slice_plot(f1(t-50) + 2*f2(t+20) ~ t, label_text="(C)",  color  = "green") %>% 
  slice_plot(1000 - f1(t) + f2(t) ~ t, color = "orange", label_text  = "(A)") %>%
  gf_labs(y = "Total  cumulative cases",  x = "Day of year")  

```

Each of the functions shown in the above graph is a "linear combination" of the red function and the blue function. The real total cumulative cases is "wave1(t)"  plus "wave2(t)". The other functions are made up and are irrelevant to the real epidemic. Still, we can ask what arithmetic combination of "wave1(t)" and "wave2(t)"  was used to create the made-up function.


```{r daily-digital-09-QA16, echo=FALSE}
askMC(
  prompt = "For made-up function (A) in  the graph, what was the linear combination?",
    "+1000 - wave1(t) + wave2(t)+",
    "2*wave1(t) - 3*wave2(t)" = "When t=0,  both wave1(t) and wave2(t) are zero, so the sum would also be zero. That is not  true in function (A)",
    "1000 + wave1(t) - wave2(t)" = "This  linear combination would rise from 1000 to 2000 (at time t=100) and  then  fall off to 1500 as the negative  wave2(t)  comes into play."
)
```
