---
title: "Arithmetic with functions"
author: "Daniel Kaplan"
---

```{r include=FALSE}

library(mosaic)
library(mosaicCalc)
```


Being able to add or multiply functions can enable us to  tell a somewhat complicated story. For example, the  graph shows a simple model of the hourly  temperature in from June through September

```{r}
temp <- makeFun(20*cos(2*pi*(t-120)/365) + 
                  10*sin(2*pi*t/1) + 50 ~ t)
slice_plot(temp(day) ~ day, domain(day=c(90, 180)),  npt=1000) %>%
  gf_labs(y = "Temperature  (F)", x = "Day of year")

```

You can see in the graph the slow seasonal change in  temperature represented by a sine  function with a period of 365 days, as well as the relatively rapid hour-by-hour variation over the course of 24 hours. There are two different mechanisms at  work, 1) the annual orbit of the Earth around the Sun  which creates the seasons,  2) the daily rotation of the Earth around its axis. Each mechanism is approximated by a simple function. The  overall  pattern can be represented by  a sum of functions.

We are going to use this strategy of modeling complex phenomena as a  sum of two or more simple functions.

**Problem 1** Epidemics are often modeled as a series of "waves." Each wave involves different  people. Let's imagine two epidemic waves, one passing through the "dodgerblue" population  and a smaller one through  the "orange3" population, like this:

```{r echo=FALSE}
f1 <- makeFun(1000 * pnorm(t, 50, 20) ~ t)
f2 <- makeFun(500 * pnorm(t, 125, 10) ~ t)
slice_plot(f1(t) ~ t, domain(t=c(0,200)), 
           color = "dodgerblue",  
           label_text = "wave1(t)") %>%
  gf_labs(y = "Cumulative cases",  x = "Day of year") %>%
  slice_plot(f2(t) ~ t, color="orange3", label_text  = "wave2(t)")
```



```{r daily-digital-09-QA15, echo=FALSE}
askMC(
  prompt = "Assuming that there were no way to distinguish between red and blue cases, the epidemic would look like just one curve. Which of the following curves would the overall epidemic look like? (The colors are just  to  help you  distinguish the different functions)",
    "A" = "The cumulative number of cases can't possibly go down!",
    "B" = "The cumulative number of cases can't possibly go down!",
    "C" = "You can't see the separate waves in (C) even though the rise in the blue wave is clearly distinct in  time from the rise in  the red wave. ",
    "+D+" = "Yes. Each wave adds another layer to the cumulative number of cases"
)
```


```{r echo=FALSE}
slice_plot(f1(t) + f2(t) ~ t, domain(t=c(0,200)), label_text="(D)") %>%
  slice_plot(2*f1(t) - 2*f2(t) ~ t, label_text="(B)",  color  = "gray") %>%
  slice_plot(f1(t-50) + 2*f2(t+20) ~ t, label_text="(C)",  color  = "green") %>% 
  slice_plot(1000 - f1(t) + f2(t) ~ t, color = "orange", label_text  = "(A)") %>%
  gf_labs(y = "Total  cumulative cases",  x = "Day of year")  

```

Each of the functions shown in the above graph is a "linear combination" of the red function and the blue function. The real total cumulative cases is "wave1(t)"  plus "wave2(t)". The other functions are made up and are irrelevant to the real epidemic. Still, we can ask what arithmetic combination of "wave1(t)" and "wave2(t)"  was used to create the made-up function.


```{r daily-digital-09-QA16, echo=FALSE}
askMC(
  prompt = "For made-up function (A) in  the graph, what was the linear combination?",
    "+1000 - wave1(t) + wave2(t)+",
    "2*wave1(t) - 3*wave2(t)" = "When t=0,  both wave1(t) and wave2(t) are zero, so the sum would also be zero. that is  not  true in function (A)",
    "1000 + wave1(t) - wave2(t)" = "This  linear combination would rise from 1000 to 2000 (at time t=100) and  then  fall off to 1500 as the negative  wave2(t)  comes into play."
)
```
