---
title: "Energy-Maneuverability Theory"
author: "Daniel Kaplan"
global-id: "JELCI"
date: "Dec 23, 2020"
---

```{r include=FALSE}
library(mosaic)
```

The "Energy-maneuverability Theory" (E-M) of aircraft performance was developed by renowned fighter pilot Col John Boyd and mathematician Thomas Christie in the 1960s. The theory posits that the available maneuverability of an aircraft is closely related to its *specific energy* $E_s$, that is, the kinetic plus potential energy of the aircraft divided by aircraft weight. To be highly maneuverable, an aircraft must be able to change it is specific energy rapidly in time. Let's call this ability the *specific power* (that is, power divided by mass), $P_s$. An aircraft with large $P_s$ is more maneuverable than one with small $P_s$.

An important formula in E-M Theory is $$P_s = \frac{T - D}{W} V$$ where $T$ is aircraft thrust, $D$ is drag, $W$ is weight, and $V$ is velocity. $(T-D)$, thrust minus drag, is the net forward force on the aircraft.

Recall these facts about the dimension of physical quantities: 

- Velocity has dimension $L^1 T^{-1}$ (e.g. meters per second)

- Acceleration has dimension $L^1 T^{-2} =$ [Velocity] $\times\  T^{-1}$ (e.g. meters per second-squared)
- Force has dimension $M \times\ $ [Acceleration]
- Energy has dimension [Force] $\times\  L$
- Power has dimension [Energy] $\times\  T^{-1}$.

```{r boyd-1, echo=FALSE, results="markup"}
askMC(
  "Which of the following is a correct dimensional formulation of power?",
  "+[Force][Velocity]+",
  "[Energy][Velocity]",
  "[Force] / [Velocity]",
  "[Energy] / [Velocity]"
)
```


```{r boyd-2, echo=FALSE, results="markup"}
askMC(
  prompt = "What is the dimension of $P_s$ in E-M Theory?",
  "+[Power] $\\times\\ M^{-1}$+" = "In other words, specific power, that is, power per mass.",
  "[Force] $\\times$ [Acceleration]" = "Just so you know, such a dimension is rarely, if ever, encountered in practice.",
  "[Force] $\\times$ [Velocity]" = "You're leaving out the division by $W$ in the E-M Theory formula.",
  "[Power]"
)
```


