---
title: "Boyle's data"
author: "Daniel Kaplan"
---

```{r include=FALSE}

library(mosaic)
library(mosaicCalc)

```






Open a sandbox to carry out some calculations with Boyle's data. To see how the data frame is organized, use the `head(Boyle)` and `names(Boyle)` commands. 

::: {.scaffolding  data-latex=""}
The scaffolding here contains a command for plotting out Boyle's data. It also includes a command, `gf_lm()` that will add a graph of the best straight-line model to the plotted points. Recall that the `#` symbol turns what follows on the line into a *comment*,  which is ignored by R. By removing the `#` selectively you can turn on the display of log axes.

```{r}
gf_point(pressure ~ volume, data = Boyle) %>%
  gf_refine(
     # scale_x_log10(),
     # scale_y_log10()
  ) %>%
  gf_lm()
```
:::


```{r daily-digital-06-QA5, echo=FALSE}
askMC(
  prompt = "In a sandbox, plot pressure versus volume using linear, semi-log, and log-log axes. Based on the plot, and the straight-line function drawn, which of these is a good model of the relationship between pressure and volume?",
    "linear" = "This would look like a straight line on linear axes.",
    "exponential" = "This would look like a straight line on semi-log axes.",
    "+power-law+"
)
```

