---
chapter: "none yet"
author: "Danny Kaplan"
difficulty: "XX"
date: 2021-06-08 
version: 0.1
tags: [first, second, third]
id: chicken-show-map
---

```{r chicken-show-map-default, include = FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
```


::: {.scaffolding  data-latex=""}
Open an R sandbox. You can use these function definitions to help you in your calculations. 
```{r}
f <- makeFun(2*exp(x+1) ~ x)
g <- makeFun(3*exp(-x) ~ x)
h <- makeFun(x*exp(x) ~ x)
```

:::


Using R, compute the *average* rate of change of the function over the given interval. Choose the closest answer for each problem.



```{r daily-digital-24-QA7, echo=FALSE}
askMC(
  prompt = "$f(x) \\equiv 2 e^{x+1}$ over [-2, 2]",
  -2.99, 1.54, 2.72, 4.68, "+9.85+", 11.32,
  random_answer_order = FALSE
)
```


```{r daily-digital-24-QA8, echo=FALSE}
askMC(
  prompt = "$g(x) \\equiv 3 e^{-x}$ over [-1, 1.5] ",
  "+-2.99+", 1.54, 2.72, 4.68, 9.85, 11.32,
  random_answer_order = FALSE
)
```

```{r daily-digital-24-QA9, echo=FALSE}
askMC(
  prompt = "$h(x) \\equiv x e^x$ over [0, 1]",
  -3.00, 1.54, "+2.72+", 4.68, 9.85, 11.32,
  random_answer_order = FALSE
)
```

It's much less work if we use the R function `c()` to define the interval, and the R function `diff()` to calculate differences. The next sandbox contains an example asking you to compute the average rate of change of $f(x) \equiv e^x$ over the interval [0, 4]. You only need lines 1, 3, and 5. The other lines show intermediate results to help you understand what `diff()` is doing.

::: {.scaffolding  data-latex=""}
```{r results="hide"}
interval <- c(0, 4)  # creates the interval
diff(interval)       # calculate the run
f <- makeFun(exp(x) ~ x)   #  create the function
f(interval)     # evaluate  function at the endpoints of  the interval
diff(f(interval)) / diff(interval)  # complete answer
```
:::

```{r daily-digital-24-TF7, echo=FALSE}
askMC(
  prompt = "True or false: The average rate of change of $f(x) \\equiv e^{x^2}$ over [0.0, 0.1] is 0.1005017",
  "+TRUE+",
  "FALSE",
  random_answer_order = FALSE)
```

```{r daily-digital-24-TF8, echo=FALSE}
askMC(
  prompt = "True or false: The average rate of change of $f(x) \\equiv \\log(x)$ over [2, 3] is 0.5062353.  
  (Hint: Change the code above so the interval variable goes from 2 to 3 and f becomes the function $f(x) \\equiv \\log(x)$)",
  "TRUE",
  "+FALSE+",
  random_answer_order = FALSE)
```

```{r daily-digital-24-TF9, echo=FALSE}
askMC(
  prompt = "True or false: The average rate of change of $f(x) \\equiv \\sin(x)$ over [0.0, 0.5] is 0.9588511",
  "+TRUE+",
  "FALSE",
  random_answer_order = FALSE)
```



