---
title: "R Graphics"
author: "Daniel Kaplan"
global-id: "KWLCie"
---

We've created a function named $\text{twins}(x,y)$ to help you practice making contour plots. You will need to open a sandbox to draw the plot. 


Here is some scaffolding for the command:

::: {.scaffolding  data-latex=""}
```{r}
twins <- mosaic::rfun(~ x + y, seed = 302, n=5)
contour_plot(twins(x, y) ~ x + y, bounds(x=c(0,1), y=c(-1,1)))
```
:::

```{r dr1-1, echo=FALSE, results="markup"}
askMC(
  "The domain of the plot should be large enough to show a mountain next to  a deep hole. Which of these domains will do the job?",
  "+`bounds(x=c(-5, 5), y=c(-5, 5)`+",
  "`bounds(x=c(1, 5), y=c(1, 5)`" = "This shows the mountain, but not the hole.",
  "`bounds(x=c(1,1), y=c(-1,1)))`" = "Some of the hole is shown, but none of the mountain.",
  "`bounds(x=c(5,10), y=c(0,10)))`" = "There is hardly anything going on in this domain. The function here is pretty flat except for a dip in the lower left." 
)
```

In a different sandbox (so you can still see the contour plot in the first sandbox), draw a **slice** through the function along the line $y=0$. Use the same $x$-domain as in the correct answer to the previous question.

In the `slice_plot()` command below, you will need to replace `__tilde-expression___` and `__domain__` with the correct syntax.

::: {.scaffolding
```{r, eval=FALSE}
twins <- mosaic::rfun(~ x + y, seed = 302, n=5)
slice_plot(__tilde-expression__, __domain__)
```
:::


```{r dr1-2, echo=FALSE, results="markup"}
askMC(
  "Which of these expressions will accomplish the task?",
  "+`slice_plot(twins(x, y=0) ~ x, bounds(x=c(-5,5)))`+",
  "`slice_plot(twins(x) ~ x, bounds(y=c(-5, 5)))`" = "The domain should be over $x$, not $y$. And `twins()` takes two inputs, even if one of them is fixed at zero.",
  "`slice_plot(twins(x, y=0) ~ x, bounds(x=c(-5, 5), y=c(-5, 5)))`" = "A slice plot has a domain that includes only one input.",
  "`slice_plot(twins(x, y=0) ~ x + y, bounds(x=c(-5, 5), y=c(-5, 5)))`" = "A slice plot has only one input on the right side of the tilde expression."
)
```
