---
id: "duck-see-socks"
created: "Wed Apr 20 13:14:35 2022"
global_id: "p6jxx5"
skill: "linear combinations"
---

PICK UP ON THIS IN ACCUMULATION BLOCK TO calculate the Gini Coefficient.

Income inequality is a matter of perennial political debate. In the US, most people support Social Security, which is an income re-distribution programming dating back almost a century. But other re-distribution policies are controversial. Some believe they are essential to a healthy society, others that the "cure" is worse than the "disease." 

Whatever one's views, it's helpful to have a way to quantify inequality. There are many ways that this might be done. A mathematically sophisticated one is called the ***Gini coefficient***.

Imagine that society was divided statistically into income groups, from poorest to richest. Each of these income groups consists of a fraction of the population and has, in aggregate, a fraction of the national income. Poor people tend to be many in number but to have a very small fraction of income. Wealthy people are few in number, but have a large fraction of income. The table shows data for US households in 2009:^[These data, as well as the general idea for the topic come from La Haye and Zizler (2021), "The Lorenz Curve in the Classroom", *The American Statistician*, 75(2):217-225]

group label  | population | aggregate income | cumulative income | cumulative pop.
:------------|-----------:|-----------------:|:-------
poorest      | 20%   | 3.4%  | 3.4%         | 20%
low-middle   | 20%   | 8.6%  | 12.0%        | 40%
middle       | 20%   | 14.6% | 26.6%        | 60%
high-middle  | 20%   | 23.2% | 47.8%        | 80%
richest      | 20%   | 50.2% | 100.0%       | 100%


The ***cumulative*** income shows the fraction of income of all the people in that group or poorer.  The cumulative population adds up the population fraction in that row and previous rows. So, a cumulative population of 60% means "the poorest 60% of the population" which, as the table shows, earn as a group 14.6% of the total income for the whole population.

A function that relates the cumulative population to the cumulative income is called a ***Lorenz function***. The data are graphed in @fig-lorenz-data and available as the `US_income` data frame in the `r sandbox_link()`. Later, in @fig-lorenz-one-fun, we'll fit parameterized functions to the data.

```{r echo=FALSE}
#| label: fig-lorenz-data
#| fig-cap: "Data on household incomes in the US in 2009."
#| fig-cap-location: margin
Income <- tibble::tribble(
  ~ income, ~ pop,
  0,     0,
  3.4,  20,
  12.0, 40, 
  26.6, 60,
  47.8, 80,
  100, 100
) %>%
  mutate(income=income/100, pop=pop/100)
P <- gf_point(income ~ pop, data = Income) %>%
  gf_labs(x="Population percentile (p)", y="Cumulative income percentile (L)") %>%
  gf_refine(coord_fixed())
P
```

Lorenz curves must:

- Be concave up, which amounts to saying that the curve gets steeper and steeper as the population percentile increases. (Why? Because at any point, poorer people are to the left and richer to the right.) 
- Connect (0,0) to (1, 1). 

Calling the income percentile $L$ a function of the population percentile $p$, a Lorenz function is $L(p)$ that satisfies the requirements in the previous paragraph.
Here are some functions that meet the requirements:

- $L_b(p) \equiv p^b$ where $1 \leq b$.
- $L_q(p) \equiv 1 - (1-p)^q$ where $0 < q \leq 1$

Notice that each of these functions has just one parameter. It seems implausible that the workings of a complex society can be summarized with just one number. We can use the curve-polishing techniques that will be introduced in @chap-fitting-polishing to find the "best" parameter value to match the data. 

```{r}
Lb <- fitModel(income ~ pop^b, data = Income, start=list(b=1.5))
Lq <- fitModel(income ~ 1 - (1-pop)^q, data = Income, start=list(q=0.5))
```

@fig-lorenz-one-fun compares the fitted functions to the data.

```{r echo=FALSE}
#| label: fig-lorenz-one-fun
#| fig-cap: "Lorenz curves $L_b(p)$ (blue) and $L_q(p)$ (magenta) fitted to the household income data."
#| fig-cap-location: margin
P %>%
  slice_plot(Lb(pop) ~ pop, color="dodgerblue", npts=500) %>%
  slice_plot(Lq(pop) ~ pop, color="orange3", npts=500) %>%
  slice_plot(.55*Lb(pop) + .45*Lq(pop) ~ pop, color="magenta")
```

Neither form $L_b(p)$ or $L_q(p)$ gives a compelling description of the data. Where should we go from here?

We can provide more parameters by constructing more complicated Lorenz functions. Here are two ways to build a new Lorenz function out of an existing one:

- The product of any two Lorenz functions, $L_1(p) L_2(p)$ is itself a Lorenz function.
- A linear combination of any two Lorenz functions, $a L_1(p) + (1-a) L_2(p)$, so long as the scalars add up to 1, is itself a Lorenz function. For instance, the magenta curve in @fig-lorenz-one-fun is the linear combination of 0.45 times the tan curve plus 0.55 times the blue curve.

Question: Is the composition of two Lorenz functions a Lorenz function? That is, does the composition meet the two requirements for being a Lorenz function?

To get started, figure out whether or not $L_1(L_2(0)) = 0$ and $L_1(L_2(1)) = 1$. If the answer is yes, then we need to find a way to compute the concavity of a Lorenz function to determine if the composition will always be concave up. We'll need additional tools for this. We'll introduce these in Block 2.

