---
id: "frog-win-fridge"
created: "Tue Aug  3 14:16:03 2021"
global_id: "mHACgC"
---

You are designing a pendulum for a planned joint NASA/ESA mission to Mars. From the orbital period and radius of Mars, its mass is known. From the mass and the observed diameter of the planet, gravitational acceleration at the surface is calculated as 3.721 m/s$^2$. According to Section @sec-pendulum-dimensions, the period is $\text{Period} = 2 \pi \sqrt{\frac{\text{Length}}{\text{Gravity}}}$.

The length of your pendulum is 3 feet.


```{r fwf8-1, echo=FALSE, results="markup"}
askMC(
  "What will be the period of your pendulum when it eventually gets to Mars? (Hint: Don't make the mistake of the engineers working on the [Mars Polar Lander](https://en.wikipedia.org/wiki/Mars_Polar_Lander) and forget to resolve the different units of length presented in the problem.)",
  "1.3 seconds",
  "1.9 seconds",
  "+3.1 seconds+",
  "9.1 seconds",
  random_answer_order = FALSE
  
)
```


```{r fwf8-2, echo=FALSE, results="markup"}
askMC(
  "What is the period of your pendulum on Earth?",
  "1.3 seconds",
  "+1.9 seconds+",
  "3.1 seconds",
  "9.1 seconds",
  random_answer_order = FALSE
)
```
