---
id: "frog-grow-pantry"
created: "Wed Sep  1 09:25:19 2021"
global_id: "VA9Dxi"
---


In this book, we are parameterizing the sinusoid using the ***period*** $P$, the duration a cycle. In many settings, such as communications engineering and physics, it is preferable to parameterize in terms of the ***frequency***, often written with the Greek letter $\omega$ ("omega").

Here's the relationship: $$\sin\left(\frac{2\pi}{P} t\right) = \sin(2\pi \omega t)$$

```{r fgp18-1, echo=FALSE, results="markup"}
askMC(
  "When the input quantity $t$ represents time, it has dimension T. The period P has the same dimension so that the overall argument to $\\sin()$ is dimensionless, as required. What is the dimension of $\\omega$?",
  "T",
  "T$^2$",
  "+T$^{-1}$+",
  "T$^{-2}$"
)
```

In an earlier exercise, we looked at human breathing. The period of a breathing cycle differs from hour to hour and from person to person. (It's also somewhat, but not completely, under conscious control.) A reasonable scale for the period of normal human breathing is 3 seconds. 


```{r fgp18-2, echo=FALSE, results="markup"}
askMC(
  prompt = "Given a respiratory period of 3 seconds/breath, what is the respiratory **frequency** in units of breaths/minute?",
    "+20 breaths/minute+" = "Right. Each breath takes 1/20th of a minute, which is 3 seconds, the period specified in the question.",
    "3 breaths/minute" = "If this were true, each breath would take 20 seconds to complete.",
    "1/3 breath per minute" = "With breaths completed every three seconds, 1/3 of a breath is completed each second. But the problem asked for breaths per minute.",
    "20 seconds per breath" = "The *period* is in the units of seconds per breath, but the *frequency* will have units of breaths per second. Frequency is the reciprocal of period (and vice versa).",
  inline = FALSE
)
```

