---
id: "lion-jump-kitchen"
created: "Fri Aug  6 18:29:01 2021"
global_id: "gmZiWh"
---

Recall Robert Boyle's data  on pressure and volume of a fixed mass of gas held at constant pressure. In Section @sec-magnitude-graphics of the text you saw a graphical analysis that enabled you to identify Boyle's Law with a power-law relationship between pressure and volume: $$P(V) = a V^n$$ On log-log axes, a power-law relationship shows up as a straight-line graphically. 

Taking logarithms translates the relationship to a straight-line function:$$\text{lnP(lnV)} = \ln(a) + n\,  \ln(V)$$
To find the parameter $n$, you can fit the model to the data. This R command will do the job:


```{r results="hide"}
fitModel(log(pressure) ~ log(a) + n*log(volume), data = Boyle) %>%
  coefficients()
```

Open a `r sandbox_link()` and run the model-fitting command. Then, interpret the parameters.


```{r daily-digital-06-QA6, echo=FALSE}
askMC(
  prompt = "What is the slope produced by `fitModel()` when fitting a power law model?",
    "Roughly -1" = "You must be a very precise person!",
    "+Almost exactly -1+" = "",
    "About -1.5" = "I'm not sure how you arrived at this answer.",
    "Slope $> 0$" = "You should be able to see from the graph you made in part (1) that the slope is negative.",
  random_answer_order = FALSE
)
```

According to the appropriate model that you found in (A) and (B), interpret the function you found relating pressure and volume.


```{r daily-digital-06-QA7, echo=FALSE}
askMC(
  "As the volume becomes very large, what happens to the pressure?",
    "+Pressure becomes very small.+" = "",
    "Pressure stays constant" = "You can see from the graph in part (A) that pressure does change with volume.",
    "Pressure also becomes large." = "You can see from the graph in part (A) that pressure goes down as volume goes up.",
    "None of the above" = "",
  inline = FALSE
)
```

Return to your use of `fitModel()` to find the slope of the straight-line fit to the appropriately log-transformed model. When you carried out the log transformation, you used the so-called "natural logarithm" with expressions like `log(pressure)`.  Alternatively, you could have used the log base-10 or the log base-2, with expressions like `log(pressure, base = 10)` or `log(volume, base = 2)`.  Whichever you use, you should use the same base for all the logarithmic transformations when finding the straight-line parameters.


```{r daily-digital-06-QA8, echo=FALSE}
askMC(
  prompt = "**(D)** Does the **slope** of the  straight line found by `fitModel()` depend on which base is used?",
    "+No+",
    "Yes" = "Did you use the same base for both logarithms in your `fitModel()` expression?",
    "There's no way to tell." = "Yes, there is. Try using `fitModel()` with the different bases of log."
)
```

```{r daily-digital-06-QA9, echo=FALSE}
askMC(
  "**(E)** Does the **intercept** of the  straight line found by `fitModel()` depend on which base is used?",
  "+Yes+" = "Good. But this will come out in the wash when you calculate the parameter $C$ in $C x^b$, since $C$ will be either $2^\\text{intercept}$ or $10^\\text{intercept}$ or $e^\\text{intercept}$ depending on the base log you use.",
  "No" = "Are you sure you tried different bases?",
  random_answer_order = FALSE
)
```

