---
chapter: "none yet"
author: "Danny Kaplan"
difficulty: "XX"
date: 2021-06-08 
version: 0.1
tags: [first, second, third]
id: lion-find-bowl
global-id: "HRTIX"
---

```{r lion-find-bowl-default, include = FALSE}
library(etude2)
```

For each exercise, you are given a series of intervals that get smaller and smaller. Your job is to calculate the *average rate of change* of the function $f(x) \equiv x^2$ for each of the intervals. As the width of the intervals approach zero, our *average* rates of change become better approximations of the *instantaneous* rate of change. You should use the results you calculate to make an informed *estimate* of the **instantaneous rate of change**.



::: {.scaffolding  data-latex=""}
```{r eval=FALSE}
interval <- c(__start__ , __end__ )
f <- makeFun(x^2 ~ x)
diff(f(interval)) / diff(interval)
```
:::


A. Use these three intervals to estimate the instantaneous rate of change $\partial_x f(x=3)$
    - [3, 3.1]
    - [3, 3.01]
    - [3, 3.001]

B. Use these three intervals to estimate the instantaneous rate of change $\partial_x f(x=5)$
    - [4.9, 5]
    - [4.99, 5]
    - [4.999, 5]
    
C. Use these three intervals to estimate the instantaneous rate of change $\partial_x f(x=-2)$
    - [-2, -1.9]
    - [-2, -1.99]
    - [-2, -1.999]


