---
id: "mag-blue"
author: "Daniel Kaplan"
global-id: "ILXEG"
---

```{r include=FALSE}

library(mosaic)
library(mosaicCalc)
```

Open a `r sandbox_link()` and make the following log-log plot of horsepower (`BHP`) versus displacement (in cc, cubic-centimeters) of the internal combustion engines listed in the `Engines` data frame. 

::: {.scaffolding  data-latex=""}
```{r results="hide"}
gf_point(BHP ~ displacement, data = Engines) %>%
  gf_refine(scale_x_log10(), scale_y_log10()) %>%
  gf_labs(x = "displacement (cc)")
```
:::

In the plot, you will see that the vertical axis has labels at 1, 10, 100, 1000, 10000. These numbers are hardly spaced evenly when plotted on a linear scale, but on the log scale they are evenly spaced. Since there is a factor of ten between consecutive labels, the interval between the labels is called a ***decade***. On the horizontal axis, the labels are at 10, 1000, 100,000, and 10,000,000. Each of those intervals spans a factor of one hundred. For instance, from 1000 is one-hundred times 10, 100,000 is one-hundred times 1000, and so on. An interval of size 100 is said to span ***two decades***, not 20 years but a factor of 100.

Based on the log-log plot, answer these questions.

```{r daily-digital-04-QA3, echo=FALSE}
askMC(
  prompt = "How many engines have a displacement of 1 liter or less?",
    "none" = "Perhaps you recognized that the left-most tick mark corresponds to a value of 1, and that no data points are 1 or smaller. But one liter corresponds to 1000 cc.",
    "7" = "This is the number of engines with displacement of 10 cc or smaller. But one liter corresponds to 1000 cc.",
    "+14+" = "Right. It is the $10^3$ tick that marks 1 liter, since 1 liter is 1000 cc.",
    "25" = "That would be true if the cut-off were 10 liters. But it is not.",
  random_answer_order = FALSE
)
```

```{r daily-digital-04-QA4, echo=FALSE}
askMC(
  prompt = "Using the log-log plot, how many decades of BHP are spanned by the data?",
     "4" = "Not a bad answer, but not the best one either. Notice that the smallest engine is about half a decade below 1 BPM, and the largest is  about half a decade above 10,000 BPH",
    "+5+" = "",
    "100" = "The number $10^{100}$ is called a *googol* and is roughly how many particles (including photons, neutrinos, etc.)  are in the universe. Imagine,  quite contrary to fact, that 1 BHP could be generated by burning one molecule of fuel per second. Then as many fuel molecules as there are particles in the universe would have  to  be burned each second to power an engine at the high end of a span of 100 decades.",
  random_answer_order = FALSE
)
```

