---
title: "multiplying functions"
author: "Daniel Kaplan"
global-id: "MWDKVA"
---

```{r include=FALSE}
library(mosaic)
library(mosaicCalc)
```

It is very common in communications systems to have a signal  that is  a sine-wave *carrier* that is *modulated* by  another function. That is,  the carrier is *multiplied*  by the other function: `waveform(t) = modulator(t)` $\times$ `carrier(t)`.

Here are four examples in each of which a sine with a period of about 2 units is modulated by another  function. In each case, that other function is one of the modeling functions we have studied.

```{r echo=FALSE}
slice_plot(exp(-t/10) * sin(2*pi*t/2.3) ~ t, 
           domain(t=c(-10, 30)),
           npts = 300) %>%
  gf_labs(title = "(A)")
```

```{r daily-digital-09-QA11, echo=FALSE}
askMC(
  prompt = "(A) Of what sort is the function  that is  modulating  the carrier?",
    "+exponential+",
    "gaussian",
    "sigmoid",
    "sine"
)
```


```{r echo=FALSE}
slice_plot(dnorm(t, 10, 7) * sin(2*pi*t/2) ~ t,
           domain(t=c(-10, 30)), npts = 500)  %>%
  gf_labs(title = "(B)")
```
```{r daily-digital-09-QA12, echo=FALSE}
askMC(
  prompt = "(B) Of what sort is the function  that is  modulating  the carrier?",
    "exponential",
    "+gaussian+",
    "sigmoid",
    "sine"
)
```



```{r echo=FALSE}
slice_plot(pnorm(t, 10, 7) * sin(2*pi*t/2) ~ t,
           domain(t=c(-10, 30)), npts = 500) %>%
  gf_labs(title = "(C)")
```

```{r daily-digital-09-QA13, echo=FALSE}
askMC(
  prompt = "(C) Of what sort is the function  that is  modulating  the carrier?",
    "exponential",
    "gaussian",
    "+sigmoid+",
    "sine"
)
```


```{r echo=FALSE}
slice_plot(sin(2*pi*t/20) * sin(2*pi*t/1.5) ~ t,
           domain(t=c(-10, 30)), npts = 500)  %>%
  gf_labs(title = "(D)")

```
```{r daily-digital-09-QA14, echo=FALSE}
askMC(
  prompt = "(D) Of what sort is the function that is modulating  the carrier?",
    "exponential",
    "gaussian",
    "sigmoid",
    "+sine+"
)
```

