---
id: "pythagorus-two"
author: "Daniel Kaplan"
---

```{r include=FALSE}

library(mosaic)
library(mosaicCalc)
```

Recall the Pythagorean theorem: $C^2 = A^2 + B^2$. Let's write this as a function that takes as inputs the lengths of the two legs and produces as output the length of the hypotenuse.

$$\text{hypotenuse}(a, b) \equiv \sqrt{\strut a^2 + b^2}$$

This can be seen as a **composition** of a function $f(x) \equiv \sqrt{x}$ into a linear combination of square functions of different inputs: $g(a, b) \equiv a^2 + b^2$. 

```{r py2-1, echo=FALSE, results="markup"}
askMC(
  "What is the function $f()$?", 
  "$f(x) \\equiv a^2$" = "This wouldn't make sense. The output of $f()$ does not depend on its input.", 
  "+$f(x) \\equiv \\sqrt{x}$+",
  "$f(x) \\equiv x^2$" = "This is part of the interior function, not the outer function.",
  "$f(x) \\equiv +$" = "Sorry, but the output of the hypothenuse() function needs to be a quantity, and $+$ is not a quantity.  "
)

```

```{r py2-2, echo=FALSE, results="markup"}
message <- "This wouldn't make sense. The output of $f()$ does not depend on its input."
askMC(
  "What is the function $g()$?", 
  "$g(x) \\equiv a^2$" = message, 
  "$g(a) \\equiv x^2$"= message,
  "+$g(x) \\equiv x^2$+" = "Right, a power-law function. It does not matter what we call the input, so long as it is used consistently in the definition.",
  "$g(x) \\equiv a^2 + b^2$" = "Sorry, but the input name is $x$ and the formula on the right side of the tilde expression has $a$ and $b$ in it."
)

```
 
```{r py2-3, echo=FALSE, results="markup"}
askMC(
  "There are two functions in the linear combination. What are they?",
  "+$g(a)$ and $g(b)$+",
  "$f(x)$ and $g(x)$",
  "$f(b)$ and $g(b)$"
)
```

```{r py2-4, echo=FALSE, results="markup"}
askMC(
  "What are the scalars in the linear combination?",
  "+$1$ and $1$+",
  "$1$ and $-1$",
  "There are no scalars." = "I think I know what you're getting at. The tradition in mathematics is that when a scalar is $1$, we don't write it down. But still, the scalar is $1$."
)
```

::: {.why  data-latex=""}
You used $a$ and $b$ as the names of the inputs to hypotenuse(). Aren't $a$ and $b$ meant to be parameters, not input names?

You can use whatever you want as input names, so long as they are used consistently between the left and right sides of the $\equiv$. Names like $x$, $t$,  and $y$ scream out to be recognized as input names, so these are what we use most of the time. But the Pythagorean theorem is usually written using A, B, and C. I'm honoring that convention so that a human reader gets a cultural hint what the inputs stand for. I'm unapologetic about this, since you got fair warning by the left side of the $\equiv$ what the argument names are.
:::


  
