---
title: "Sigmoid intro"
author: "Daniel Kaplan"
global-id: "CKSLE"
skill: "Gi-Cal-404-ai" 
---

```{r include=FALSE}
library(mosaic)
library(mosaicCalc)
```

Gaussian functions and sigmoidal functions come in pairs. For  every possible sigmoid, there is a corresponding gaussian that gives, for each value of the input, the slope of the sigmoid. 

Each of the following graphs shows a sigmoid and a gaussian function. The two might or might not correspond to one another. That is, the output of the gaussian might be the slope of the sigmoid, or the gaussian might correspond to  some other sigmoid. Remember, you're comparing the *output* of the gaussian to the *slope* of the sigmoid.

For each graph, say whether the gaussian and the sigmoid correspond to one another. Is so, choose **correspond**. If not, choose one of these reasons why not:

- **center**: The peak of the gaussian does not occur at the same value of $x$ at which the sigmoid is steepest.
- **amplitude**: The numerical value of the output of the gaussian function is, for all $x$, much larger than the numerical value  of the *slope*  of the sigmoid.

```{r echo=FALSE}
slice_plot(pnorm(x, mean = 5, sd = 3) ~ x, domain(x=c(-5,15)),
           label_text = "sigmoid") %>%
  slice_plot(dnorm(x, mean = 5, sd = 3) ~ x, color = "dodgerblue",  label_text = "gaussian") %>%
  gf_labs(title  = "(A)")
```

```{r daily-digital-08-QA7, echo=FALSE}
askMC(
  prompt = "Graph (A)",
    "+correspond+" = "",
    "center" = "For what $x$ is the sigmoid the steepest? For what $x$ is the gaussian the highest?",
    "amplitude" = "Did you  calculate the numerical value  of  the slope of the sigmoid?",
  inline = FALSE,
  random_answer_order = FALSE
)
```


```{r echo=FALSE}
slice_plot(pnorm(x, mean = 4, sd = 3) ~ x, domain(x=c(-5,15)),
           label_text = "sigmoid") %>%
  slice_plot(dnorm(x, mean = 2, sd = 1) ~ x, color = "dodgerblue",  label_text = "gaussian") %>%
  gf_labs(title = "(B)")
```

```{r daily-digital-08-QA8, echo=FALSE}
askMC(
  prompt = "Graph (B)",
    "correspond" = "",
    "+center+" = "The gaussian peaks at about $x=2$ while the steepest part of the sigmoid is at about $x=4$",
    "The numerical value of the output of the gaussian function is much larger than the numerical value  of the *slope*  of the sigmoid." = "Did you  calculate the numerical value  of  the slope of the sigmoid?",
  inline = FALSE,
  random_answer_order = FALSE
)
```

```{r echo=FALSE}
slice_plot(pnorm(-x, mean = -5, sd = 3) ~ x, domain(x=c(-5,15)),
           label_text = "sigmoid") %>%
  slice_plot(-dnorm(x, mean = 5, sd = 3) ~ x, color = "dodgerblue",  label_text = "gaussian", label_x = 0.1) %>%
  gf_labs(title  = "(C)")
```


```{r daily-digital-08-QA9, echo=FALSE}
askMC(
  prompt = "Graph (C)",
    "+correspond+" = "",
    "center" = "For what $x$ is the sigmoid the steepest? For what $x$ is the gaussian the highest?",
    "amplitude" = "Did you  calculate the numerical value  of  the slope of the sigmoid?",
  inline = FALSE,
  random_answer_order = FALSE
)
```

