---
author: "Danny Kaplan"
difficulty: "XX"
date: 2020-07-19 
tags: [mortality, logs, third]
id: tiger-have-fork
global-id: "SELIX"
skill: "Gc-Fun-117 Gi-Fun-117"
---

```{r include = FALSE}
library(mosaic)
library(mosaicCalc)

```

The data frame `SSA_2007` comes from the [US Social Security Administration](https://www.ssa.gov/oact/STATS/table4c6.html) and contains mortality in the US as a function of age and sex. ("Mortality" refers to the probability of dying in the next year.) 

Run the following commands in your R console to see the organization of the data.

```{r results="hide"}
data(SSA_2007)
head(SSA_2007)
```


Once you understand the data organization, delete the previous command and make a plot of mortality versus age:

```{r results="hide"}
gf_point(mortality ~ age, color = ~ Sex, data = SSA_2007)
```

There is a *slight* mistake in the way the command is written,
so an error message will be generated. To figure out what's wrong, read the error message, check the variable names, and so on until you successfully make the plot.

```{r tiger-have-fork-QA1, echo=FALSE}
askMC(
  prompt = "What was the mistake in the plotting command in the above code box?",
    "+Variable names didn't match the ones in the data.+" = "",
    "The *tilde* in the argument `color = ~ sex`" = "The `color =` argument is right. The value being used, `~ sex`, is a one-sided formula and is used for things like color, shape, transparency, ....",
    "The data frame name is spelled wrong." = "No.",
    "There is no function `gf_point()`." = "No. `gf_point()` is one of the more commonly used plotting functions",
  inline=FALSE
)
```

**Essay**:
What's the obvious (simple) message about mortality and age shown by graphic?

Now you are going to use semi-log and log-log scales to look at the mortality data again. To do this, you will use the `gf_refine()` function. 

```{r eval=FALSE}
gf_point( __and_so_on__) %>%
  gf_refine(
    scale_y_log10(),
    scale_x_log10()
  )
```
Fill in the `__and_so_on__` details correctly and run the command in your sandbox.

As written, both vertical and horizontal axes will be on log scales. This may not be what you want in the end.

Arrange the plotting command to make a semi-log plot of mortality versus age. Interpret the plot to answer the following questions. Note that labels such as those along the vertical axis are often called "decade labels."

```{r tiger-have-fork-QA2, echo=FALSE}
message = "Hint: How much is the change between successive labels on the y axis?"
askMC(
  "The level of mortality in year 0 of life is how much greater than in year 1 and after?",
    "About twice as large." = message,
    "About five times as large" = message,
    "+About 10 times as large+" = "",
    "About 100 times as large" = message,
  inline=FALSE
)
```

```{r tiger-have-fork-QA3, echo=FALSE}
message = "Hint: Due to the nature of logs, a difference of half a decade corresponds to a change of $\\sqrt{10}$."
askMC(
  "Near age 20, the mortality of males is how much compared to females?",
    "Less than twice as large." = message,
    "+A bit more than three times as large+" = "",
    "About 8 times as large" = message,
    "About 12 times as large" = message,
  inline = FALSE
)
```

```{r tiger-have-fork-QA4, echo=FALSE}
askMC(
  prompt = "Between the ages of about 40 and 80, how does mortality change with age?",
    "It stays about the same." = "But the curve is sloping up, isn't it?",
    "It increases as a straight line." = "It would be fair to say this about the logarithm of mortality. But a straight line in log mortality means that mortality itself is increasing exponentially.",
    "+It increases exponentially.+" = "",
    "It increases, then decreases, then increases again." = "Interesting that you would say this when the function is monotonically increasing above age 30.",
  inline = FALSE
)
```

Remake the plot of mortality vs age once again, but this time put it on log-log axes. The sign of a power-law relationship is that it shows up as a straight line on log-log axes. 

```{r tiger-have-fork-QA5, echo=FALSE}
askMC(
  prompt = "Between the ages of about 40 and 80 is the increase in mortality better modeled by an exponential or a power-law process?",
    "Power-law" = "But it is hard to find any straight line on the log-log  plot.",
    "+Exponential+" = "Right. The graph is much closer to a straight line on semi-log scales than on log-log scales.",
    "No reason to prefer one or the other." = "One is much closer to a straight line than the other."
)
```
