---
id: "birch-dig-radio"
created: "Mon Sep 13 14:31:01 2021"
global_id: "jAOmME"
---



We would like to write a Taylor-series polynomial for $\dnorm(x)$. To do this, we will need the numbers
$$\dnorm(x_0),\ \ \partial_x \dnorm(x_0),\ \ \partial_{xx} \dnorm(x_0),\ \ \partial_{xxx} \dnorm(x_0), ...$$
Here are the functions for the derivatives

Order | formula for derivative
------|---------
0th   | $\dnorm(x)$
1st   | $\left[{\large\strut}-x\right]\dnorm(x)$
2nd   | $\left[{\large\strut}x^2 - 1\right]\dnorm(x)$
3rd   | $\left[{\large\strut}3x - x^3\right]\dnorm(x)$
4th   | $\left[{\large\strut}3\,x^2-x^4     \right]\dnorm(x)$
5th   | $\left[{\large\strut}3-6\,x^3 + x^5     \right]\dnorm(x)$
6th   | $\left[{\large\strut}-15\,x^2 + 10\,x^4-x^6     \right]\dnorm(x)$
7th   | $\left[{\large\strut}-15 + 45\,x^3-15\,x^5 + x^7     \right]\dnorm(x)$
8th   | $\left[{\large\strut}105\,x^2-105\,x^4 + 21\,x^6-x^8     \right]\dnorm(x)$

1. Write a Taylor polynomial function for $\dnorm(x)$ around $x_0=0$). To do this, you will have to evaluate each of the 9 formulas above at $x=0$, producing a number. For example, the first five formulas will give 0.3989, 0, -0.3989, 0, 1.1968. This would expand into the following Taylor function:

```{r}
T_at_0 = makeFun(0.3989 + 
                   0*(x-x0) +
                   - 0.3989*(x-x0)^2/factorial(2) + 
                   0*(x-x0)^3/factorial(3) +
                   1.1968*(x-x0)^4/factorial(4) ~ x, 
                 x0=0)
```
You will need to add in the 6th, 7th, and 8th terms.

2. Plot `T_at_0(x) ~ x` over the domain $0 \leq x \leq 2$, and pipe the result to a slice-plot of `dnorm(x) ~ x` so that you can compare the two plots.

3. Define another function `T_at_1()` which will be much like `T_at_zero()` but with `x0=1` and the coefficients changed to be the formulas evaluated at $x=1$.

4. Add a layer showing `T_at_1()` to your previous plot. 

Say over what domain `T_at_0()` is a good approximation to `dnorm()` and over what domain `T_at_1()` is a good approximation to `dnorm()`. Do the two domains overlap?

5. Write a piecewise function of this form:

```{r eval=FALSE}
T <- makeFun(ifelse(abs(x) < 0.5, 
                    T_at_0(abs(x)), 
                    T_at_1(abs(x))) ~ x)
```
Plot this out on top of `dnorm()` to show whether `T()` is a good approximation to `dnorm()`.

You could continue on to define `T_at_2()` and incorporate that into the piecewise `T()`, and so on, to construct an approximation to `dnorm()` that is accurate over a larger domain.

<!--

RECURSIVE CONSTRUCTION OF THE POLYNOMIALS that multiply dnorm(x) for the successive derivatives.

```{r eval=FALSE, echo=FALSE}
funp <- function(coefs, ...) {
  coefs <- unlist(c(coefs, list(...)))
  powers <- 0:(length(coefs)-1)
  res <- function(x) {
    sum(coefs * x^powers)
  }
  attributes(res) <- list(coefs = coefs)
  class(res) <- c(class(res), "polynomial")
  res
}
get_coefs <- function(f) {
  if (inherits(f, "polynomial")) attributes(f)$coefs
  else f
}
derivp <- function(f) {
  coefs <- get_coefs(f)
  if (length(coefs) == 1) coefs <- 0
  else coefs <- coefs[-1]
  funp(coefs * 1:length(coefs))
}
intp <- function(f, C=0) {
  coefs <- get_coefs(f)
  funp(c(C, coefs / 1:length(coefs)))
}
xp <- function(f) {
  coefs <- get_coefs(f)
  funp(c(0, coefs))
}
padp <- function(f, n) {
  coefs <- get_coefs(f)
  # pad the coefficients with trailing zeros

  if (n > length(coefs)) c(coefs, rep(0, n-length(coefs)))
  else coefs
}
addp <- function(f1, f2) {
  coefs1 <- get_coefs(f1)
  coefs2 <- get_coefs(f2)
  n <- max(length(coefs1), length(coefs2))
  padp(coefs1, n) + padp(coefs2, n)
}
nextp <- function(f) {
  coefs <- get_coefs(f)
  funp(addp(xp(-coefs), derivp(coefs)))
}
as_latex <- function(f) {
  coefs <- get_coefs(f)
  res = ifelse(coefs[1]==0, character(0), as.character(coefs[1]))
  for (k in 2:length(coefs)) {
    if (coefs[k] == 0) next
    else if (coefs[k] < 0) res = paste0(res, coefs[k], "\\,x^", k)
    else res <- paste0(res, " + ", coefs[k], "\\,x^", k)
  }
  gsub("^NA ?\\+ ?", "", res) %>%
    gsub("1\\\\,", "", .)
}
```

-->
