---
id: "Exponentials"
author: "Daniel Kaplan"
global-id: "RDWKW"
---

```{r include=FALSE, context="setup"}

library(mosaic)
library(mosaicCalc)
```

In this question, we ask you to estimate the slope from a graph of the function. But the function is **exponential**, so not a straight line. 

A fundamental idea in calculus is that even a function with a curved graph, if you zoom in closely around a given point, will look like a straight line. And you know how to calculate the slope of a straight line.

When the graph is curved, the slope will be different at different points along the graph. So there's not a single slope for the function. Still, we can talk about the "slope at a point."

One way to specify a point on a function's graph is to give the horizontal coordinate: the **input** to the function. But here we're going to give you the **output** of the function. So long as the function passes the ***horizontal-line test***, as the exponential does, specifying any particular output in the function's range *uniquely* identifies a corresponding input. 

Estimate the slope of the exponential function $g(x) \equiv e^x$ at several inputs, which we'll call $x_1$, $x_2$, $x_3$ and $x_4$. We won't give you numerical values for the $x_i$ points, but we will tell you the output of the function at each of those inputs. the values of $x$ where:

a. $g(x_1) = 1$
b. $g(x_2) = 5$
c. $g(x_3) = 10$
d. $g(x_4) = 0.1$

For each of (a)-(d), use a sandbox to plot the exponential function $e^x$ on a domain zoomed in  around around the appropriate value of $x_i$.  Then calculate the slope  of  the curve at that $x_i$.


```{r daily-digital-02-QA1, echo=FALSE}
# indicate correct choices with +_+ in the name of the list item.
askMC(
  prompt = "Using your answers for the slope at the points given in (a)-(d), choose the best answer to this question: What is the pattern in the slope as $x$ varies?",
    "+The slope at each value $x_i$ is the same as $e^{x_i}$.+" = "This property of the exponential function becomes important when describing a wide range of phenomena, from nuclear isotope decay to population growth.",
    "The slope at each value $x_i$ is the same as $x_i$." = "That would being saying the slope at $x_3$ is $\\approx 2.30$. Is that what you got?",
    "The slope at each value of $x_i$ is the same as $x_i^2$." = "That would being saying the slope at the $x_3$ is $\\approx 5.30$. Is that what you got?",
    "The slope at each value of $x$ is the same as $\\sqrt{x}.$" = "That would being saying the slope at the $x_i$ is $\\approx 1.52$. Is that what you got?"
)
```

