---
id: "fawn-hear-kayak"
created: "Mon Jul 19 18:05:58 2021"
global_id: "IlNSF0"
---

```{r}
source("www/polycomp.R")
```

We are all busy people these days and nobody has the time to compute an infinite number of coefficients to program the polynomial computer. So, we will have to use just a few coefficients. How many is enough?

For example, the coefficients for $e^x$ (around $x_0 = 0$) are $1, 1, 1/2, 1/6, 1/24, 1/120, 1/720, 1/5040, 1/40320, 1/362880, ...$ Factorials get big fast! 

The sandbox shows a simple comparison of the polynomial approximation to the actual exponential function. it plots the magnitude of the error $|e^x - p(x)|$ on a log-log scale.

```{r cis1-0, exercise=TRUE, exercise.cap="The polynomial computer", exercise.nlines=10, eval=FALSE}
p <- poly_comp(x0 = 0, 1, 1, 1/2, 1/6, 1/24, 1/120)
error_fun <- makeFun(abs(exp(x) - p(x)) ~ x)
slice_plot(error_fun(x) ~ x, bounds(x=c(0.01, 10))) %>%
  slice_plot(abs(exp(-x) - p(-x)) ~ x, color="orange3") %>%
       gf_refine(scale_y_log10(), scale_x_log10()) %>%
  gf_labs(y = "Absolute error")
```

Read the graph carefully to make sure you understand what is being displayed.

* it is easy to think about the "absolute error" scale as indicating the number of zeros *after* the decimal point in the size of the error. For example `1e-6` is 0.000001. The approximation has that much error or less for $|x| < 0.3$.
* we are using log-log scales because we anticipated that the error would be something like a power law. (Why? Because the first term omitted from the approximation will produce a function $a_{n+1} x^{n+1}$, a power law. And when a power-law function is plotted on log-log axes, it appears as a straight line) But we cannot graph negative values of $x$ on a log scale. So we are plotting both $p(x)$ and $p(-x)$ together. The error for negative $x$ values are plotted in red.



```{r cis1-1, echo=FALSE, results="markup"}
askMC(
  "For $x \\approx 2.0$, the magnitude of the error is, to judge from the graph, `1e-1`. When written as a decimal number, how many leading zeros are after the decimal point?",
  -2, -1, 0, 1, 2, right_one = 0,
  random_answer_order = FALSE
)
```

```{r cis1-2, echo=FALSE, results="markup"}
askMC(
  "For $x \\approx 10.0$, the magnitude of the error is, to judge from the graph, `1e4`. What is this when written as a decimal number?",
  10, 100, 1000, 10000, right_one = 10000,
  random_answer_order = FALSE
)
```

