---
chapter: "none yet"
author: "Danny Kaplan"
difficulty: "XX"
date: 2021-06-09 
version: 0.1
tags: [first, second, third]
id: finch-trim-kayak
global-id: ykels
---

Although we created an R function named `slopeFun()` for the purposes of demonstration, it's better to use the R/mosaic operator `D()` which calculates the derivative, sometimes using symbolic methods and sometimes using a finite-difference method.


As an example of the use of `D()`, here is some more R code that defines a function `f()` and finds $\partial_x f()$, calling it `d_f()`. Then a slice plot is made of both `f()` and `d_f()`. 

```{r}
f   <- makeFun(sqrt(exp(x)) - x^2 ~ x)
d_f <- D(f(x) ~ x)
slice_plot(f(x) ~ x, domain(x=c(0, 5))) %>%
  slice_plot(d_f(x) ~ x, color = "orange3")
```
:::

For each of the following functions, write a brief comparison of the function to it's differenced version. You can combine phrases such as "same shape", "different shape. larger in amplitude", "smaller in amplitude", "same period", "shorter period", "longer period", or whatever seems appropriate. For instance, for the original example in the sandbox, a reasonable comparison might be, "f() is concave down but Diff(f) is concave up."


```{r daily-digital-26-Essay1, echo=FALSE}
etude2::etudeEssay(
  prompt = "A. For the function $f(x) \\equiv 3 x$, compare $f()$ to $\\partial_x f$."
)
```

```{r daily-digital-26-Essay2, echo=FALSE}
etude2::etudeEssay(
  prompt = "B. For the function $f(x) \\equiv x^2$, compare $f()$ to $\\partial_x f$."
)
```

```{r daily-digital-26-Essay3, echo=FALSE}
etude2::etudeEssay(
  prompt = "C. For the function $f(x) \\equiv e^x$, compare $f()$ to $\\partial_x f$."
)
```

```{r daily-digital-26-Essay4, echo=FALSE}
etude2::etudeEssay(
  prompt = "D. For the function $f(x) \\equiv e^{-0.3 x}$, compare $f()$ to $\\partial_x f$."
)
```

```{r daily-digital-26-Essay5, echo=FALSE}
etude2::etudeEssay(
  prompt = "E. For the function $f(x) \\equiv \\sin(x)$, compare $f()$ to $\\partial_x f$."
)
```

```{r daily-digital-26-Essay6, echo=FALSE}
etude2::etudeEssay(
  prompt = "F. For the function $f(x) \\equiv \\sin(2 \\pi x)$, compare $f()$ to $\\partial_x f$)."
)
```

```{r daily-digital-26-Essay7, echo=FALSE}
etude2::etudeEssay(
  prompt = "G. For the function $f(x) \\equiv \\sin(\\frac{2 \\pi}{20} x)$, compare $f()$ to $\\partial_x f$)."
)
```


