---
id: "fish-hurt-ring"
created: "Sun Dec 19 09:43:01 2021"
global_id: "pseisj"
---

It is worth remembering the first couple of terms in the Taylor polynomial approximation of the sine and cosine functions.

\begin{eqnarray}
\sin(x) &\approx & x - \frac{1}{2\cdot 3} x^3\\
 
\cos(x) &\approx & 1 - \frac{1}{2} x^2$$
\end{eqnarray}

```{r fhrw9-1, echo=FALSE, results="markup"}
askMC("What is the order of the approximation to $\\sin(x)$ given above?",
      "First",
      "Second",
      "+Third+" = "The highest-order term in the polynomial is $x^3$. So, third-order.",
      "Fourth",
      random_answer_order = FALSE
      )
```

```{r fhrw9-2, echo=FALSE, results="markup"}
askMC("What is the order of the approximation to $\\cos(x)$ given above?",
      "First",
      "+Second+"  = "The highest-order term in the polynomial is $x^2$. So, second-order.",
      "Third",
      "Fourth",
      random_answer_order = FALSE
      )
```

Here are several different quantities to calculate involving limits. You will carry out the calculation by substituting the Taylor Polynomials given above, cancelling out $x$ terms as permitted, and examining if the resulting formula makes sense. If it does, calculate the numerical value of the limit.

Example: 
$$\lim_{x \rightarrow 0} \frac{1 - \sin(x)}{x^2} \ \ \text{becomes}\ \ \lim_{x \rightarrow 0} \frac{1 - x + x^3/6}{x} = \lim_{x \rightarrow 0} \left(\strut\frac{1}{x} - 1 + \frac{x^2}{6}\right)$$
The $-1$ and $x^2/6$ terms are perfectly sensible in the limit as $x \rightarrow 0$, but the $1/x$ term implies division by zero. There is no numerical value for the limit. 


$$\text{a.}\ \ \lim_{x\rightarrow 0}\frac{1 - \cos(x)}{x}$$
$$\text{b.}\ \ \lim_{x\rightarrow 0}\frac{1 - \cos(x)}{x^2}$$

$$\text{c.}\ \ \lim_{x\rightarrow 0}\frac{\sin(x)}{x}$$

