---
id: "fox-dig-room"
created: "Mon Jul 19 13:04:42 2021"
global_id: "zdrsLb"
---

In the next few weeks, we're going to be using four of our basic modeling functions very heavily:

1. exponential function
2. sine function (which it turns out is related to the exponential function)
3. cosine function (just like the sin if you shifted it backward in time by 1/4 period)
4. logarithm function (the inverse of the exponential)

These functions are all very intricately related to one another. It will help to see the relationships if we write each of them in a common form. To that end, we will write each as a linear combination of power-law functions with integer exponents. Each of these linear series involves an infinite number of power-law functions.

1. exponential function$$e^x = 1 + \frac{x}{1} + \frac{x^2}{2!} + \frac{x^3}{3!} + \frac{x^4}{4!} + \frac{x^5}{5!} + \cdots$$

2. sine function $$\sin(x) = 0 + \frac{x}{1!} + 0 - \frac{x^3}{3!} + 0 + \frac{x^5}{5!} + \cdots$$

3. cosine function $$\cos(x) = 1 + 0 - \frac{x^2}{2!} + 0 + \frac{x^4}{4!} + 0 + \cdots$$

4. "natural" logarithm$$\ln(1+x) = 0 + \frac{x}{1} - \frac{x^2}{2} + \frac{x^3}{3} - \frac{x^4}{4} + \frac{x^5}{5} + \cdots$$

We ask you to memorize each of these four infinite series. Since it's impractical to memorize an ionfinite number of things, we're going to give you a system so that there are only a small number of facts needed.

a. Each term in each series will be given an index $n = 0, 1, 2, 3, 4, 5, \cdots
b. Every term involves multiplying the power law function $x^n$ by a number. Except for the $\ln(1+x)$ function, that number is always $1/n!$ multiplied by a *sign*, which can be $-1, 0$, or $1$. 

Here are the patterns of the signs:

function  |x^0^|x^1^|x^2^|x^3^|x^4^|x^5^|x^6^| $\cdots$
----------|----|----|----|----|----|----|----|----
$e^x$     | +  |   +|   +|   +|   +|   +|   +| $\cdots$
$\cos(x)$ | +  |   0|   -|   0|   +|   0|   -| $\cdots$
$\sin(x)$ | 0  |   +|   0|   -|   0|   +|   0| $\cdots$
$\ln(1+x)$| 0  |   +|   -|   +|   -|   +|   -| $\cdots$

Notice that the signs of the $\cos(x)$ function cycle with a period of 4, so the sign of term $n+4$ equals the sign of term $n$. Same with the $\sin(x)$, but the signs are shifted one slot. 

The series all extend to $n\rightarrow\infty$. Consequently, it's not practical to use these polynomial expansions for *exact* calculations. Mathematicians call them *transcendental functions*. 

```{r trans1, echo=TRUE}
askMC(
  "Which of these functions has a 0 for the $x^0$ term of the series? Choose the most complete correct answer.",
  "$e^x$",
  "+$\\sin(x)$+",
  "$\\cos(x)$",
  "$\\sin(x)$ and $\\cos(x)$",
  "$\\sin(x)$ and $e^x$",
  "$\\cos(x)$ and $\\ln(x+1)$"
)
```

```{r trans2, echo=TRUE}
askMC(
  "Which of these functions has a 1 for the $x^0$ term of the series? (Choose the most complete correct answer.",
  "$e^x$",
  "$\\sin(x)$",
  "$\\cos(x)$",
  "$\\sin(x)$ and $\\cos(x)$",
  "$\\sin(x)$ and $e^x$",
  "+$\\cos(x)$ and $e^x$+"
)
```

```{r trans3, echo=TRUE}
askMC(
  "What's the first zero term in the expansion of $e^x + \\cos(x)$?",
  "$x^0$ term",
  "$x^1$ term",
  "+$x^2$ term+",
  "$x^3$ term",
  "$x^4$ term",
  "$x^5$ term",
  random_answer_order = FALSE
)
```

```{r trans4, echo=TRUE}
askMC(
  "What's the first negative term in the expansion of $\\sin(x) + \\cos(x)$?",
  "$x^0$ term",
  "$x^1$ term",
  "+$x^2$ term+",
  "$x^3$ term",
  "$x^4$ term",
  "$x^5$ term",
  "there are none",
  random_answer_order = FALSE
)
```

```{r trans5, echo=TRUE}
askMC(
  "What's the *second* negative term in the expansion of $\\sin(x) + \\cos(x)$?",
  "$x^0$ term",
  "$x^1$ term",
  "$x^2$ term",
  "+$x^3$ term+",
  "$x^4$ term",
  "$x^5$ term",
  "there are none",
  random_answer_order = FALSE
)
```

```{r trans6, echo=TRUE}
askMC(
  "What's the first zero term in the expansion of $\\sin(x) + \\cos(x)$?",
  "$x^0$ term",
  "$x^1$ term",
  "$x^2$ term",
  "$x^3$ term",
  "$x^4$ term",
  "$x^5$ term",
  "+there are none+",
  random_answer_order = FALSE
)
```

