---
id: "girl-wake-bottle"
created: "Mon Jul 19 18:03:49 2021"
global_id: "KmDiXI"
---


Recall the formula for the coefficients $a_n$ to approximate a function $f(x)$:

$$a_n = \frac{f^{(n)}(x_0)}{n!}$$
where $f^{(n)}$ means the "n^th^ derivative. For simplicity, we'll look at examples where $x_0 = 0$, so that the polynomial itself will be 

$$p(x) = a_0 x^0 + a_1 x^1 + a_2 x^2 + a_3 x^3 + \cdots$$
Of course, $x^0$ equals 1.

Let's construct the polynomial coefficients for the function $f(x) = e^x$ around $x_0 = 1$. Start with the sequence of derivatives $f^{(n)}(0)$.

```{r bc1-1, echo=FALSE, results="markup"}
askMC(
  "Which of these gives the sequence $f^{(n)}(0)$ (where $f(x) = e^x$) for $n=0, 1, 2, 3, ...$ ?",
  "0, 0, 0, 0, 0, ...",
  "+1, 1, 1, 1, 1, ...+",
  "1, 2, 3, 4, 5, ...", 
  "0, 1, 2, 3, 4, ...",
  random_answer_order = FALSE
)
```

```{r bc1-2, echo=FALSE, results="markup"}
askMC(
  "Which of these gives the sequence $n!$ for $n=0, 1, 2, 3, ...$ ?",
  "0, 0, 0, 0, 0, ...",
  "1, 2, 3, 4, 5, ...", 
  "0, 1, 2, 3, 4, ...",
  "+1, 1, 2, 6, 24, ...+",
  random_answer_order = FALSE
)
```


```{r bc1-3, echo=FALSE, results="markup"}
askMC(
  "One of these is **not** the polynomial expansion for $e^x$. Which one?",
  "$1 + x + \\frac{x^2}{2} + \\frac{x^3}{6} + \\frac{x^4}{24} + \\cdots$",
  "+$1 + x + \\frac{x^2}{2} + \\frac{x^3}{3} + \\frac{x^4}{4} + \\cdots$+",
  "$1 + x + \\frac{x^2}{2!} + \\frac{x^3}{3!} + \\frac{x^4}{4!} + \\cdots$"
)
```

Now let's construct the polynomial expansion of $\sin(x)$ using the same technique: Find the sequence of derivatives $f^{(n)}$, then divide each of them by $n!$ to construct the $a_n$ coefficients.

```{r bc1-4, echo=FALSE, results="markup"}
askMC(
  "Which of these gives the sequence $f^{(n)}(0)$ (where $f(x) = \\sin(x)$) for $n=0, 1, 2, 3, ...$ ? (Remember, $x_0$ will be set to zero.)",
  "0, 1, -1, 1, -1, ...",
  "1, 0, -1, 0, 1, ...",
  "+0, 1, 0, -1, 0, ...+", 
  "1, 0, 0, -1, 1, ...",
  random_answer_order = FALSE
)
```

```{r bc1-5, echo=FALSE, results="markup"}
askMC(
  "One of these **is** the polynomial expansion for $\\sin(x)$. Which one?",
  "$1 + x + \\frac{x^2}{2} + \\frac{x^3}{6} + \\frac{x^4}{24} + \\cdots$",
  "$x - \\frac{x^3}{3} - \\frac{x^5}{5} + \\cdots$",
  "+$x - \\frac{x^3}{6} - \\frac{x^5}{120} + \\cdots$+",
  "$1 + x - \\frac{x^3}{6} - \\frac{x^5}{24} + \\cdots$"
)
```

```{r bc1-6, echo=FALSE, results="markup"}
askMC(
  "Take the correct polynomial expansion for $\\sin(x)$ from the previous question and differentiate it term by term to get the polynomial expansion for $\\cos(x)$. Which one of these is right",
  "$1 - \\frac{x^2}{2} + \\frac{x^3}{6} - \\frac{x^4}{24} + \\cdots$",
  "$1 - \\frac{x^2}{2} - \\frac{x^4}{4} + \\cdots$",
  "+$1 - \\frac{x^2}{6} + \\frac{x^4}{24} - \\cdots$+",
  "$1 - \\frac{x^2}{2} + \\frac{x^4}{120} + \\cdots$"
)
```


Finally, let's construct the polynomial expansion of the function $f(x) \equiv \frac{1}{1-x}$ at $x_0 = 0$.

```{r bc1-7, echo=FALSE, results="markup"}
askMC(
  "Which of these gives the sequence $f^{(n)}(0)$ (where $f(x) = 1/(1-x)$) for $n=0, 1, 2, 3, ...$ ? (Remember, $x_0$ will be set to zero.)",
  "+1, 1, 1, 1, 1 ...+",
  "1, -1, 1, -1, 1, ...",
  "0, 1, 0, 1, 0, ...", 
  "1, 0, 1, 0, 1, ..."
)
```

```{r bc1-8, echo=FALSE, results="markup"}
askMC(
  "What are the coefficients $a_n$ in the polynomial expansion of $1/(1-x)$, for $n = 0, 1, 2, 3, 4, ...$",
  "1, 1, 1, 1, 1, ...",
  "1, 1/2, 1/6, 1/24, 1/120, ...",
  "+1, 1, 1/2, 1/6, 1/24, ...+",
  "1, 1, 2!, 3!, 4!"
)
```

