---
id: "lamb-put-bulb"
created: "Wed Jul 28 14:52:46 2021"
global_id: "MW03ay"
---

Sometimes a bit of algebra can help us see what's going on with the instantaneous rate of change. Consider the exponential function $e^x$.

Rather than writing the slope function definition with a 0.1, let's substitute in the symbol $h$. This gives $${\cal D}\exp(x) = \frac{e^{x+h} - e^x}{h}$$
Extracting out the common term $e^x$ in the numerator, we arrive at $${\cal D}\exp(x) = e^x \left[\frac{e^h - 1}{h}\right]$$
Since $h$ is a number, $[e^h - 1]/h$ is a number, not a function of $x$. So, for any given value of $h$, the slope function of the exponential is proportional to the exponential itself.

```{r lpb4-1, echo=FALSE, results="markup"}
askMC(
  "Using a sandbox, find the value of $[e^h - 1]/h$ when $h=0.1$. Which of these is it?",
  "1.271828",
  "1.694343",
  "+1.718282+",
  "e = 2.718282",
  random_answer_order=FALSE
  )
```

The instantaneous rate of change involves making $h$ very small, but not quite zero. If you make $h$ exactly zero, the result will be ambiguous.

```{r lpb4-2, echo=FALSE, results="markup"}
askMC(
  "Using a sandbox, compute $[e^0 - 1]/0$. What's the result?",
  "`Inf`" = "This would be the result if the numerator were positive, however small. But both the numerator and the denominator are zero.",
  "+`NaN`+" = "Meaning, 'not a number'.",
  "`Bogus`" = "This is **not** a value and can't be used in arithmetic. But `Inf` and `NaN` each can be legitimately used in arithmetic.",
  "An error message result." = "The point of `Inf` and `NaN` is to avoid creating error conditions while still signalling that the result is ambiguous.",
  random_answer_order = FALSE
)
```

```{r lpb4-3, echo=FALSE, results="markup"}
askMC(
  "Using a sandbox, compute $[e^h - 1]/h$. Make $h$ as small as you can, for instance $0.00001$ or $0.00000001$. What's the result?",
  "+1+" = "Amazingly, the instantaneous rate of change of $e^x$ is exactly $e^x$.",
  "It varies with $h$, but is always around 1.5 for $h$ small enough",
  "There is no particular result.",
  random_answer_order = FALSE
)
```

