---
id: "optim-blue"
author: "Daniel Kaplan"
---

```{r include = FALSE}

```


In the previous problem, you used graphics to find the argmin of a function. You can, of course, successively reset the domain of the graphing window to zoom in on the location of the argmin.

To find the argmin numerically, you start with a (pretty good) guess for the argmin. Recall that a function (such as `total_time()`) can be modeled with a low-order polynomial. When the function has an extreme point the polynomial should include a quadratic term. Suppose you have guessed that $x=10$ is a good starting guess for the argmin of the air-to-water problem. You can compute the first and second derivatives of `total_time(x)` with respect to `x` and evaluate those derivative functions at $x=10$. (Use the `D()` function to construct the derivatives.) In the sandbox, use the value of those two derivatives at the $x=10$ guess to find the approximation to the minimum of the function.

To help remind you, the quadratic approximation to a function $t(x)$ around a reference point $x_0$ is: $$q(x) \equiv t(x_0) + \partial_x t(x_0) [x - x_0] + \frac{1}{2} \partial_{xx} t(x_0) [x - x_0]^2$$
Find the derivative $\partial_x q(x)$ and set this equal to zero. Inserting appropriate values for $x_0$, $\partial_x t(x_0)$, and $\partial_{xx} t(x_0)$ will enable you to calculate the value of $x$ at the minimum of the approximating quadratic.

```{r daily-digital-36-QA4, echo=FALSE}
askMC(
  prompt = "iii) Letting the blue area be water (1/1.3 speed factor) and starting with a guess $x_0=10$, find the $x$ that minimizes the approximating quadratic.",
  9.912,10.748,11.017,11.151,11.267,11.327,11.515,11.713,11.810,12.074,12.276,12.321,12.438,"12.461","+12.579+", 12.618,12.637,12.993,13.261,13.360,13.527,13.888,
  random_answer_order = FALSE
  )
```

```{r daily-digital-36-QA5, echo=FALSE}
askMC(
  prompt = "iv) Using the correct answer from (iii) for a new $x_0$, find the value of $x$ that minimizes the new approximating quadratic.",
  9.912,10.748,11.017,11.151,11.267,11.327,11.515,11.713,11.810,12.074,12.276,12.321,12.438,"+12.461+", 12.579, 12.618,12.637,12.993,13.261,13.360,13.527,13.888,
  random_answer_order = FALSE
  )
```

## Tree harvesting

Your uncle Bob is writing a business plan for a tree farm for lumber. Having heard that you are taking Math 141Z, he emails you giving some information asking for some numbers. In particular, Bob sends you a report saying that, for the species of tree he plans to plant, the amount of usable lumber is a function of growth time $t$ looking like the function defined in the sandbox: 

::: {.scaffolding  data-latex=""}
```{r}
lumber <- makeFun(ifelse(t<10, 0, 100*exp((t-25)/8)/(1 + exp((t-25)/8))) ~ t)
dt_lumber <- D(lumber(t) ~ t)
ave_growth_rate <- makeFun( your_expression_here ~ t)
slice_plot(lumber(t) ~ t, bounds(t=0:100))
```
:::

Bob has heard that the time to harvest is when the tree is growing fastest.

```{r daily-digital-36-QA6, echo=FALSE}
askMC(
  prompt = "What is the value of `t` (in years) at which $\\partial_t$ `lumber(t)` is largest?",
  15, 16, 17, 18, 19, 20, 21, 22, 23, 24, "+25+", 26, 27, 28, 29, 30,
  random_answer_order = FALSE
)
```

You patiently explain to your uncle that you certainly do not want to harvest trees when they are growing the fastest. You say, "You want to wait until the average growth rate up to that point is fastest. That will be a little while before the tree reaches its adult volume."



```{r daily-digital-36-QA37, echo=FALSE}
askMC(
  prompt = "What is the value of t (in years) for which the average growth rate, up to that time, is fastest.",
  15, 17, 19, 21, 23, 25, 27, 29, 31, 33, "+35+", 37, 39, 41,
  random_answer_order = FALSE
)
```

