---
id: "optim-red"
author: "Daniel Kaplan"
---

```{r include = FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
```

::: {.todo}
Turn this into a problem about algebra.

Later in accumulation, figure out the mile-markers on the road.
:::


There is a tradition in mathematics education of using *geometrical*, distance-related problems to illustrate optimization. As it happens, some such problems can provide some valuable insight into physical situations.

```{r echo=FALSE}
make_dist_fun <- function(..., method=c("euclidean", "sphere", "minkowski")) {
  method <- match.arg(method)
  origin <- unlist(list(...))
  if (method == "euclidean") {
    function(...) {
      target <- list(...)
      if (length(target) != length(origin))
        stop("Origin point and target must have the same number of coordinates.")
      accumulator <- 0
      for (k in 1:length(target)) {
        accumulator <- accumulator + (target[[k]] - origin[k])^2
      }
      sqrt(accumulator)
    }
  } else if (method == "sphere") {
    if (length(origin) != 2)
      stop("Supply two arguments in degrees, lat and long, for spherical distances")
    origin <- pi*origin/180
    origin_lat <- origin[1]
    origin_long <- origin[2]
    function(lat, long, radius = 1) {
      lat <- pi*lat/180
      long <- pi*long/180
      del_lambda <- long - origin_long
      del_phi <- lat - origin_lat
      # Haversign formula
      radius* 2 * asin(sqrt(sin(del_phi/2)^2 +
                      cos(origin_lat)*cos(lat)*sin(del_lambda/2)^2))
    }
  }
}
```

Suppose you have a function $f(x)$ whose graph represents the path of a road through the jungle. There is a lion at coordinates $(3, 2)$ in the jungle. At what point as you travel along the road will you be closest to the lion.

The trick here is to see that your position on the road at any value of $x$ is $(x, f(x))$. The distance to the lion as a function of $x$ is $$ \text{lion.dist}(x) \equiv \sqrt{(x - 3)^2 + (f(x)-2)^2} .$$

The road and the lion are graphed in the sandbox for the road $f(x) \equiv x^2 e^{-x}$. You can see that the road comes closest to the lion at roughly $x=3$. When you implement the lion.dist$(x)$ function, you'll be able to find the *argmin*, that is, the value of the input $x$ for which the output lion_dist$(x)$ is minimal. The value of the function lion_dist$(x)$ at the argmin is the *min* of the function.

```{r}
slice_plot(x^2 * exp(-x) ~ x, domain(x=0:5)) %>%
  gf_label(2 ~ 3, label="lion") %>%
  gf_refine(coord_fixed())

```

To find the argmin of the distance to the lion more precisely, implement the $\text{lion.dist}(x)$ function in the sandbox. Plot this out and look for the minimum. Alternatively, find $\partial_x \text{lion.dist}(x)$ and find the $x$ of the zero crossing.



```{r daily-digital-36-QA38, echo=FALSE}
askMC(
  prompt = "Using the sandbox above, find the argmin for which the graph of the function $g(x) \\equiv x^2 \\exp(-x)$ comes closest to the point $(3, 2)$? (Choose the best answer.) ",
  2.03, 2.07, 2.21, 2.27, 2.28, 2.33, 2.37, 2.39, 2.43, 2.48, 2.58, 2.71, "+2.80+", 2.94,
  random_answer_order = FALSE
)
```


