---
id: "optim-green"
author: "Daniel Kaplan"
---

```{r include = FALSE}

library(mosaic)
library(mosaicCalc)
```


Your uncle Bob is writing a business plan for a tree farm for lumber. Having heard that you are taking Math 141Z, he emails you giving some information asking for some numbers. In particular, Bob sends you a report saying that, for the species of tree he plans to plant, the amount of usable lumber is a function of growth time $t$ looking like the function `lumber()`: 

::: {.scaffolding  data-latex=""}
```{r}
lumber <- makeFun(ifelse(t<10, 0, 100*exp((t-25)/8)/(1 + exp((t-25)/8))) ~ t)
dt_lumber <- D(lumber(t) ~ t)
ave_growth_rate <- makeFun( your_expression_here ~ t)
slice_plot(lumber(t) ~ t, bounds(t=c(0,100)))
```
:::

Bob has heard that the time to harvest is when the tree is growing fastest.

```{r daily-digital-36-QA6, echo=FALSE}
askMC(
  prompt = "A) What is the value of `t` (in years) at which $\\partial_t$ `lumber(t)` is largest?",
  15, 16, 17, 18, 19, 20, 21, 22, 23, 24, "+25+", 26, 27, 28, 29, 30,
  random_answer_order = FALSE
)
```

You patiently explain to your uncle that you certainly do not want to harvest trees when they are growing the fastest. You say, "You want to wait until the average growth rate up to that point is fastest. That will be a little while before the tree reaches its adult volume."



```{r daily-digital-36-QA37, echo=FALSE}
askMC(
  prompt = "B) What is the value of t (in years) for which the average growth rate, up to that time, is fastest.",
  15, 17, 19, 21, 23, 25, 27, 29, 31, 33, "+35+", 37, 39, 41,
  random_answer_order = FALSE
)
```
