---
id: "partial-purple"
origin: "141 DD29"
global-id: "9iowVB"
---

```{r include=FALSE}
library(Zcalc)
```

Open a sandbox and use the following commands to make a contour plot of the function $g(x)$ centered  on  the reference point $(x_0\!=\!0,\, y_0\!=\!0)$. 

```{r results="hide"}
g <- rfun( ~ x + y, seed = 802, n = 15)
x0 <-  0
y0 <-  0
size <- 5
contour_plot(g(x, y) ~ x + y,
             domain(x = x0 + size*c(-1, 1),
                    y = y0 + size*c(-1, 1))) 
```


By making `size` smaller, you can zoom in around the reference point. Zoom in gradually (say, `size = ` 1.0,  0.5, 0.1, 0.05, 0.01) until you reach a  point where the surface plot is (practically) a pretty simple inclined  plane.


From the contour plot, zoomed in so that the graph shows an inclined plane, figure out the sign of $\partial_x g(0,0)$ and $\partial_y g(0,0)$.

```{r daily-digital-29-QA9, echo=FALSE}
askMC(
  prompt = "Which answer best describes the signs of the partial derivatives of $g(x,y)$ at the reference point $(x_0=0, y_0=0)$? ",
    "$\\partial_x g(0,0)$ is pos, $\\partial_y g(0,0)$ is pos ",
    "$\\partial_x g(0,0)$ is pos, $\\partial_y g(0,0)$ is neg ",
    "$\\partial_x g(0,0)$ is neg, $\\partial_y g(0,0)$ is neg ",
    "+$\\partial_x g(0,0)$ is neg, $\\partial_y g(0,0)$ is pos.+",
    "$\\partial_x g(0,0)$ is 0, $\\partial_y g(0,0)$ is pos", 
  inline = FALSE
)
```

