---
id: "partial-tan"
origin: "141 DD29"
author: "Daniel Kaplan"
---

```{r include=FALSE}

library(mosaic)
library(mosaicCalc)
```


Consider this close up of a function around a reference point at the center of the graph.

```{r echo=FALSE}
g <- rfun( ~ x + y, seed = 802, n = 15)
x0 <- -2.
y0 <-  -5
size <- 0.5
gg <- makeFun(-g(x,y) ~ x + y)
contour_plot(gg(x, y) ~ x + y, skip = 0, contours_at=seq(0,3,by=0.1),
             domain(x = x0 + size*c(-1, 1),
                    y = y0 + size*c(-1, 1)))  +
  scale_y_continuous(breaks=seq(-6,-4,by=0.1)) +
  scale_x_continuous(breaks=seq(-3, -1, by=0.1)) +
  coord_fixed()
```

By eye, estimate these derivatives of the function at the reference point $(x_0=-2, y_0=-5)$.




```{r daily-digital-29-QA10, echo=FALSE}
askMC(
  prompt = "What is the numerical value  of $\\partial_x g(x,y)$ at the reference point?",
  -1, "-0.50", -0.25, 0, 0.25, "+0.50+", 1,
  random_answer_order = FALSE
)
```

```{r daily-digital-29-QA11, echo=FALSE}
askMC(
  prompt = "What is the numerical value  of $\\partial_y g(x,y)$ at the reference point?",
  -1, "-0.50", -0.25, 0, 0.25, "0.50", "+1+",
  random_answer_order = FALSE
)
```

The next questions ask about second-order  partial derivatives. As you know, the second derivative is about how the first derivative changes with x or y. Insofar as the function is a simple inclined plane, where the contours would be straight, parallel,  and evenly spaced,  the second derivatives would all  be zero. But you can see that it is not such a plane: the contours curve a bit.

In  determining the second derivatives by eye from the graph, you  are encouraged to compare first derivatives at the opposing edges of the graph,  as opposed to at very nearby points.


```{r daily-digital-29-QA12, echo=FALSE}
askMC(
  prompt = "What is the sign  of $\\partial_{xx} g(x,y)$ at the reference point?",
  "+negative+", "positive",
  random_answer_order = FALSE
)
```

```{r daily-digital-29-QA13, echo=FALSE}
askMC(
  prompt = "What is the sign  of $\\partial_{yy} g(x,y)$ at the reference point?",
  "negative", "+positive+",
  random_answer_order = FALSE
)
```

```{r daily-digital-29-QA14, echo=FALSE}
askMC(
  prompt = "What is the sign  of $\\partial_{xy} g(x,y)$ at the reference point?",
  "negative", "+positive+",
  random_answer_order = FALSE
)
```

```{r daily-digital-29-QA15, echo=FALSE}
askMC(
  prompt = "What is the sign  of $\\partial_{yx} g(x,y)$ at the reference point?",
  "negative", "+positive+",
  random_answer_order = FALSE
)
```
