---
id: "rooster-red"
author: "Daniel Kaplan"
---

```{r include = FALSE}

library(mosaic)
library(mosaicCalc)
```

It often happens in decision making that there are multiple criteria. For instance, in selecting cadets for pilot training, two obvious criteria are the cadet's demonstrated flying aptitude and the leadership potential of the cadet. Let's assume that the merit $M$ of a candidate is a function of flying aptitude $F$ and leadership potential $L$. 

Currently, the merit score is a simple function of the $F$ and $L$ scores: $$M_{current}(F, L) \equiv F + L$$

The general in charge of the training program is not satisfied with the current merit function. "I'm getting too many cadets who are great leaders but poor pilots, and too many pilot hot-shots who are poor leaders.  I would rather have an good pilot who is a good leader than have a great pilot who is a poor leader or a poor pilot who is a great leader." (You might reasonably agree or disagree with this point of view, but the general is in charge.)

You've been tasked to develop an improved formula for the merit score $M_{improved}(F, L)$. 

A low-order polynomial model, without quadratic terms, is $$M(F, L) = d_0 + d_1 F + d_2 L + d_{12} F L .$$ 

You are trying to decide whether $d_{12}$ should be positive, negative, or zero. To guide your decision, you have only this statement from the general in charge of the program: "

Take a few minutes now to think about how you would decide whether $d_{12}$ should be positive or negative or zero to implement the general's view. The rest of the problem will guide you to a solution, but you'll understand better if you try on your own, first.

Using the low-order polynomial model, find algebraically these two partial derivatives

- $\partial_F M(F, L)$ and
- $\partial_L M(F, L)$

```{r daily-digital-33-QA7, echo=FALSE}
askMC(
  prompt = "Which of these possibilities is true about $\\partial_F M$ as a function of $F$?",
  "Increases with F", "Decreases with F", "+Is not a function of F.+",
  inline = FALSE,
  random_answer_order = FALSE
)
```

```{r daily-digital-33-QA8, echo=FALSE}
askMC(
  prompt = "Which of these possibilities is true about $\\partial_L M$ as a function of $L$?",
  "Increasing", "Decreasing", "+Is not a function of L.+",
  inline = FALSE,
  random_answer_order = FALSE
)
```

```{r daily-digital-33-QA9, echo=FALSE}
askMC(
  prompt = "Which one of these describes the relationship between $\\partial_F M()$ as a function of $L$ and $\\partial_L M()$ as a function of $F$? (Hint: Remember that $d_1$ and $d_2$ are specified as being positive.)",
    "If $\\partial_F M()$ is an increasing function of $L$, then $\\partial_L M()$ must be a decreasing function of $F$." = "Notice that both partials involve the same $d_{12}$ coefficient.",
    "+If $\\partial_F M()$ is an increasing function of $L$, then $\\partial_L M()$ must be an increasing function of $F$.+" = "",
    "Neither of the above needs to be true." = "Notice that both partials involve the same $d_{12}$ coefficient.",
  inline=FALSE,
  random_answer_order = FALSE
)
```

Think now about the general's statement and how to translate it into mathematical terms. Here's a hint: Imagine Drew has very high $F$ score and low $L$ score. Consider another cadet, Blake, with scores $F-\epsilon$ and $L + \epsilon$. Blake's flying and leadership scores  are closer together than Drew's. Both Drew and Blake have exactly the same average of $F$ and $L$. But would the general find Blake superior or inferior to Drew? If superior, then you would want $\partial_L M > \partial_F M$. If inferior, then your model should have $\partial_L M < \partial_F M$. If the general would be indifferent between Drew and Blake, then the two partial derivatives should be equal.


```{r daily-digital-33-QA10, echo=FALSE}
askMC(
  prompt = "Based on the general's statement, do you want $d_{12}$ to be positive, negative, or zero?",
    "+Positive+",
    "Zero",
    "Negative",
    "It has nothing to do with the general's statement.",
  inline=FALSE,
  random_answer_order = FALSE
)
```
