---
id: "singularity-numerics"
origin: "141 DD30"
author: "Daniel Kaplan"
---

```{r include = FALSE}
library(etude2)
library(mosaic)
library(mosaicCalc)
```


Consider the function $f(x)  \equiv  \frac{x-3}{x-3}$. Simple algebra suggests that  the  output will be 1 for any $x$ value input, but the rules are division mean that the output is undefined for input $x=3$. We say,  the  function has a "singularity" at $x=3$. We want to test whether this singularity is removable, that is, whether  the function has  a  limit as $x \rightarrow 3$. The computer can help us out.  The basic idea is to try inputs closer  and closer to $x=3$  and see if the output behaves nicely. To help you distinguish between removable and  non-removable singularities (or, in the words of the song "Santa Claus is Coming to Town", to know "who's naughty or nice"), the sandbox defines the function $f()$ and  a similar function $g()$. Try out both $f()$ and $g()$ for a sequence of inputs that get closer and closer to $x=3$ but are not *exactly*  3.

```{r daily-digital-30-sandbox2, exercise=TRUE, exercise.cap="Naughty or nice", exercise.eval=FALSE, exercise.lines=10, exercise.completion=FALSE, exercise.startover=TRUE, exercise.diagnostics=TRUE}
f <- function(x) { (x-3) / (x-3)}
g <- function(x) { (x-2.5) / (x-3)}
x_sequence <- 3 - 10^-(1:15)
options(digits=20)
x_sequence
f(x_sequence)
g(x_sequence)
```

**Essay 1**: Explain what the `options(digits=20)` statement is doing. You  can do some experiments by changing the 20 to something much smaller.

```{r daily-digital-30-QA3, echo=FALSE}
askMC(
  prompt = "Which of these statements is true about the successive values in  `x_sequence`?",
  "+It approaches 3 from below.+",
              "It approaches 3 from above.",
              "It oscillates, getting nearer and farther from 3.",
  random_answer_order = FALSE)

```

```{r daily-digital-30-QA3b, echo=FALSE}
askMC(
  prompt = "Which of these statements is true about the successive values in  `f(x_sequence)`?",
  "It approaches 1 from below.",
              "It approaches 1 from above.",
              "+Its value is constant at 1.+",
  random_answer_order = FALSE)

```

**Essay 2**: Explain what about the output of the function $g()$ qualifies it for being numerically 'naughty' near $x=3$.

