---
chapter: "none yet"
author: "Danny Kaplan"
difficulty: "XX"
date: 2021-06-08 
version: 0.1
tags: [first, second, third]
id: snail-tear-blanket
global-id: "KEWIX"
---

```{r include = FALSE}

library(mosaic)
library(mosaicCalc)
```


We will be working extensively with  the *change* in output value  of a function when the *input value changes*. 

- The change in the output value of a function $f()$ when the input changes from $x = a$ to $x = b$ is $$f(x=b) - f(x=a)$$ Notice that when we talk about the change from $x=a$ to $x=b$ we subtract $f(a)$ from  $f(b)$. That  change is sometimes called the **rise** in the value of the function.  Rise always compares (by subtraction) the two **output values** corresponding to two specific input values. Remember that $a$ and $b$ stand for  specific numbers. 

- Corresponding with  the idea of the  change in output being $f(b) - f(a)$ the  change in the input value to a function is $b - a$. This is often  called the  *run* in the value of the input.


```{r echo=FALSE, warning=FALSE}
Pts2 <- tribble(
  ~ x, ~ y, ~label,
  1, 4, "(a,f(a))",
  3.5, 3, "(b,f(b))"
)
gf_point(y ~ x, data = Pts2) %>%
  gf_text(label = ~ label, vjust = 1, nudge_y = .15) %>%
  gf_lm() %>%
  gf_lims(x = c(0,5)) %>%
  gf_labs(title = "Graph I.", y="f(x)")
```


```{r daily-digital-24-TF1, echo=FALSE}
askMC(
  prompt = "True or false: In Graph I, the *rise* from a to b is positive.",
  "TRUE" = "$f(x=a) > f(x=b)$, so the rise $f(x=b) - f(x=a)$ is *negative*.",
  "+FALSE+",
  random_answer_order = FALSE
  )
```

```{r daily-digital-24-TF2, echo=FALSE}
askMC(
  prompt = "True or false: In Graph I, the *run* from $x=a$ to $x=b$ is positive.",
  "+TRUE+",
  "FALSE" = "The *run* is about the relative positions of $x=a$ and $x=b$ on the x-axis. Since $a < b$, the run from $x=a$ to $x=b$ is positive.",
  random_answer_order = FALSE)
```

```{r echo=FALSE}
f <- makeFun(sin(x) ~ x)
Pts2 = tribble(
  ~x, ~label,
  .1, "a",
  2.5, "c",
  3.5, "b"
)
slice_plot(f(x) ~ x, domain(x=c(0, 5))) %>%
  gf_point(f(x) ~ x, data = Pts2) %>%
  gf_text(f(x) ~ x, label = ~label, data = Pts2, nudge_y =-.1) %>%
  gf_labs(title = "Graph II")
```

```{r daily-digital-25-TF3, echo=FALSE}
askMC(
  prompt = "True or false: In Graph II, the *run* from a to b is positive.",
  "+TRUE+"  = "                                   ",
  "FALSE" =  "The *run* is about the relative positions of a and b on the x-axis. Since a is to the left of b, the run from a to b is positive.",
  random_answer_order = FALSE)
```

```{r daily-digital-26-TF4, echo=FALSE}
askMC(
  prompt = "True or false: In Graph II, the *rise* from a to b is positive.",
  "TRUE" = "Remember,  the *rise* from $x=a$ to $x=b$ is $f(x=b) - f(x=a)$",
  "+FALSE+" = "                                   ",
  random_answer_order = FALSE)
```

```{r daily-digital-24-TF5, echo=FALSE}
askMC(
  prompt = "True or false: In Graph II, the *run* from b to c is positive.",
  "TRUE" = "The *run* from $x=b$  to $x=c$ is $c - b$. Since $b>c$ b to c is negative.",
  "+FALSE+",
  random_answer_order = FALSE)
```

```{r daily-digital-24-TF6, echo=FALSE}
askMC(
  prompt = "True or false: In Graph II, the *rise* from b to c is positive.",
  "+TRUE+",
  "FALSE" = "The *rise* from $x=b$ to $x=c$ is $f(x = c) - f(x = b)$. Since $f(x=c)   > f(x=b)$, the rise is positive.",
  random_answer_order = FALSE)
```


```{r daily-digital-24-QA5, echo=FALSE}
askMC(
  prompt = "For an interval [2, 6] what is the value of the run? (The answer is independent of any particular graph/function.)",
    "+4+" = "The run is always the second number in the interval minus the first number. that is  $6 - 2$ here.",
    "-4" = "You got it backwards! The second number in the interval, 6, is numerically to the right of 2, so the run is positive."
)
```

```{r daily-digital-24-QA6, echo=FALSE}
askMC(
  prompt = "Which is the run of the interval [6, 2]? (Again, the answer is independent of any particular graph/function.)",
    "4" = "Sorry. The run from $x=6$  to $x=2$  is $2 - 6$ which is $-4$.",
    "+-4+" = "The run is $2 - 6$, the second number in the interval minus the first number."
)
```




