---
author: "Danny Kaplan"
origin: "141 DD27"
date: 2021-06-09 
id: spider-blow-lamp
global-id: "72ldw"
---

```{r spider-blow-lamp-default, include = FALSE}
```

In economic theory, the quantity of the demand for any good is a decreasing function of the price of that good and an *increasing* function of the price of a competing good.

The classical example is that apple juice competes with orange juice. The demand for orange juice is  in  units of thousands of liters of orange juice. The price is in units of dollars  per  liter.

Here's a graph with the input quantities unlabeled. The contour labels indicate the demand for orange juice.



```{r echo=FALSE}
f <- function(x,y) {
  angle = atan2(y, x)
  300* ((pi/2) - angle)
}
contour_plot(f(x, y) ~ x + y, domain(x=c(3,1.3), y=c(.4,1.3)),
             skip = 0) %>%
  gf_labs(title = "Demand quantity", x = "?", y = "?")
```

The concept of  partial derivatives makes it much easier to think about the situation. There are two partial derivative functions relevant  to the function in the graph. Well denote the inputs `apple` and `orange`, but remember that  these are the prices of those commodities in  dollars per liter.

* $\partial_\text{apple} \text{demand}()$ -- how the *demand* changes when apple-juice price goes up, *holding orange-juice  price constant*. (Another notation that is more verbose but perhaps easier to read $\frac{\partial\, \text{demand}}{\partial\,\text{apple}}$)
* $\partial_\text{orange} \text{demand}()$ -- how the *demand* changes when orange-juice price goes up, *holding apple-juice  price constant*. (Another notation:  $\frac{\partial\, \text{demand}}{\partial\,\text{orange}}$)

Notice that the notation  names both the output and the single input which is to be changed--the other inputs will be held constant.

The first  paragraph of this problem gives the economic theory which amounts to saying that one of the partial derivatives is  positive and the other negative.

```{r daily-digital-27-QA2, echo=FALSE}
askMC(
  prompt = "What is the proper translation of the notation $\\partial_\\text{apple}\\text{demand}()$?",
    "+The partial derivative of orange-juice  demand *with repect to* apple-juice price+" = "",
    "The partial derivative of apple-juice price *with respect to* demand for orange juice" = "Derivatives are always with respect to an *input*. Demand for orange juice is the output.",
    "The partial derivative of apple-juice demand *with respect to* price of apple juice" = "This problem is about orange-juice demand, not apple-juice demand.",
    "The partial derivative of orange-juice price *with respect to* apple-juice price." = "Both of these are *inputs* to the demand function. A partial derivative is always about the change in output when one of the inputs changes, not the change in one input when another input changes."
)
```


```{r daily-digital-27-QA3, echo=FALSE}
askMC(
  prompt = "According to the economic theory described above, one of the partial derivatives will be positive and the other negative. Which will be *positive*.",
    "+$\\partial_\\text{apple}  \\text{demand}()$+" = "",
    "$\\partial_\\text{orange} \\text{demand}()$" = "If orange juice prices went up, would you drink more orange juice?"
)
```


```{r daily-digital-27-kayak-QA4, echo=FALSE}
askMC(
  prompt = "What does the vertical axis measure?",
    "+Price of orange juice+" = "",
    "Quantity of apple juice" = "",
    "Quantity of orange juice" = "",
    "Price of apple juice" = ""
)
```

```{r daily-digital-27-QA5, echo=FALSE}
askMC(
  prompt = "Consider the magnitude (absolute value) of the partial derivative of demand with respect to orange-juice price. Is this magnitude greater toward the top of the graph or the bottom?",
    "+top+" = "",
    "bottom" = "The horizontal distance between contour lines is larger at the bottom of the graph than at the top. Far-apart contours mean that the function is flatter.",
    "neither" = "Remember, we're talking about the partial derivative with respect to orange juice price. That corresponds to the slope when moving in a *horizontal* direction."
)
```





