---
id: "tree-harvesting"
author: "Daniel Kaplan"
---

```{r include=FALSE}

library(mosaic)
library(mosaicCalc)
```


The graph shows a simplified model of the amount of usable wood that can be harvested from a typical tree in a managed forest of Ponderosa Pine. (You can see some actual forestry research models [here](https://www.fs.fed.us/rm/pubs/rmrs_gtr292/1992_milner.pdf).)

```{r aver-tree, echo=FALSE}
tree_volume <- makeFun(1000*pnorm(year, mean=15, sd=10)^3 ~ year)
# do the deriv numerically to avoid bug in derivative
# of normal function.
tree_growth <- numD(tree_volume(year) ~ year)
slice_plot(tree_volume(year) ~ year, domain(year = c(0, 50))) %>%
  gf_labs(y = "Usable wood product (board-feet)",
          x = "Year after planting")
```

You are writing a business plan for a proposed pine forest. Among other things, you have to forecast the revenue that will be generated and when you will have salable product.

They say that "time is money." Every year you wait before harvest is another year that you don't have the money. On the other hand, every year that you wait means more wood at the end. How to decide when to harvest?

The tree continues to grow until year 50, when it seems to have reached an equilibrium: perhaps growth goes to zero, or rot balances what growth there is. There is no point waiting until after year 50.

At year 25, the tree is growing as fast as it ever will. You will get about 600 board-feet of lumber. Should you harvest at year 25? No! That the tree is growing so fast means that you will have a lot more wood at year 26, 27, and so on. The time to harvest is when the growth is getting smaller, so that it is not worth waiting an extra year.

The quantity of interest is the average rate of growth from seedling to harvest. Harvesting at year 25 will give a total change of 600 board feet over 25 years, giving an average rate of change of $600 \div 25 = 24 \text{board-feet-per-year}$. But if you wait until year 35, you will have about 900 board feet, giving an average rate of change of $900 \div 35 = 25.7 \text{board-feet-per-year}$.
:::

We've been presenting the average rate of change as a **number**: $$\frac{f(t_B) - f(t_A)}{t_B - t_A}$$

But it is helpful to think of it as a **function** that takes two inputs, which we can call $t_A$ and $t_B$:
$$\text{ave_rate_of_change}(t_A, t_B) \equiv \frac{f(t_B) - f(t_A)}{t_B - t_A}$$
This is a very subtle maneuver. The formula is the same, but by writing the quantities $t_A$ and $t_B$ as inputs (by putting them in parentheses on the left side), we turn the formula into a function.

Sometimes the average-rate-of-change function presents the information in a clear way.

::: {.practice data-latex=""}


Move this to the DERIVATIVES CHAPTER, OR INSTANTANEOUS RATE OF CHANGE.

Back to the forest ... Here's a graph of the average-rate-of change function. The function has two inputs, but botany dictates that we can only be growing wood from the time the seedling is planted. So we will set $t_A = 0$ and look at the corresponding *slice* of the function:

```{r echo=FALSE}
slice_plot(tree_volume(year)/year ~ year, domain(year = c(1, 50))) %>%
  gf_labs(y = "Annual growth (board-feet)",
          x = "Year after planting")
```
The graph makes it clear that the maximum average growth from planting to harvest will occur at about year 32.


**ANOTHER ELABORATION**

The context of the situation being modeled determines whether it is appropriate to look at an average rate of change or an instantaneous rate of change. @fig-instant-tree shows the instantaneous rate of change in the volume of wood.

```{r echo=FALSE, eval=!exists("tree_growth"), fig-cap: "A hump function makes a nice model of the rate of growth of a tree: slowly when young, increasing almost exponentially as it gets bigger, then slowing down as it ages."}
# This is a redefinition from Fun-slopes.Rmd for when we do partial compilation.
tree_volume <- makeFun(1000*pnorm(year, mean=15, sd=10)^3 ~ year)
# do the deriv numerically to avoid bug in derivative 
# of normal function.
tree_growth <- numD(tree_volume(year) ~ year)
```

```{r instant-tree, echo=FALSE}
slice_plot(tree_growth(year) ~ year, domain(year=c(0,50)))
```
It is tempting to look to the year where the growth rate is highest as the optimal harvest year. This is a mistake. The volume of wood being harvested is the ***accumulated growth*** not the instantaneous growth. Even though the instantaneous growth is higher at year 23 than year 30, it is still pretty high at year 30 and waiting until then (or later) accumates those years of higher-than-average growth. That is why the ***average rate of growth*** is a better thing to look at to determine optimal harvest time. Still, is it the right thing to look at? 

Between year 30 and 32, there is hardly any change in the value of the average-rate-of-change function. It is increasing a little, but is it really worthwhile to wait? One argument is that at year 29 you already have a valuable resource: wood that could be money in the bank. If the money were in the bank, you could invest it and earn more money *and* at the same time get a new seedling in the ground to start its growth. You're doing two things at once. Efficient!

To know what is the best year for harvest from this point of view, you want to calculate the effective "interest rate" on the present amount of wood that you earn in the form of new wood. That interest rate is the ratio of the *instantaneous* rate of growth of new wood divided by the amount of existing wood. @fig-tree-interest shows this function:

```{r tree-interest, echo=FALSE, warning=FALSE}
slice_plot(100* tree_growth(year)/tree_volume(year) ~ year,
           domain(year=c(0,50))) %>%
  gf_labs(y = "Growth relative to volume (%/year)") %>%
  gf_refine(scale_y_log10()) %>%
  gf_hline(yintercept=5, color="dodgerblue")
```
Early in the tree's life, the growth is high compared to the volume of the tree. That is because the tree is small. As the years pass, the tree gets bigger. Even though the rate of growth increases through year 23, the accumulated volume increases even faster, so there is a fall in the rate of return. 

The best time to harvest is when the annual "interest rate" paid by the growing tree falls to the level of the next best available investment. Suppose that investment would pay 10% per year. Then harvest the tree at year 24. If the next best investment paid only 5% (blue horizontal line), the harvest should be made at about year 29.



