---
id: "walnut-write-futon"
created: "Tue Aug 17 12:36:48 2021"
global_id: "ph92lT"
skill: "Sc-Opt-655"
---


Return to the problem of finding the optimal radius of a cylindrical tank with spherical ends. The point is to choose the sphere radius $r$ and the cylinder length $h$ that minimize the surface area of the tank while enclosing a specified volume. In terms of $h$ and $r$, the volume and surface area are:
$$V(r, h) = \pi r^2 h + \frac{4}{3} \pi r^3 \ \ \ \text{and}\ \ \ A(r,h) = 2 \pi r h + 4 \pi r^2$$

```{r tb1-1, echo=FALSE, results="markup"}
askMC(
  "Which of these is correct? (Hint: Only one of the answers is dimensionally consistent.)",
  "+$h(r, V) = (V-4\\pi r^3/3)/ \\pi r^2$+",
  "$h(r, V) = (V-4\\pi r^2/2)/ \\pi r^2$",
  "$h(r, V) = (V-4\\pi r^3/3)/ \\pi r^3$",
  "$h(r, V) = \\sqrt{(V-4\\pi r^3/2)/\\pi r^3}$"
)
```


```{r tb1-2, echo=FALSE, results="markup"}
askMC(
  "Which of these is the correct expression for $A(r, V)$",
  "$A(r, V) = 2 V/r + \\frac{8 \\pi}{3} r^2$",
  "+$A(r, V) = 2 V/r + \\frac{4 \\pi}{3} r^2$+",
  "$A(r, V) = V/r + \\frac{20 \\pi}{3} r^2$",
  "$A(r, V) = V/r + 3 \\pi r^2$"
)
```

```{r tb1-3, echo=FALSE, results="markup"}
askMC(
  "Find $\\partial_r A(r, V)$ and set to zero. Solve for $r^\\star$ in terms of $V$. Which of these is correct?",
  "+$r^\\star = \\sqrt[3]{\\frac{3}{4\\pi} V}$+",
  "$r^\\star = \\sqrt[3]{\\frac{4}{3\\pi} V}$",
  "$r^\\star = \\sqrt[3]{\\frac{3\\pi}{4} V}$",
  "$r^\\star = \\sqrt[3]{\\frac{3}{2\\pi} V}$"
)
```


Find the optimum value of $r$ to minimize $A(r,V)$ when $V = 1000$ liters. 

```{r tb1-4, echo=FALSE, results="markup"}
askMC(
  "What is the optimal value of $r$ in cm to a precision of one micron?",
  "6.2035" = "Check that you were using the right units for $V$.",
  "46.0351",
  "52.0351",
  "+62.0351+",
  random_answer_order = FALSE
)
```  

Use a sandbox to plot a graph of $A(r, V)$ versus r, setting $V = 1000$ liters. 



```{r rb2-2-solution, echo=FALSE, eval=FALSE}
A <- makeFun(2*(V - 4*pi*r^3/3)/r + 4*pi*r^2 ~ r, V = 1000000)
slice_plot(A(r) ~ r, bounds(r=55:70)) %>%
  gf_labs(y="Surface area of tank (square cm)")
```

```{r tb1-5, echo=FALSE, results="markup"}
askMC(
  "From the graph of $A(r, V)$ versus $r$ at $V=1000$ liters, read off a range of $r$ that produces $A$ no worse than 1% greater than the minimum. How wide is that range, approximately?",
  "$\\pm 0.1$cm",
  "$\\pm 1$cm",
  "+$\\pm 5$cm+",
  "$\\pm 10$cm",
  random_answer_order = FALSE
)
```

