---
id: "wolf-talk-kayak"
created: "Wed Aug 11 13:45:37 2021"
global_id: "n78LrF"
---

As you know, given a function $g(x)$ it's easy to construct a new function ${\cal D}_x g(x)$ that will be an approximation to the derivative $\partial_x g(x)$. The approximation function, which we call the slope function, can be $${\cal D}_x g(x) \equiv \frac{g(x + 0.1) - g(x)}{0.1}$$

Open a `r sandbox_link()` and use `makeFun()` to create a function $g(x) \equiv e^x$ and another function called `slope_of_g()` using the definition of ${\cal D} g(x)$.

```r
g <- makeFun(exp(x) ~ x)
slope_of_g <- makeFun( _your_tilde_expression_here_ )
```

```{r wtf1-1, echo=FALSE, results="markup"}
h <- makeFun(exp(x) ~ x)
slope_of_h <- makeFun( (h(x + 0.1) - h(x))/0.1 ~ x )
askMC(
  "What's the value of `slope_of_g(1)`?",
  "0.37",
  "0.85",
  "1.37",
  "1.85",
  "+2.85+",
  random_answer_order = FALSE
)
```

Using your sandbox, plot both `g()` and `slope_of_g()` (in blue) on a domain $-1 \leq x \leq 1$. This can be done with `slicePlot()` in the following way:

```{r results="hide"}
slice_plot(g(x) ~ x, domain(x=c(-1,1))) %>%
  slice_plot(slope_of_g(x) ~ x, color="blue")
```
```{r wtf1-2, echo=FALSE, results="markup"}
askMC(
  "Which of these statements best describes the graph of $g()$ compared to `slope_of_g()`?",
  "`slope_of_g()` is negative compared to `g(x)`.+",
  "`slope_of_g()` is shifted left by about $1$ compared to `g(x)`.",
  "`slope_of_g()` has a much smaller amplitude than `g()`.", 
  "+`slope_of_g()` is practically the same function as `g()`. That is, for any input the output of the two functions is practically the same.+"
)
```

