---
id: "octopus-toss-gloves"
created: "Thu Dec 16 14:30:51 2021"
global_id: "tm0XB4"
---

In this exercise, you will calculate the net present value of the social security payments for a woman retiring at age 70. We will assume that benefits are paid at a rate of $24,000 per year. 

The following R/mosaic commands will construct a function, `survive(age)` that gives the proportion of people alive at age 70 who will survive to an older age. (This is estimated from the Social Security mortality tables for 2014.)

```{r}
survive <- M2014F %>%
  filter(age >= 70) %>%
  select(age, nliving) %>%
  mutate(survival = nliving / 82818) %>%
  spliner(survival ~ age, data = .)
```

For a person living to age 100, the net present value at age 70 of the $24,000 annual social security benefit is 
$$\int_{70}^{100} 24000 \exp(-k*(age-70))\, dt$$
Calculate this amount for a continuously compounded interest rate of $k=0.03$ per year

```{r echo=FALSE, results="hide"}
Integrate(24000*exp(-0.03*(age-70)) ~ age, domain(age=70:100))
```
          
2. Benefit payments stop when a beneficiary dies. Of course, it is not known ahead of time when a retiree will die, but the `survive(age)` function gives a probability of surviving to any given age.

Modify the integral to incorporate the survival probability into the net present value calculation. That is, calculate the expectation value of the net present value.

```{r echo=FALSE, results="hide"}
Integrate(survive(age)*24000*exp(-0.03*(age-70)) ~ age, domain(age=70:100))
```

3. Is `survive(age)` a probability density function? Explain why or why not.

<!-- The output is a probability, not a probability density. -->




