---
id: "pony-write-shirt"
created: "Fri Jan 21 13:34:36 2022"
global_id: "xL9gWV"
---

`r insert_calcZ_exercise("XX.XX", "xL9gWV", "Exercises/pony-write-shirt.Rmd")`

```{r include=FALSE}
library(Zcalc)
```

Likelihood with a ball painted in two colors: red and blue.

Let $r$ be the proportion covered by red, so $(1-r)$ is the proportion covered in blue.

The likelihood for a red observation is $r$; for a green observation it is $1-r$.

With three reds and two blues, the likelihood is $$L(rrrbb) = r^3 (1 - r)^2$$.

Show the likelihood for $n$ reds and $m$ blues

```{r}
likelihood <- makeFun(r^n * (1-r)^m ~ r, n=2, m=2)
loglikelihood <- makeFun(n*log(r) + m*log(1-r) ~ r, n=2, m=2)
slice_plot(loglikelihood(r, n=700, m=300) ~ r, domain(r=.65:.75))

```




## Three colors: rgb

Domain is triangular: (x + y) <=1

```{r}
L <- function(x, y, r=5,g=5,b=5) {
  ifelse(x+y > 1, NA,
         x^r * y^g * (1-(x+y))^b)
}
contour_plot(L(x, y, r=25, g=16, b=19) ~ x & y,
             domain(x=0:1, y=0:1), labels=FALSE)
```
