---
id: "shark-hang-radio"
created: "Mon Jul 19 18:40:24 2021"
global_id: "l5Higs"
---

Many homeowners do not completely own their home. Instead, a bank or other financial institution in part owns the home through the terms of a *mortgage*. The mortgagee (e.g., a bank) will receive a monthly payment from the mortgagor (that is, the home owner). In setting the amount of the monthly payment, the mortgagee takes into account the initial size of the loan covered by the mortgage, the duration of the mortgage (typically 360 months, a 30-year mortgage), the calculated credit-worthiness of the mortgagor and other risk factors such as the loan's initial amount as a fraction of the value of the house. 

This is not as intricate a calculation as it might seem. The mortgagee typically has a cost of money (expressed as an interest rate) and uses a credit report on the home owner to find an interest increment ("risk premium") deemed suitable to cover the risk of default by the mortgagor.

Given a yearly interest rate $r$ and the initial loan amount (the "principal") $P$, how to calculate the monthly payment $m$ amount so that it covers the interest on the principal while also reducing the principal to bring it to exactly zero at the end of the mortgage term.

Since this is a calculus course, we will make a slight simplification and treat the mortgage payment as a continuous instantaneous payment. This allows us to use the tools of calculus and differential equations, simplifying the calculation to those who understand the tools and completely obscuring it to those who do not.

Let's look at the principal $P(t)$ as a function of time. We will denominate $t$ in units of months. As time goes by, the payments accumulate to reduce the principal while the interest rate tends to increase the principal. The overall equation of motion is:
$$\dot{P} = r P - m$$

```{r mort1-1, echo=FALSE, results="markup"}
askMC(
  "Given that $P$ has units of dollars, what are the units of $m$?",
  "dollars",
  "+dollars/month+",
  "months",
  "months/dollar"
)
```

Suppose that the bank offers the loan at an effective rate of 5% per year. Since this is a continuous-time process, and because of the units of time, this does not mean that $r$ is set to 0.05. Instead, One needs to do a calculation based on the compound interest dynamics, $\dot{P} = r P$. 

```{r mort1-2, echo=FALSE, results="markup"}
askMC(
  "Which of these is an appropriate equation for computing the value of $r$ so that the effective interest rate is 5% per year? (Remember, time is in months.)",
  "+$e^{12 r} = 1.05$+",
  "$e^{12 r} = 0.05$",
  "$12 r = 1.05$",
  "$12 r = 0.05$"
)
```

