---
id: "beech-iron-clock"
created: "Wed Oct 20 12:06:30 2021"
global_id: "Gw42pX"
---

*Note: The radio-button multiple choice questions don't allow us to display a column vector as such. Instead, we use the notation involving a superscript T. For instance, $(2, -3)^T$ is a column vector which we would ordinarily write $\left(\begin{array}{c}2\\-3\end{array}\right)$ A vector written like  $(2, -3)$, without the $^T$, is a row vector.*


Suppose $$\vec{\mathbf{u}}  \equiv \left(\begin{array}{c}2\\-3\end{array}\right) \ \ \hbox{and}\  \ \vec{\mathbf{v}}  \equiv \left(\begin{array}{c}4\\1\end{array}\right)$$
Compute the following linear combinations (arithmetically) ...


```{r bicR-6, echo=FALSE}
askMC(
  prompt = "$- \\vec{\\mathbf{u}}  -8  \\vec{\\mathbf{v}}$",
    "+$(-34, -5)^T$+"  =  "",
    "$(-34, -5)$"  =  "The linear combination of two column vectors will be a column  vector.",
    "$(-34, 15)$"  = "This is a row vector,  not a column vector.",
    "$(34, 15)^T$"  =  "Watch your arithmetic.",
    "Invalid combination"  = ""
)
```


```{r bicR-7, echo=FALSE}
askMC(
  prompt = "$2.1 \\vec{\\mathbf{u}}  -1.3  \\vec{\\mathbf{v}}$",
  
    "+$(-1,  -7.6)^T$+"  =  "",
    "$(0.8)$"  =  "The linear combination of two column vectors will be a column  vector.",
    "$(-1,  -7.6)$"  = "This is a row vector,  not  a column vector.",
    "$(2.1, -1.3)^T$"  =  "These are the scalar multipliers in the linear  combination, not  the value of the linear combination.",
    "Invalid combination"  = ""
)
```
```{r bicR-8, echo=FALSE}
askMC(
  prompt = "$-0.5 \\vec{\\mathbf{u}}  + 3.2  \\vec{\\mathbf{v}}$",
  
    "+$(11.8, 4.7)^T$+"  =  "",
    "$(2.7)$"  =  "The linear combination of two column vectors will be a column  vector.",
    "$(13.8,  1.7)$"  = "This is a row vector",
    "$(34, 16)^T$"  =  "Watch your arithmetic.",
    "Invalid combination"  = ""
)
```
```{r bicR-5, echo=FALSE}
message <- "Check  your  arithmetic!"
askMC(
  prompt = "$7 \\vec{\\mathbf{u}}  + 5  \\vec{\\mathbf{v}}$",
  
    "+$(34, -16)^T$+"  =  "",
    "$(32, -14)^T$"  =  message,
    "$(34, -14)^T$"  = message,
    "$(32, -16)^T$"  =  message,
    "Invalid combination"  = ""
)
```

