---
id: "cheetah-go-jacket"
created: "Sun Feb  6 11:16:48 2022"
global_id: "EFgeD4"
---

`r insert_calcZ_exercise("XX.XX", "EFgeD4", "Exercises/cheetah-go-jacket.Rmd")`

::: {.underconstruction}
:::

Suppose we're interested in a domain $0 \leq t \leq 10$ and a set of sinusoid functions:

$$s_1(t) \equiv \sin(2 \pi t/10)\\
s_2(t) \equiv \sin(4 \pi t/10)\\
s_3(t) \equiv \sin(6 \pi t/10)\\
c_1(t) \equiv \cos(2 \pi t/10)\\
c_2(t) \equiv \cos(4 \pi t/10)\\
c_3(t) \equiv \cos(6 \pi t/10)
$$

Each of these functions goes through an integer number of cycles over $0 \leq t  \leq 10$, as you can confirm by graphing them, e.g.
```{r results="hide"}
s1 <- makeFun(sin(2*pi*t/10) ~ t)
s2 <- makeFun(sin(4*pi*t/10) ~ t)
# and so on
c2 <- makeFun(cos(4*pi*t/10) ~ t)
c3 <- makeFun(cos(6*pi*t/10) ~ t)
slice_plot(s1(t) ~ t, domain(t=0:10)) %>%
  slice_plot(c2(t) ~ t, color="blue")
```
Using the definition of the dot product between two functions as
$$f(t) \bullet g(t) \equiv \int_0^{10} f(t)\, g(t) dt\ ,$$


a. Calculate the "length" of each of the functions $s_1(t), \ldots, c_3(t)$.
b. Calculate the angle between each pair of functions.

To simplify your calculations, you might want to make use of these "helper" functions:

```{r}
fdot <- function(tilde1, tilde2, domain) {
  f <- makeFun(tilde1)
  g <- makeFun(tilde2)
  Integrate(f(t)*g(t) ~ t, domain)
}
fangle <- function(tilde1, tilde2, domain) {
  fdot(tilde1, tilde2, domain) /
    sqrt(fdot(tilde1, tilde1, domain) * 
         fdot(tilde2, tilde2, domain))
}
```

c. To judge from the set of angles you calculated, what about the functions $s_1(t), \ldots, c_3(t)$ would make it easy to project a function $h(t)$ onto the subspace spanned by the functions? 
