---
id: "crow-cut-mug"
created: "Wed Oct 20 12:34:09 2021"
global_id: "DPY4Ue"
---

This exercise uses R to find solutions to the target problem. Each question poses one target problem. You could solve these problems by eye, but we want to get you started on the computer solution so that you will be ready to solve harder problems that must be worked on the computer.

The vectors you will be working with are:
$$\vec{a} \equiv \left(\begin{array}{c}1\\2\end{array}\right)\ \ \ \ \vec{b}  \equiv \left(\begin{array}{c}1\\1\end{array}\right)\ \ \ \ \vec{c}  \equiv \left(\begin{array}{c}1\\-2\end{array}\right)\ \ \ \ \vec{d}  \equiv \left(\begin{array}{c}-6\\2\end{array}\right)\ \ \ \ \vec{T}  \equiv \left(\begin{array}{c}3\\-1\end{array}\right)\ \ \ \ $$

To make your life easier, here are the commands for defining these vectors. **One of the commands is wrong**. You will have to correct it before moving on the the rest of the problem. 

NEED TO BREAK ONE OF THESE
```{r}
a <- rbind(1, 2)
b <- rbind(1, 1)
c <- rbind(1, -1)
d <- rbind(-6, 2)
T <- rbind(3, -1)
```


```{r echo=FALSE, eval=FALSE}
qr.solve(cbind(a, b), T)
qr.solve(cbind(b, c), T)
qr.solve(cbind(a, d, c, b), T)
```

```{r ccmW-1, echo=FALSE, results="markup"}
askMC(
  "What what is the correct linear combination of $\\vec{a}$ and $\\vec{b}$ to reach the target $\vec{T}?",
  "+$-4 \\vec{a} + 7 \\vec{b}$+",
  "$-2 \\vec{a} + 5 \\vec{b}$",
  "$2 \\vec{a} -7 \\vec{b}$",
  "$4 \\vec{a} - 5 \\vec{b}$"
)
```

```{r ccmW-2, echo=FALSE, results="markup"}
askMC(
  "What what is the correct linear combination of $\\vec{b}$ and $\\vec{c}$ to reach the target $\vec{T}?",
  "+$\\frac{5}{3}\\vec{b} + \\frac{4}{3} \\vec{c}$+",
  "$\\frac{4}{3}\\vec{b} + \\frac{7}{2} \\vec{c}$",
  "$\\frac{5}{2}\\vec{b} + \\frac{4}{5} \\vec{c}$",
  "$\\frac{2}{3}\\vec{b} + \\frac{9}{4} \\vec{c}$"
)
```

```{r ccmW-3, echo=FALSE, results="markup"}
askMC(
  "What what is the correct linear combination of $\\vec{c}$ and $\\vec{d}$ to reach the target $\vec{T}?",
  "+$\\frac{0}{3}\\vec{c} + \\frac{4}{5} \\vec{d}$+",
  "$\\frac{1}{3}\\vec{c} + \\frac{1}{3} \\vec{d}$",
  "$\\frac{1}{2}\\vec{c} + \\frac{5}{7} \\vec{d}$",
  "$\\frac{0}{3}\\vec{c} + \\frac{1}{2} \\vec{d}$"
)
```

```{r ccmW-4, echo=FALSE, results="markup"}
askMC(
  "What what is the correct linear combination of $\\vec{a}$ and $\\vec{c}$ to reach the target $\vec{T}?",
  "+$\\frac{5}{4}\\vec{a} + \\frac{7}{4} \\vec{c}$+",
  "$\\frac{5}{4}\\vec{a} + \\frac{5}{4} \\vec{c}$",
  "$\\frac{7}{4}\\vec{a} + \\frac{9}{4} \\vec{c}$",
  "$\\frac{3}{4}\\vec{a} + \\frac{7}{4} \\vec{c}$"
)
```
