---
id: "maple-draw-closet"
created: "Sun Jan 16 08:45:33 2022"
global_id: "zK7fQn"
---

Construct a random 4-by-4 matrix named `A` whose columns are mutually orthogonal. Here's the process:

i. Construct a vector with four elements that has random elements. A command to do so is
    ```{r}
    v1 <- cbind(rnorm(4))
    ```
    This will be the first column of the matrix.

ii. Construct a new vector with random elements.
    ```{r}
    w <- cbind(rnorm(4))
    ```
    `w` will not be orthogonal to `v1` as you can confirm by calculating the dot product between `v1` and `w`. However, you can construct from the two vectors a new one that will be perpendicular to `v1`.
    ```{r}
    v2 <- w - (w %onto% v1)
    ```
iii. Take the mutually orthogonal vectors you already have and package them into a matrix `M`. Then construct a new random vector `w` and project it onto `M`.
    ```{r}
    M <- cbind(v1, v2)
    w <- cbind(rnorm(4))
    v3 <- w - (w %onto% M)
    ```
    Continue the process of step (iii) until you have 4 mutually orthogonal vectors, and collect them into the matrix `A`.
    
Use dot products to verify that `v1`, `v2`, `v3`, and `v4` are mutually orthogonal.
    
