


## Partial derivatives algebraically

The model we developed for the speed of a bicycle $V$ as a function of steepness $s$ of the road and bike gear $g$ is a second-order polynomial in $s$ and $g$ with five terms:

$$V(s, g) = a_0 + a_s s + a_g g + a_{sg} s g + a_{gg}g^2$$


```{r daily-digital-33-QA1, echo=FALSE}
askMC(
  prompt = "The complete second-order polynomial with two inputs has six terms. Which one is missing in $V(s, g)$",
    "+$a_{ss} s^2$+" = "",
    "$a_{gg} g^3$" = "That's  a third-order term.",
    "$a_{gg} g^{-2}$" = "The low-order polynomial framework does not include negative powers.",
    "$a_{g} g$" = "That's  in the model!",
    "$a_{sg} g/s$" = "The low-order polynomial framework does not include negative powers.",
  inline = FALSE
)
```

```{r daily-digital-33-QA2, echo=FALSE}
askMC(
  prompt = "Which of these is $\\partial_g V(s, g)$?",
    "+$a_{g} + a_{sg} s + 2 a_{gg} g$+" = "",
    "$a_0 + a_{g} + a_{sg} s + 2 a_{gg} g$" = "",
    "$a_{g} g + a_{sg} g + 2 a_{gg} g$" = "",
    "$a_{s} s + a_{sg} gs + a_{gg} g^2$" = "",
    "$a_{g} + a_{sg} s + 2 a_{gg}$" = "",
  inline = FALSE
)
```

```{r daily-digital-33-QA3, echo=FALSE}
askMC(
  prompt = "Which of these is $\\partial_s V(s, g)$?",
    "+$a_{s} + a_{sg} g$+" = "",
    "$a_{g} g + a_{sg} s + 2 a_{ss} s$" = "The function $V(s,g)$ does not have any $a_{ss} g^2$ term.",
    "$a_{g} g + a_{sg} s + 2 a_{gg} g$" = "There is a $a_{gg} g^2$ term in the model. But that does not contribute anything to $\\partial_s V()$ because that term has no dependence on $s$.",
    "$a_s s + a_{sg} sg$" = "You forgot to differentiate these terms with repect to $s$.",
  inline = FALSE
)
```

```{r daily-digital-33-QA4, echo=FALSE}
askMC()
  prompt = "Which of these is $\\partial_{sg} V(s, g)$?",
    "+$a_{sg}$+" = "",
    "$a_{sg} s$" = "",
    "$a_{sg} g$" = "",
    "$a_{sg} sg$" = "",
  inline = FALSE
)
```

```{r daily-digital-33-QA5, echo=FALSE}
askMC(
  prompt = "Which of these is $\\partial_{ss} V(s, g)$?",
    "+$0$+" = "",
    "$2a_{ss} s$" = "",
    "$a_{ss} s$" = "",
    "$2 a_{ss} g$" = "",
    "$a_{sg}$" = "",
  inline = FALSE
)
```


## Bicycling with missing terms

The following code chunk will fit the low-order polynomial model of the bicycle to the data used in class. The results are shown in 4 different ways:

1. The coefficients on the model
2. A contour plot of the model
3. A surface plot of the model
4. A slice plot showing speed as a function of gear for three different slopes of road.

You might find some of these displays more useful than others. Feel free to comment out (with a `#`) the ones that you don't find useful.

Notice that the "model formula" in the `lm()` function is

`V ~ s + g + I(s*g) + I(g^2)`

This expression contains just the input quantities in the model. The `lm()` function does the work of finding the best coefficients for a linear combination of those terms. In the following questions, you're going to remove terms (such as `+ I(s*g)` from the model formula) to see what happens to the model. In one of the questions, you will extend the formula with a `- 1` (which suppresses the intercept term that is ordinarily included in models).

<span title='daily-digital-33-sandbox1'>...</span>
```{r daily-digital-33-sandbox1, exercise=TRUE, exercise.cap="Adding and excluding model terms", exercise.eval=FALSE, exercise.lines=30, exercise.completion=FALSE, exercise.startover=TRUE, exercise.diagnostics=TRUE, warning=FALSE}
Bicycle_speed <- tibble::tribble(
    ~ s, ~ g, ~ V,
    8, 1, 2,
    8, 5, 1,
    8, 10,0,
    0, 1, 9,
    0, 5, 12,
    0, 10, 6,
   -8, 1, 12,
   -8, 5, 16,
   -8, 10, 20
)
# fit the model to the data
mod <- lm(V ~ s + g + I(s*g) + I(g^2) , data = Bicycle_speed)
knitr::kable(coef(mod))
mod_fun <- makeFun(mod) # turn the statistical model into a function
dom <- bounds(s = c(-8, 8), g = c(1, 10))
contour_plot(mod_fun(s, g) ~ s + g, dom)
interactive_plot(mod_fun(s, g) ~ s + g, dom)
slice_plot(mod_fun(s=0, g) ~ g, bounds(g=c(1,10)),
           color = "red", label_text = "flat") %>%
  slice_plot(mod_fun(s = -5, g) ~ g, color = "black",
             label_text = "downhill") %>%
  slice_plot(mod_fun(s = 5, g) ~ g, color = "blue",
             label_text = "uphill")%>%
  gf_labs(y = "Bike velocity (mph)", x = "Gear #")
```


> **Essay**: The `lm()` function automatically adds an \"intercept\" term to the model. You can suppress this by ending the model formula with `-1`. Explain briefly what happens when you suppress the intercept and to what extent that model makes sense for the bicycle situation.


Restore the sandbox to its original before you answer this question.

> **Essay**: The interaction term in the model is included by the `+ I(s*g)` component of the model formula. (Don't get confused: \"Interaction\" and \"intercept\" are completely different things.) Take out the interaction term, refit and re-display the model. Explain briefly what happens when you suppress the interaction term and to what extent that model makes sense for the bicycle situation.

Restore the sandbox to its original before you answer this question.

> **Essay**: Suppose you add in a quadratic term in `s` to the model. Explain briefly whether this changes the model a lot or not. Also, look at the coefficients found by `lm()` for this extended model. What about those coefficients accounts for whether the model changed by a little or a lot.

Restore the sandbox to its original before you answer this question.
 
> **Essay**: Add a new plot to the code box. It should be just like the slice-plot that was originally there, but instead of each slice holding road slope constant and showing velocity as a function of gear, change things so that gear is held constant and the plot shows velocity as a function of road slope. Explain in everyday terms what this new plot displays about the model and say whether you think it makes sense.
