


::: {.example data-latex=""}
**From Taylor to Lagrange**

In Chapter `r Chaps$taylor` we met a method introduced by Brook Taylor (1685–1731) to construct a polynomial of order-$n$ that approximates any smooth function $f(x)$ close enough to some center $x_0$. The method made use of the ability to differentiate $f(x)$ at $x_0$ and produced the general formula:
$$f(x) \approx f(x_0) + \frac{f'(x_0)}{1} \left[x-x_0\right] + \frac{f''(x_0)}{2!} \left[x-x_0\right]^2 + \frac{f'''(x_0)}{3!} \left[x-x_0\right]^3 + \cdots + \frac{f^{(n)}(x_0)}{n!} \left[x-x_0\right]^n$$ where $f'(x_0) \equiv \partial_x f(x)\left.{\Large\strut}\right|_{x=x_0}$ and so on.

Using polynomials as approximating functions has been an important theme in mathematics history. Brook Taylor was neither the first nor the last to take on the problem.

In 1795, [Joseph-Louis Lagrange](https://en.wikipedia.org/wiki/Joseph-Louis_Lagrange)
(1736 – 1813) published another method for constructing an approximating polynomial of order $n$. Whereas the Taylor polynomial builds the polynomial that exactly matches the first $n$ derivatives at the center point $x_0$, the Lagrange polynomial has a different objective: to match exactly the values of the target function $f(x)$ at a set of ***knots*** (input values) $x_0$, $x_1$, $x_2$, $\ldots, x_n$. @fig-lagrange-sine shows the situation with the knots shown as orange dots.

```{r echo=FALSE}
#| label: fig-lagrange-sin
#| fig-cap: "The Lagrange polynomial of order $n$ is arranged to pass exactly through $n+1$ points on the graph of a function $f(x)$."
f <- sin
make_lagrange_unit <- function(x, x0) {
  bottom <- paste("(",
                  paste(
                    paste0("(", x0," - ", setdiff(x, x0), ")"), collapse = "*"),
                  ")")
  top <- paste(paste0("(x - ", setdiff(x, x0), ")"), collapse = "*")
  f <- function(x) {}
  body(f) <- parse(text=paste(top, "/", bottom))

  f
}
make_lagrange_poly <- function(xs, ys) {
  funs <- lapply(xs, function(p) make_lagrange_unit(xs, p))
  f <- function(x) {
    res <- 0
    vals <- lapply(funs, function(f) do.call(f, list(x)))
    for (k in 1:length(vals))
      res <- res + vals[[k]]*ys[k]

    return(res)
  }
  f
}
tp <- makeFun(x - x^3/6 + x^5/120 ~ x)
Pts <- tibble(x=c(-1, 0, 1, 3, 5, 7), y=f(x))
lp <- make_lagrange_poly(Pts$x, Pts$y)
  slice_plot(tp(x) ~ x, bounds(x=-3:8), color="green", size=2, alpha=0.7,
             label_text="Taylor polynomial", label_x=0.1) %>%
  slice_plot(f(x) ~ x, bounds(x=-3:8), label_text="f(x)") %>%
  slice_plot(lp(x) ~ x, bounds(x=-3:8), color="magenta", size=2, alpha=0.5,
             label_text="Lagrange polynomial", label_x = 0.8) %>%
  gf_point(y ~ x, data = Pts, color="orange", size=6, alpha=0.5) %>%
  gf_lims(y=c(-1.5, 1.5))
```

The Lagrange polynomial is constructed of a linear combinations of functions, one for each of the knots. In the example of Figure \@ref(fig:lagrange-sine), there are 6 knots, hence six functions being combined. For knot 2, for instance, has coordinates $\left(\strut x_2, f(x_2)\right)$ and the corresponding function is:

$$p_2(x) = \frac{(x-x_1)}{(x_2 -x_1)}\left[\strut\cdot\right]\frac{(x-x_3)(x-x_4)(x-x_5)(x-x_6)}{(x_2 -x_3)(x_2 -x_4)(x_2 -x_5)(x_2 -x_6)}$$
The gap indicated by $\left[\strut\cdot\right]$ marks where a term being excluded. For $p_2(x)$ that excluded term is $\frac{(x-x_2)}{(x_2 - x_2)}$. The various functions $p_1(x)$, $p_2(x)$, $p_3(x)$ and so on each leave out an analogous term.

Three important facts to notice about these ingenious polynomial functions:

i. They all have the same polynomial order. For $k$ knots, the order is $k-1$.
i. Evaluated at $x_i$, the value of $p_i(x_i) = 1$. For instance, $p_2(x_2) = 1$.
i. Evaluated at $x_j$, where $j\neq i$, the value of $p_j(x_i) = 0$. For example, $p_2(x_3) = 0$.

The overall polynomial will be the linear combination $$p(x) = y_1\, p_1(x) +
y_2\, p_2(x) + \cdots + y_k\, p_k(x)\ .$$
Can you see why?
:::


## Time and tide

Periodicities from <https://tidesandcurrents.noaa.gov/harcon.html?id=8451552&type=>

```{r}
hour <- with(RI_tide, hour)
b    <- with(RI_tide, level)
sin1 <- sin(2*pi*hour/12.41)
cos1 <- cos(2*pi*hour/12.41)
sin2 <- sin(2*pi*hour/23.94)
cos2 <- cos(2*pi*hour/23.94)
sin3 <- sin(2*pi*hour/12)
cos3 <- cos(2*pi*hour/12)
sin4 <- sin(2*pi*hour/12.66)
cos4 <- cos(2*pi*hour/12.66)
A <- cbind(1, sin1, cos1, sin2, cos2, sin3, cos3, sin4, cos4)
mod1 <- b %onto% cbind(1, sin1, cos1)
mod2 <- b %onto% cbind(1, sin1, cos1, sin2, cos2)
x <- qr.solve(A, b)
mod3 <- A %*% x
gf_point(level ~ hour, data = RI_tide)  %>%
  gf_line(mod3 ~ hour, color="magenta")

```

Anchorage, AK

Components:
- M2 12.42 hours
- S2 12 hours
- N2 12.658 hours
- K1 23.935 hours

```{r}
hour <- with(Anchorage_tide, hour)
b    <- with(Anchorage_tide, level)
sin1 <- sin(2*pi*hour/12.42)
cos1 <- cos(2*pi*hour/12.42)
sin2 <- sin(2*pi*hour/23.935)
cos2 <- cos(2*pi*hour/23.935)
sin3 <- sin(2*pi*hour/12)
cos3 <- cos(2*pi*hour/12)
sin4 <- sin(2*pi*hour/12.658)
cos4 <- cos(2*pi*hour/12.658)
A <- cbind(1, sin1, cos1, sin2, cos2, sin3, cos3, sin4, cos4)
mod1 <- b %onto% cbind(1, sin1, cos1)
mod2 <- b %onto% cbind(1, sin1, cos1, sin2, cos2)
x <- qr.solve(A, b)
mod3 <- A %*% x
resid <- b - mod3
gf_line(level ~ hour, data = Anchorage_tide)  %>%
  gf_line(mod3 ~ hour, color="magenta") %>%
  gf_lims(x = c(0,1000))
gf_line(resid ~ hour) 

```

## Fourier transform

```{r}
# Fill in the missing data
Raw <- Anchorage_tide |> select(hour, level) %>%
  mutate(hour = round(hour, 1))
Even <- tibble(hour=seq(min(Raw$hour), max(Raw$hour), by=0.1))
Both <- Even |> full_join(Raw)
# fill in the missing data
Fix1 <- Both |> 
  mutate(level = ifelse(is.na(level), lag(level), level)) %>%
  mutate(level = ifelse(is.na(level), lag(level), level)) %>%
  mutate(level = ifelse(is.na(level), lag(level), level)) %>%
  mutate(level = ifelse(is.na(level), lag(level), level)) %>%
  mutate(level = ifelse(is.na(level), lag(level), level)) %>%
# Fill in a constant value for the missing days
  mutate(level = ifelse(is.na(level), 4.867, level))
FFT <- abs(fft(Fix1$level))
# bin 974 is 1 per day
FFT2 <- 
  tibble(freq=2*(1:10000)/974, amp=FFT[2:10001]) %>%
  mutate(period = 24/freq) %>%
  mutate(amp = ifelse(amp < .3e4, 0, amp)) %>%
  mutate(speed = 360/period)
gf_line(amp ~ period, size=.1,data = FFT2 |> filter(period<100)) |> gf_lims(x=c(10, 14))
```
