---
id: "bee-wake-bottle"
created: "Mon Jul 19 12:57:10 2021"
global_id: "3A3pMv"
skill: "N-SCL6M"
---

```{r echo=FALSE}
source("draw_flow.R", local=TRUE)
```

Consider this two-dimensional flow field:

```{r echo=FALSE, out-width: "100%", message=FALSE, warning=FALSE}
draw_flow(2736)
```

There are three fixed points visible. The next plots zoom in on each of the fixed points.

```{r echo=FALSE, out-width: "75%", message=FALSE, warning=FALSE, fig.show="hold"}
draw_flow(2736, center=c(-1.75, 1.6), width=.2, ngrid=11) %>% gf_labs(title = "(A)")
```

```{r zi1-1, echo=FALSE, results="markup"}
askMC(
  "Is the fixed point in (A) stable or not?",
  "Stable",
  "+Unstable+",
  random_answer_order = FALSE
)
```

```{r echo=FALSE, out-width: "75%", message=FALSE, warning=FALSE, fig.show="hold"}
draw_flow(2736, center=c(4.6, 2.95), width=.2, ngrid=11) %>% gf_labs(title = "(B)")
```

```{r zi1-2, echo=FALSE, results="markup"}
askMC(
  "Is the fixed point in (B) stable or not?",
  "Stable",
  "+Unstable+",
  random_answer_order = FALSE
)
```

```{r echo=FALSE, out-width: "75%", message=FALSE, warning=FALSE, fig.show="hold"}
draw_flow(2736, center=c(1.25, 2.5), width=.2, ngrid=11) %>% gf_labs(title = "(C)")
```

```{r zi1-3, echo=FALSE, results="markup"}
askMC(
  "The fixed point in (C) is called a \"saddle.\" It is stable in one direction and unstable in another. Which of these is correct?",
  "Stable in y-direction and unstable in the x-direction",
  "+Stable in the x-direction and unstble in y+",
  random_answer_order = FALSE
)
```

Here's a system which has 4 fixed points in the region shown.

```{r echo=FALSE, out-width: "100%", warning=FALSE, message=FALSE}
draw_flow(2742)
```
Plots (D) and (E) zoom in on two regions.

```{r echo=FALSE, out-width: "75%", warning=FALSE, message=FALSE, fig.show="hold"}
draw_flow(2742, center=c(3.7, 0.9), width=.2, ngrid=11) %>%
  gf_labs(title = "(D)")
```

```{r zi1-4, echo=FALSE, results="markup"}
askMC(
  "Which of the following is the best description of the behavior near the fixed point in (D)?",
  "Stable and rotating clockwise",
  "+Stable and rotating counter-clockwise+",
  "Unstable and rotating clockwise",
  "Unstable and rotating counter-clockwise",
  random_answer_order = FALSE
)
```


```{r echo=FALSE, out-width: "75%", warning=FALSE, message=FALSE, fig.show="hold"}
draw_flow(2742, center=c(-0.5, -1.8), width=.5, ngrid=17) %>%
  gf_labs(title = "(E) with two fixed points") %>%
  gf_refine(coord_fixed(xlim = c(-1, 0), ylim=c(-2.3, -1.3), 
                        clip = "on"))
```

```{r zi1-5, echo=FALSE, results="markup"}
askMC(
  "Which of the following is the best description of the behavior near the upper left fixed point in (E)? (Neutal stability means neither stable nor unstable; the trajectory just orbits around the fixed point.)",
  "Unstable and rotating clockwise",
  "Stable and rotating counter-clockwise",
  "+\"Neutral stability\" and rotating clockwise+",
  "\"Neutral stability\" and rotating counter-clockwise",
  random_answer_order = FALSE
)
```

Let's look a little more closely at the upper-left fixed point in graph (E):

```{r echo=FALSE, out-width: "75%", warning=FALSE, message=FALSE, fig.show="hold"}
draw_flow(2742, center=c(-0.79, -1.455), width=.025, ngrid=15) %>%
  gf_labs(title = "(F) A \"center\"") %>%
  gf_refine(coord_fixed(xlim = c(-0.815, -0.765), 
                        ylim = c(-1.48, -1.43),
                        clip="on"))

```

The pattern in figure (F) is clockwise rotation around the fixed point. This kind of pattern is of fundamental importance in physics and engineering.

