---
id: "birch-bring-bed"
created: "Tue Mar 22 10:53:38 2022"
global_id: "uYXvfX"
skill: "L-ZPDIJ dimensions"
---


Newton's Law of Cooling is about how a hot (or cold) object comes into equilibrium with the ambient temperature. For instance, you might have a cup of coffee at $200^\circ$F in a room at $70^\circ$F. Unless you drink it, the coffee will cool with time until it reaches the room's temperature. In this setting, $x_\text{fixed} = 70^\circ$F. The value of $a$ depends on how insulated the cup is (and details of evaporation, etc.). But if the coffee reaches very near to room temperature in 60 minutes, $a \approx 0.5$ with units 1/minute. `r mark(6300)`


```{r bbb03-a, echo=FALSE, results="markup"}
askMC(
  "What are the units of the output of $x(t)$?",
  "degrees F per minute",
  "+degrees F+",
  "1/minute",
  "1/degrees F"
)
```

```{r bbb03-b, echo=FALSE, results="markup"}
askMC(
  "What are the units of $\\dot{x}$?",
  "+degrees F per minute+",
  "degrees F",
  "1/minute",
  "1/degrees F"
)
```

```{r bbb03-c, echo=FALSE, results="markup"}
askMC(
  "What is $x$?",
  "the room temperature",
  "the initial temperature of the coffee",
  "+the instantaneous coffee temperature as a function of time+",
  "the fixed rate at which the coffee cools",
  "the instantaneous rate at which the coffee cools as a function of time"
)
```

```{r bbb03-d, echo=FALSE, results="markup"}
askMC(
  "What is $\\dot{x}$?",
  "the room temperature",
  "the initial temperature of the coffee",
  "the instantaneous coffee temperature as a function of time",
  "+the instantaneous rate at which the coffee cools as a function of time+"
)
```

