---
id: "chicken-say-lamp"
created: "Mon Jul 19 16:57:58 2021"
global_id: "N9iexG"
skill: "F-189B3"
---

At the very beginning of [this 1987 video](http://www.ukclimbing.com/news/item.php?id=49981), the climber, [Catherine Destivelle](https://en.wikipedia.org/wiki/Catherine_Destivelle), is dangling from a  rope. Make an estimate of how long the rope is (specifically the length of the rope from the climber's harness to the bolts at the top of the route), based on the eigenvalues of the linear dynamics of swinging.

The standard differential-equation model for the changing angle $\theta$ of a pendulum is $$L \partial_{tt} \theta = - g \theta$$ where $L$ is the length of the pendulum (in meters) and $g$ is the acceleration due to gravity (9.8 m/s^2 on the Earth's surface).


```{r csl57-1, echo=FALSE, results="markup"}
askMC(
"What are the eigenvalues of the standard model for swinging?",
  "$\\lambda_1 = L$ and $\\lambda_2 = g$",
  "$\\lambda_{1,2} = \\pm \\frac{L}{g}$",
  "$\\lambda_{1,2} = \\pm \\sqrt{L/g}$",
  "+$\\lambda_{1,2} = \\pm ⅈ \\sqrt{L/g}$+"
)
```

```{r csl57-2, echo=FALSE, results="markup"}
askMC(
  "Units for $L$ and $g$ were given in the paragraph above. Using these units, what are the corresponding units of the eigenvalues?",
  "Since ⅈ is \"imaginary\", it makes no sense to talk about units." = "Since we multiply $\\lambda$ by $t$ in the solution $e^{\\lambda t}$, there had better be some units to $\\lambda$!", 
  "seconds/meter",
  "+1/seconds+",
  "meters per second"
)
```

```{r csl57-3, echo=FALSE, results="markup"}
askMC(
"From the video, estimate the period of the oscillation. Which of these is closest to the duration of a full back-and-forth swing?",
  "2 seconds",
  "3 seconds" = "This is a good estimate of *half* a cycle.",
  "4 seconds",
  "+6 seconds+",
  random_answer_order = FALSE
)
```

```{r csl57-4, echo=FALSE, results="markup"}
askMC(
"It's conventional to give separate names to the components of $\\lambda$.  The \"real\" part is often denoted $k$, the ⅈ part is called $\\omega$. The period (in seconds, say) is the quantity $P$ such that $\\omega P = 2 \\pi$. What are the units of $\\omega$ for $P$ in seconds?",
  "seconds",
  "+1/seconds+",
  "meters per second",
  "It has no units.",
  random_answer_order = FALSE
)
```

Put together these three facts to find a formula for $L$. Remember: $\omega = \sqrt{L/g}$  and $6\,\text{sec}\,\omega = 2 \pi$ and $g = 9.8$m/s. 

```{r csl57-5, echo=FALSE, results="markup"}
askMC(
  "What's a good estimate of $L$?",
  "5 meters",
  "+10 meters+",
  "15 meters",
  "20 meters",
  random_answer_order = FALSE
)
```

PS. Glad to report that in 2022, Catherine turned 62 years old.

