---
id: "eagle-choose-mattress"
created: "Sun Mar 20 08:43:23 2022"
global_id: "owE1WY"
skill: "X-HGCV2"
---



```{r echo=FALSE}
flow_contour <- function(f, g, f2=f, g2=g, 
                         dom=bounds(u=-3:3, v=-3:3)) {
    contour_plot(f2(u,v) ~ u & v, 
               domain=dom, contours_at = 0,
               contour_color="blue", 
               labels=FALSE) %>%
    contour_plot(g2(u,v) ~ u & v, 
                 domain=dom,
                 contours_at = 0, contour_color="red",
                 labels=FALSE, inherit=FALSE) %>%
    vectorfield_plot(f(u,v) ~ u & v, g(u,v) ~ u & v, 
                     dom, 
                     transform=function(x) x^0.25) 
}
```

In each of the following graphs, a flow field is annotated with a red contour and a blue contour. Your task is to determine whether a contour corresponds to a zero of the horizontal component of flow, a zero of the vertical component of flow, or neither. (Remember, if the contour is at a zero of the horizontal flow, the flow on the contour will be entirely vertical. And *vice versa*.)

```{r echo=FALSE}
f <- doodle_fun( ~ u & v, seed=577)
g <- doodle_fun( ~ u & v, seed=186)
f2 <- doodle_fun( ~ u & v, seed=111)
flow_contour(f, g, f2, dom=bounds(u=-3:3, v=-3:3)) %>%
  gf_labs(subtitle="Flow A") %>%
  gf_refine(coord_fixed()) %>%
  gf_lims(y=c(-3,3), x=c(-3,3))

```
1. Plot A $\color{blue}{\text{blue contour}}$:

    i. Zero of horizontal flow
    ii. Zero of vertical flow
    iii. Neither <!-- Correct -->

2. Plot A $\color{red}{\text{red contour}}$:

    i. Zero of horizontal flow
    ii. Zero of vertical flow <!-- Correct -->
    iii. Neither

```{r echo=FALSE}
f <- doodle_fun( ~ u & v, seed=578)
g <- doodle_fun( ~ u & v, seed=187)
f2 <- doodle_fun( ~ u & v, seed=1112)
g2 <- doodle_fun( ~ u & v, seed=2322)
flow_contour(f, g, dom=bounds(u=-3:3, v=-4:2)) %>%
  gf_labs(subtitle="Flow B") %>%
  gf_refine(coord_fixed()) %>%
  gf_lims(y=c(-4,2), x = c(-3,3))
```

3. Plot B $\color{blue}{\text{blue contour}}$:

    i. Zero of horizontal flow
    ii. Zero of vertical flow
    iii. Neither <!-- Correct -->

4. Plot B $\color{red}{\text{red contour}}$:

    i. Zero of horizontal flow
    ii. Zero of vertical flow <!-- Correct -->
    iii. Neither
