---
id: "fir-meet-shirt"
created: "Tue Mar 22 14:19:20 2022"
global_id: "adxrU4"
skill: "WP-103-16 G-PVCX5"
---


We have seen that the solution $x(t)$ to the linear dynamical system in two state variables
$$\begin{array}{c}\partial_t x  = ax  +  b y \\ \partial_t y = c x +  dy\end{array} = \left[\begin{array}{cc}a & b\\c & d\end{array}\right] \left[\begin{array}{c}x\\y\end{array}\right]$$
can be written as a linear combination of two exponentials:

$x(t) = A e^{\lambda_1 t} + B e^{\lambda_2 t}$.

Let's call the two components of this linear combination the "A-part" and the "B-part."

In each of the following, you are given two specific numerical values for $\lambda_1$ and $\lambda_2$. Your task is to determine whether the A-part or the B-part (or both or neither) is transient.

```{r fms02-1, echo=FALSE, results="markup"}
askMC(
  "For $\\lambda_1 = -1$ and $\\lambda_2 = -0.5$, which parts are transient?",
  "A-part",
  "B-part",
  "+both parts+",
  "neither part",
  random_answer_order = FALSE
)
```

```{r fms02-2, echo=FALSE, results="markup"}
askMC(
  "For $\\lambda_1 = -0.01$ and $\\lambda_2 = 0.01$, which parts are transient?",
  "+A-part+",
  "B-part",
  "both parts",
  "neither part",
  random_answer_order = FALSE
)
```

```{r fms02-3, echo=FALSE, results="markup"}
askMC(
  "For $\\lambda_1 = 0.01$ and $\\lambda_2 = -0.3$, which parts are transient?",
  "A-part",
  "+B-part+",
  "both parts",
  "neither part",
  random_answer_order = FALSE
)
```

In answering the next two questions, keep in mind that for large $t$,  $$A e^{\lambda_1 t} + B e^{\lambda_2 t} \approx A e^{\lambda_1 t}\ \ \text{when}\ \ \lambda_2 < \lambda_1\ .$$


```{r fms02-4, echo=FALSE, results="markup"}
askMC(
  "For $\\lambda_1 = 0.01$ and $\\lambda_2 = 0.001$, which parts are transient?",
  "A-part",
  "+B-part+",
  "both parts",
  "neither part",
  random_answer_order = FALSE
)
```

```{r fms02-5, echo=FALSE, results="markup"}
askMC(
  "For $\\lambda_1 = -0.1 $ and $\\lambda_2 = -0.001$, which parts are transient?",
  "A-part",
  "B-part",
  "+both parts+" = "Both parts are exponential decay.",
  "neither part",
  random_answer_order = FALSE
)
```
