---
id: "fox-look-magnet"
created: "Sun Mar 20 12:23:27 2022"
global_id: "tNLucT"
skill: "U-5PMG7 V-4KF8Q K-LH9CX N-9IU5J"
---


This system of differential equations, called the ***Lorenz equations*** is famous as the first widely recognized example of a type of motion called ***chaos***.

\begin{eqnalign}
\partial_t x & = \sigma (y - x), \\
\partial_t y & = x (\rho - z) - y, \\
\partial_t z & = x y - \beta z
\end{eqnalign}

1. What are the state variables and what are the parameters?

2. What is the dimension of the state space?

3. For initial conditions $x_0 = -5.276, y_0 = -7.152, z_0=19.452$ and $\rho=28$, $\sigma=10$, and $\beta=8/3$, use `integrateODE()` to integrate the equations over $0 \leq t \leq 100$ with a stepsize of $dt=0.01$. (This means you should set use the arguments `bounds(t=0:50), dt=0.01`. Call your trajectory `T1`.

```{r echo=FALSE, results="hide"}
T1 <- integrateODE(dx ~ sigma*(y-x), dy ~(x*(rho-z) - y), 
                  dz ~ (x*y - beta*z),
                  x=-5.276, y=-7.152, z=19.452, 
                  rho=28, sigma=10, beta = 8/3, 
                  bounds(t=0:50), dt=0.01)
T2 <- integrateODE(dx ~ sigma*(y-x), dy ~(x*(rho-z) - y), 
                  dz ~ (x*y - beta*z),
                  x=-5.277, y=-7.153, z=19.453, 
                  rho=28, sigma=10, beta = 8/3, 
                  bounds(t=0:50), dt=0.01)
traj_plot(x(t) ~ t, T1)
traj_plot(y(t) ~ x(t), T1, npts=10000)
traj_plot(x(t) ~ t, T1, color="blue") %>% traj_plot(x(t) ~ t, T2, color="red")
```

4. Based on your results in (3), what are the values of $x$, $y$, and $z$ at time $t=50$.

5. Make a time series plot of $x(t)$. Note that $x(t)$ jumps between a  oscillation with negative $x$ and the same kind of oscillation with positive $x$.

6. Plot out the $y$ component of the trajectory versus the $x$ component of the trajectory. To get a smooth plot, you will have to use the `npts=10000` argument to `traj_plot()`. The trajectory will appear to cross itself, but this is because you are only plotting two of the state variables.

7. Create another trajectory `T2` in the same way you made `T1`. But change each of the initial conditions in the third decimal point. Then plot out the $x(t)$ time series for `T1` (with `color="blue"`) and the $x(t)$ time series for `T2` (with color="red"`) on the same axes. The two time series will track one another very closely, but then come out of sync with one another. 

8. The characteristic feature of chaos is that a small change in initial conditions can lead to a very large change in the time series. How long does it take for the two time series in (7) to become utterly out of sync with one another?

9. The function `traj_plot_3D()` provides a simple way to look at the trajectory in 3 dimensions. Such a plot shows that the trajectory does not in fact cross itself. Use this command:
```{r}
traj_plot_3D(x, y, z, T1, npts=5000)
```


