---
id: "kangaroo-wear-knife"
created: "Tue Mar 22 15:19:36 2022"
global_id: "CUfA2c"
skill: "WP-103-16"
---

We will work with the linear dynamical system

$$\partial_t x = -x\\
\partial_t y = -11 x - 2 y$$

Using `integrateODE()` find the solution $x(t)$ for $0 \leq t \leq 10$ for each of these three initial conditions:

Trajectory 1) $x_0=1, y_0=-1$

Trajectory 2) $x_0=0, y_0=1$

Trajectory 3) $x_0=1, y_0=0$

```{r eval=FALSE}
T1 <- integrateODE( ...., x=1, y=-1, bounds(t=0:10))
T2 <- integrateODE( ...., x=0, y=1,  bounds(t=0:10)) 
T3 <- integrateODE( ...., x=1, y=0,  bounds(t=0:10))
```

The three solutions will be the functions `T1$x()`, `T2$x()`, and `T3$x()`.

Since the dynamics are linear, there will be a simple relationship between the three solutions, just as there is a simple relationship between three vectors in a 2-dimensional vector space.

Find this simple relationship and make a graph that demonstrate it.

<!-- solution -->
```{r eval=FALSE, echo=FALSE}
T1 <- integrateODE(dx ~ -x, dy ~ -11*x - 2*y, x=1, y=-1, bounds(t=0:10))
T2 <- integrateODE(dx ~ -x, dy ~ -11*x - 2*y, x=0, y=1,  bounds(t=0:10)) 
T3 <- integrateODE(dx ~ -x, dy ~ -11*x - 2*y, x=1, y=0,  bounds(t=0:10))

slice_plot(T1$x(t) + T2$x(t) - T3$x(t) ~ t, bounds(t=0:10))
```


