---
id: "puppy-take-pantry"
created: "Tue Mar 22 09:18:42 2022"
global_id: "tzVzaH"
skill: "P-3LIKO"
---


Lanchester's model of combat is 
$$\partial_t R = - b B\\
\partial_t B = -r R$$ with both parameters $r$ and $b$ taken to be positive. 

Obviously, the state variables $R$ and $B$ are to be thought of as functions of time. 

1. Locate the fixed point of the model of combat. <!--0,0 -->

2. Is it stable? In answering this question, consider two different state spaces and explain why your answer is different in the two spaces:
    a. All four quadrants of the $(R, B)$ plane.
    b. Only the first quadrant---that is, $0 \leq R, 0 \leq B$---of the $(R, B)$ pane. 
    
At a stable fixed point the state quantities---$R$ and $B$ here---do not change. But sometimes there are other functions of the state variables that do not change even when the state is off of any fixed point. These are called ***conserved quantities***. Conserved quantities such as momentum and energy are important in physics.

The text introduced this conserved quantity for Lanchester's model: $$Q(R, B) \equiv rR^2 - b B^2\ .$$ 
Let's explore whether the components of the quantity are conserved individually.

```{r ptp47-3, echo=FALSE, results="markup"}
askMC(
  "3. Calculate $\\partial_t r R^2$. Which of these is right? (Hint: Use the chain rule of differentiation and Lanchester's equations to simplify more.)",
  "+$-2 rb RB$+",
  "$2rR$",
  "$-2r\\dot{R}$",
  "$2r\\dot{R} B$ "
)
```

```{r ptp47-4, echo=FALSE, results="markup"}
askMC(
  "4. Perform the similar calculation  $\\partial_t b B^2$. Using your result and the result from question (3), is it the case that $\\partial_t r R^2 = \\partial_t b B^2$?",
  "+Yes+",
  "No",
  "Depends on $b$ and $r$.",
  "Depends on $B$ and $R$.",
  random_answer_order=FALSE
)
```




Here are two more candidates for conserved quantities:

a. $rR - bB$
b. $rB - bR$

Take the derivative with respect to time of each of them to determine if they are conserved.

```{r ptp47-5, echo=FALSE, results="markup"}
askMC(
  "5. Which of (a) and (b) are conserved?",
  "Just (a).",
  "Just (b).",
  "Both (a) and (b).",
  "+Neither (a) nor (b).+",
  random_answer_order = FALSE
)
```
