---
id: "rhinosaurus-sharpen-knob"
created: "Tue Mar 22 10:56:22 2022"
global_id: "4jkrKy"
skill: "L-ZPDIJ dimensions"
---


Our generic model for limited growth is

$$\partial_t x = r x (1-x/k)\ ,$$
which you can see as a modification of the proportional-growth model $$\partial_t x = r x\ .$$

The proportional-growth model will lead to $x(t)$ increasing without limit. Sometimes that is  a good model over short times, before $x(t)$ has a chance to get unrealistically big. But over long periods of time, something's gotta give. 

The limited growth model involves a ***carrying capacity*** $k$. For $t$ when $x(t)$ is very small, the population growth is well approximated by the usual proportion-growth model.  For intermediate $t$, the population has grown to a large enough size that it is consuming a substantial fraction of the available resources and population growth slows. When the population reaches the carrying capacity the growth stops (e.g. birth rate = death rate). 

```{r rsk65-a, echo=FALSE, results="markup"}
askMC(
  "If the units of $x(t)$ is, say, rabbits, what is the units of $k$?",
  "+rabbits+",
  "rabbits per day",
  "rabbits per week",
  "rabbits per year"
)
```

```{r rsk65-b, echo=FALSE, results="markup"}
askMC(
  "Suppose the units of $\\dot{x}$ is, rabbits per month, what is the units of $r$?",
  "+1/month+",
  "rabbits per month",
  "months per rabbit",
  "rabbits per month-squared"
)
```

```{r rsk65-c, echo=FALSE, results="markup"}
askMC(
  "Even without finding the full solution $x(t)$ to the differential equation, you can figure out how big the population will be when growth falls to zero. How big?",
  "+$k$+",
  "$k/r$",
  "$r/k$",
  "$1/r$"
)
```
