---
id: "squirrel-break-tv"
created: "Thu Mar 17 16:19:13 2022"
global_id: "aCMm4M"
skill: "J-YOLEG"
---


The graphs below show two dynamical functions, $f(x)$ and $g(x)$, in the two differential equations $\partial_t x = f(x)$ and $\partial_t x = g(x)$.

```{r echo=FALSE}
g <- makeFun((x-2.5)*(x-5)*(7.5+x)*(2.5 + x)*(x+10.1)*(x-10.2)/10000 ~ x)
slice_plot(g(x) ~ x, bounds(x=-10:10)) %>%
  gf_hline(yintercept= ~ 0, color="brown", size=2, alpha=0.5) %>%
  gf_labs(title = "f(x)")
```


```{r echo=FALSE}
f1 <- rfun(~ x, seed=7392)
f2 <- rfun(~ x, seed=992)

slice_plot((f1(x/2)*f2(x/2))/10 ~ x, bounds(x=-10:10)) %>%
  gf_hline(yintercept= ~ 0, color="brown", size=2, alpha=0.5) %>%
  gf_labs(title = "g(x)")
```


a. For function $f(x)$, 
    i. locate all of the fixed points in the domain $-10 \leq x \leq 10$ and determine whether each one is stable or not
    ii. find the state value $x$ where the state is growing the fastest.
    iii. find the state value $x$ where the state is decreasing the fastest.
    
b. Answer the questions in (a) but for $g(x)$.
