---
id: "tiger-sharpen-fork"
created: "Wed Mar  9 10:51:44 2022"
global_id: "CMBvRg" 
skill: "F-189B3"
---



Let's plot $e^{i\omega t}$ over the domain $0 < t < 10$ using $\omega = 2$. We need to be a little careful, since our plotting functions are not arranged to display complex numbers. But there is an easy workaround: plot the "real" and "imaginary" parts separately. 

```{r cn1}
f <- makeFun(exp(1i * omega * t) ~ t, omega = 2)
slice_plot(Re(f(t)) ~ t, 
           domain(t=c(0, 10)), color = "magenta") %>%
  slice_plot(Im(f(t)) ~ t, color="brown")
```

```{r cn2, echo=FALSE, results="markup"}
askMC(
  "Which part of $e^{i \\omega t}$ is the cosine function?",
  "+The \"real\" part+",
  "The \"imaginary\" part",
  "The negative of the \"imaginary\" part",
  "The negative of the \"real\" part"
)
```

Now let's consider $e^{(k + i\omega) t}$, where the input to the exponential function has a real part $k$ and an imaginary part $\omega$. As before, the output of the exponential will be a complex number, so we will plot the real and imaginary parts separately. 
```{r cn3}
g <- makeFun(exp((k + 1i * omega) * t) ~ t, omega = 2, k=-1)
slice_plot(Re(g(t)) ~ t, 
           domain(t=c(0, 10)), color = "orange3", npts=500) %>%
  slice_plot(Im(g(t)) ~ t, color="dodgerblue", npts=500)
```


```{r cn4, echo=FALSE, results="markup"}
askMC(
  "At what time $t$ does the sine-like function complete one full oscillation?",
  "At about $t=1.6$",
  "+At about $t=3.14$+",
  "At about $t=4.7$",
  "At about $t=6.28$",
  random_answer_order = FALSE
)
```

```{r cn5, echo=FALSE, results="markup"}
askMC(
  "Find a value for $\\omega$ that will produce one full oscillation every two time units. Graph it out to make sure that you have five full oscillations over the $0 < t < 10$ domain. What is that $\\omega$?",
  "$\\omega = 1$",
  "$\\omega = \\pi/2$",
  "+$\\omega = \\pi$+",
  "$\\omega = 2 \\pi$",
  random_answer_order = FALSE
)
```

```{r cn6, echo=FALSE, results="markup"}
askMC(
  "Keeping $\\omega$ at the value that produce five complete cycles over $0 < t < 10$, find a value for $k$ such that the amplitude of the oscillation at $t=10$ will be half that of the amplitude at $t=0$. What is $k$?",
  "$k \\approx -0.70$",
  "+$k \\approx -0.07$+",
  "$k \\approx 0.07$",
  "$k \\approx 0.70$",
  random_answer_order = FALSE
)
```

```{r cn7, echo=FALSE, results="markup"}
askMC(
  "Set $\\omega$ at the value that produces 20 complete cycles over $0 < t < 10$ and $k$ at the value such that the amplitude of the oscillation at $t=10$ will be *twice* that of the amplitude at $t=0$. What are $\\omega$ and $k$?",
  "$\\omega = \\pi,\\ k \\approx 0.35$",
  "$\\omega = 2 \\pi,\\ k \\approx 0.35$",
  "+$\\omega = 4 \\pi,\\ k \\approx 0.070$+",
  "$\\omega = 6 \\pi,\\ k \\approx 0.7$",
  random_answer_order = FALSE
)
```


