---
id: "walnut-tell-pot"
created: "Sun Dec  5 16:41:45 2021"
global_id: "iRZfBY"
---


In this exercise, you'll work with a particular function `f()` of two variables. Construct the function this way:

```{r}
f <- rfun( ~ x + y, seed=103)
```

The goal of this exercise is to explore the connections between the optimization method of ***gradient ascent*** or ***descent*** and dynamical systems. 

Here's a plot of the function and its ***gradient field***.

```{r warming=FALSE}
contour_plot(f(x, y) ~ x + y, domain(x=-4:4, y=-4:4)) %>%
  gradient_plot(f(x, y) ~ x + y) 
```

You can construct the $x$ and $y$ components of the gradient $\partial_x f(x,y)$ and $\partial_y f(x,y)$ of `f()` this way:

```{r warning=FALSE}
dx_f <- D(f(x, y) ~ x)
dy_f <- D(f(x, y) ~ y)
```

Use these two functions to define a dynamical system:
$$ \partial_t x = \partial_x f(x,y)\\
\partial_t y = \partial_y f(x,y)$$


1. Use `integrateODE()` to integrate the equations numerically from the initial condition $(x=0, y=-3)$. Plot the resulting trajectory as a layer on top of the contour plot and gradient field. Make the `tdur=` parameter in `integrateODE()` large enough to get very close to the high-point in the contour plot.

2. Modify the differential equations so that they correspond to gradient **descent** rather than ascent. Using the initial condition $(x=-1, y=1)$, numerically integrate the differential equations and, as in (1), plot the trajectory as a layer on the contour-plot/gradient-field. Make the `tdur=` parameter large enough to get very close to a local minimum of $f()$.

3. From (2), change the initial condition to $(x=-1.25, y=1)$ and plot the trajectory. What's different from the result in (2).


<!-- Solution -->
```{r echo=FALSE, results="hide", warning=FALSE}
soln1 <- integrateODE(dx ~ dx_f(x=x, y=y), 
                     dy ~ dy_f(x=x, y=y), 
                     x = 0, y = -3,
                     tdur = 10)
soln2 <- integrateODE(dx ~ -dx_f(x=x, y=y), 
                     dy ~ -dy_f(x=x, y=y), 
                     x = -1., y = 1,
                     tdur = list(from=0, to=10, dt=0.1))
soln3 <- integrateODE(dx ~ -dx_f(x=x, y=y), 
                     dy ~ -dy_f(x=x, y=y), 
                     x = -1.25, y = 1,
                     tdur = list(from=0, to=10, dt=0.1))

Common <- contour_plot(f(x, y) ~ x + y, domain(x=-4:4, y=-4:4)) %>%
  gradient_plot(f(x, y) ~ x + y) 
Common %>% traj_plot(y(t) ~ x(t), soln1, color="dodgerblue") %>%
  gf_labs(title="Plot 1")
Common %>%   
  traj_plot(y(t) ~ x(t), soln2, color="orange3") %>%
  traj_plot(y(t) ~ x(t), soln3, color="magenta") %>%
  gf_labs(title="Plot 2")
```


