---
id: "maple-think-ring"
created: "Mon Jul 19 15:05:12 2021"
global_id: "3D9C7d"
---

There is another format, called an **auto-regressive equation** for writing two dimensional finite-difference system. The auto-regressive format is widely used fields such as econometrics. Here is a system in [abcd]-matrix format and it's equivalent in auto-regressive format.

$$\underbrace{\left( \begin{array}{c}x_{n+1}\\y_{n+1}\end{array} \right) = 
\left( \begin{array}{cc}a & b \\1 &  0 \end{array} \right) \left( \begin{array}{c}x_{n}\\y_{n}\end{array} \right)}_\text{state-space format}\ \ \ \ \ \ \ \ 
\underbrace{x_{n+1} = a x_{n} + b y_{n}\strut}_{\text{auto-regressive format}\strut}$$

To see the equivalence, look to the equation for $y_{n+1}$ in the state-space format: $y_{n+1} = x_{n}$. In English, this translates to something like, "Tomorrow's y-component of the  state will be the same as today's x-component." But, unlike the usual situation in everyday life where a rule that applies this week might not apply the next week, the rules of finite difference equations apply for all $n$ without changing from one "week" to another or one "month" or "year" to another. Thus, the English statement is equivalent to, "Today's y-component will be the same as yesterday's x-component." Or, going back to mathematical notation, $y_n = x_{n-1}$. Substitute this relationship into the equation for $x_{n+1}$ in the state-space format to get the "auto-regressive" format:

$$x_{n+1} = \underbrace{a x_n + b y_n}_\text{state-space format} = \underbrace{a x_n + b x_{n-1}}_\text{auto-regressive format}$$
You can work with whatever format you prefer; they are the same dynamics. As you will see, we'll mostly use the state-space format because that's easier to display graphically and makes explicit the connection to the math of linear combinations. For some algebraic operations, the auto-regressive format can be more straightforward and is the more common format in fields such as economics. Some kinds of engineers use mainly one format, other kinds the other.

We're going to use the auto-regressive format to examine an *ansatz* to the AB10 discrete-time system. Recall the one-dimensional linear system with a fixed point at $x_n=0$. This had a simple solution: 
$$\underbrace{x_{n+1} = \alpha x_n}_\text{dynamics}\ \ \ \ \implies\ \ \ \ \underbrace{x_n = \alpha^n x_0}_\text{solution}$$ 

Let's try an *ansatz* of this form in the two-dimensional equation. But instead of $\alpha$ we'll use another Greek letter, lambda: $\lambda$. The ansatz will be: $$x_n = x_0 \lambda^n \ \ \mbox(ansatz)$$

Following the usual procedure we plug the ansatz into the dynamics to see if things work out.
$$x_0 \lambda^{n+1} = a x_0 \lambda^{n} + b x_0 \lambda^{n-1}$$
Dividing both sides of the equation by $x_0 \lambda^{n-1}$ gives
$$\lambda^2 = a \lambda + b$$

Since $a$ and $b$ will be what they will (depending on the system being modeled), not just any value of $\lambda$ will satisfy the relationship. But you can translate the relationship into one familiar from high-school, $\lambda^2 - a \lambda - b = 0$ and recall the formula for values of $\lambda$ that satisfy the relationship. There are two:

$$\lambda_1 = \frac{a + \sqrt{a^2 + 4b\strut}}{2}\\
\lambda_2 = \frac{a - \sqrt{a^2 + 4b\strut}}{2}$$
For good reasons that we haven't explained yet, $\lambda_1$ and $\lambda_2$ are called "characteristic values" of the AB10 system.

**Problem 1**: Consider the simple system $x_{t+1} = x_t + x_{t-1}$. This system is famous as the generator of the Fibonacci numbers. 

```{r tl1-1, echo=FALSE, results="markup"}
askMC(
  "Starting at $x_0=1$ and $x_1 = 0$, compute the first 10 elements of the Fibonacci numbers. Which one of these is it?",
  "+0, 1, 1, 2, 3, 5, 8, 13, 21, 34+",
  "0, 1, 1, 3, 5, 13, 21, 34, 55, 89", 
  "0, 0, 1, 1, 2, 3, 5, 8, 13, 21",
  "0, 1, 2, 3, 5, 8, 13, 34, 89"
)
```

```{r tl1-2, echo=FALSE, results="markup"}
askMC(
  "Using the correct answer from the previous question, calculate the ratio $x_{10}/x_9$. What is it?",
  "1.461", "1.528",
  "+1.612+", "1.841",
  random_answer_order = FALSE
)
```

```{r tl1-3, echo=FALSE, results="markup"}
askMC(
  "Recall the *ansatz* $x_n = x_0 \\lambda^n$. Under this ansatz, what is the ratio $x_{10}/ x_9$?", 
  "+$\\lambda$+",
  "$x_0 \\lambda$" = "The $x_0$ is on both the top and bottom of the ratio, cancelling it out.",
  "$\\lambda^2$",
  "$1-\\lambda$"
)
```

```{r tl1-4, echo=FALSE, results="markup"}
askMC(
  "Using the formulas for $\\lambda_1$ and $\\lambda_2$ and $a=b=1$ (the values used in the Fibonacci system), calculate numerical values for $\\lambda$. What are they?",
  "+(1.619, -0.619)+",
  "(1.619, 0.619)", 
  "(1.916, 0.916)",
  "(1.916, -0.916)",
  random_answer_order = FALSE
)
```

Notice that the $\lambda$ for the Fibonacci dynamics is the same as the ratio of successive elements (for large enough $n$) in the numerical Fibonacci sequence.

Using the sandbox, iterate the Fibonacci system (in state-space format) starting at $x_0 = 52.57$ and $y_0 = -85.07$.

```{r lt1-5, exercise=TRUE, exercise.cap="Fibonacci dynamics", exercise.nlines=4, eval=FALSE}
fib_dyn <- makeFun(c(x + y, x) ~ x + y)
Iterate(fib_dyn, x0=c(52.57, -85.07), n=10)
```
Perhaps strangely, the x-value gets smaller and smaller (and oscillates positive and negative), in contrast to the famous Fibonacci numbers which get bigger and bigger. But the underlying dynamics are the same.

```{r lt1-6, echo=FALSE, results="markup"}
askMC(
  "Consider the trajectory for the Fibonacci dynamics starting at $x_0 = 52.57$ and $y_0 = -85.07$. What is the ratio $x_{3} / x_2$?", 
  "$\\lambda_1$",
  "+$\\lambda_2$+",
  "$1/\\lambda_1$" = "Calculate $x_{3}/x_{2}$, not $x_2 / x_3$.",
  "$1/\\lambda_2$" = "Calculate $x_{3}/x_{2}$, not $x_2 / x_3$."
)
```

```{r lt1-7, echo=FALSE, results="markup"}
askMC(
  "Consider the trajectory for the Fibonacci dynamics starting at $x_0 = 52.57$ and $y_0 = -85.07$. What is the ratio $x_{10}} / x_9$?", 
  "+$\\lambda_1$+",
  "$\\lambda_2$",
  "$1/\\lambda_1$" = "Calculate $x_{3}/x_{2}$, not $x_2 / x_3$.",
  "$1/\\lambda_2$" = "Calculate $x_{3}/x_{2}$, not $x_2 / x_3$."
)
```

