---
id: "titmouse-see-jacket"
created: "Mon Jul 19 12:48:07 2021"
global_id: "VesZTL"
---

Consider the following discrete-time dynamics:
$$x_{n+1} = \alpha x_n \exp(-\beta x_n^3)$$

In the sandbox, implement the dynamical function and call it simply `f`. You're going to use this function in answering the later functions.

Note that you should write `alpha` in your function for the symbol $\alpha$, and `beta` for $\beta$. The two arguments at the end of the definition set default values for these two parameters. Also note that the dynamical function have been written in terms of $x$, not $x_n$. Of course, it does not really matter what you choose for the name of the argument to a function, so long as you use that name consistently.

The dynamical system uses the function $f()$. But to find fixed points, you will want to look for zeros of another function: $f(x) - x$.

a. Sketch the graph of $f(x) - x$. Your graphing domain should be large enough to display any local maxima as well as any fixed points of the dynamics.

<!--eval=FALSE lets this chunk be rendered in plain text. -->
```{r und-1-10, exercise=TRUE, exercise.cap="Implement the dynamics", exercise.nlines=6, eval=FALSE, warning=FALSE}
f <- makeFun( ..your.stuff.. ~ x, alpha = 2.72, beta=0.33)
slice_plot( ..your.stuff.. ~ x, domain(x=0:2)) 
```





b. For $\alpha = 2.72$ and $\beta=0.33$ (the defaults, as implemented in the original code in the sandbox), determine the location of the fixed point(s) and their stability.

```{r und-10-1-b, echo=FALSE, results="markup"}
askMC(
  "Where are the fixed points and are they stable?",
  "+At $x=0$ (unstable) and 1.447 (stable)+",
  "At $x=0$ (unstable) and 1.447 (unstable)",
  "At $x=0$ (stable) and 1.447 (stable)",
  "At $x=0$ (stable) and 1.447 (unstable)"
)
```

c. Iterate the dynamics for 10 steps, starting at $x_0 = 1.45$. (  Command: `Iterate(f, x0=1.45, n=10)`  ) 

```{r und-10-1-c, echo=FALSE, results="markup"}
askMC(
  "What does the trajectory tell you about the stability of the fixed point near $x_n = 1.45$?",
  "+The trajectory diverges from the fixed point, alternating between one side and the other.+",
  "The trajectory diverges from the fixed point, steadily moving to the right.",
  "The trajectory diverges from the fixed point, steadily moving to the left.",
  "The trajectory converges to the fixed point."
)
```

