---
id: "turtle-become-magnet"
created: "Mon Jul 19 12:47:01 2021"
global_id: "F0nHmq"
---

Consider the simple dynamical system $$x_{n+1} = g(x_n) \equiv \frac{1}{2} x_n$$

This has a fixed point $x^\star = 0$. You can see this by plugging in $x^\star = 0$ every place that $x_n$ or $x_{n+1}$ appears in the equation describing the dynamics and confirming that the left side is indeed equal to the right side.

Is the fixed point stable? Calculate $\partial_x g(x^\star)$ and check whether its absolute value is less than 1. If so, the fixed point is stable. Since $| \partial_x g(x^\star) | = \frac{1}{2} < 1$, the fixed point is stable.

Now consider $$x_{n+1} = g(x_n) \equiv 2 x_n$$ Again, this system has a fixed point $x^\star = 0$. Is it stable? Well, $\partial_x g(x^\star) = 2$. The magnitude $| \partial_x g(x^\star) |$ is greater than 1, so the fixed point is *unstable*.

We could play this game for any number in place of the $\frac{1}{2}$ or $2$. It would quickly get boring, so better to deal with the question once and for all for all possible numbers. We will *parameterize* the system, replacing the quantity that has been $\frac{1}{2}$ or $2$ in the previous examples with a symbol that stands for "some number here":
$$x_{n+1} = \alpha\, x_n$$
Does this system have a fixed point and where is it? The dynamical function is $g(x) = \alpha x$, so solve $g(x) - x = \alpha x^\star - x^\star = 0$. This becomes $$(\alpha - 1) x^\star = 0$$ which implies that $x^\star = 0$ is a fixed point or, if $\alpha=1$, any value for $x^\star$ will satisfy the fixed-point condition.

It is a common style in mathematical discourse to use parameters rather than specific numbers so that we can look at the same time at all systems of the same "form." For instance, we can summarize the stability of fixed points in the system $x_{n+1} = \alpha x_n$ by saying,

- There is a single fixed point at $x^\star = 0$, regardless of $\alpha$.
- That fixed point is stable if $|\alpha| < 1$, otherwise it is unstable. (Pedantic: In the singular case where $|\alpha| = 1$, the fixed point is "neutrally stable" and all points are fixed points.)

The advantage of using parameters is that we can often find the location of fixed points and their stability once and for all *as a function of the parameters*, and then use observed or desired behavior to set conditions on the range of parameter values that will produce that behavior.

```{r s4-1, echo=FALSE, results="markup"}
askMC("Consider the parameterized dynamical system $$x_{n+1} = \\alpha x_n + \\beta$$ What is the location of the fixed points in terms of the parameters $\\alpha$ and $\\beta$?",
  "+$x^\\star = \\frac{\\beta}{1-\\alpha}$+",
  "$x^\\star = \\frac{\\alpha}{1-\\beta}$",
  "$x^\\star = 0$",
  "$x^\\star = \\beta / \\alpha$"
  )
```

```{r s4-2, echo=FALSE, results="markup"}
askMC("In the system $$x_{n+1} = \\alpha x_n + \\beta$$ what is the condition for the fixed point to be stable?",
  "+$|\\alpha | < 1$+",
  "$|\\alpha| < |\\beta|$",
  "$|\\alpha| < \\beta$",
  "$|\\beta| < 1$"
)
```

```{r s4-3, echo=FALSE, results="markup"}
askMC(
  "You are modeling a real-world system with $$x_{n+1} = \\alpha x_n + \\beta$$ and you observe that the system tends toward a fixed value $x^\\star = 3.82$. Which one of these gives possible values for $\\alpha$ and $\\beta$?",
  "+$\\beta = 1$ and $\\alpha = 1 - 1/3.82$+",
  "$\\beta = 1$ and $\\alpha = 1 + 1/3.82$",
  "$\\beta = 0.5$ and $\\alpha = 1 - 1/3.82$",
  "$\\beta = 0.5$ and $\\alpha = 1 + 1/3.82$"
)
```

```{r s4-4, echo=FALSE, results="markup"}
askMC(
  "Now consider the dynamical system $$x_{n+1} = x_n^2 + x_n + \\alpha$$ In terms of the parameter $\\alpha$, where is/are the fixed point/s?",
  "+$x^\\star = \\pm \\sqrt{\\alpha}$+",
  "$x^\\star = \\pm \\alpha$",
  "$x^\\star = \\pm 1/\\alpha$",
  "$x^\\star = \\pm 1/\\sqrt{\\alpha}$"
  )
```

```{r s4-5, echo=FALSE, results="markup"}
askMC(
  "Again looking at the dynamical system $$x_{n+1} = x_n^2 + x_n + \\alpha$$ what are the conditions in terms of the parameter $\\alpha$ for the fixed point to be stable?",
  "+$| \\, 2 \\alpha + 1 \\, | < 1$+",
  "$| \\alpha | < 1$",
  "$| \\alpha^2 + 2 \\alpha | < 1$",
  "$\\alpha^2 < | \\alpha |$"
  )
```

